### 2015-10-15 PA-letter/manuscript.tex
catching up, PA letter version

% PA Letter
% http://www.oxfordjournals.org/our_journals/polana/for_authors/general.html
%
% 2-4 pages
% 1500-3000 words
% 1-3 tables/figures
%
% word count, cut past latex code here
% http://app.uio.no/ifi/texcount/online.php

\documentclass[letter]{article}
\usepackage[utf8x]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,array}
\usepackage{booktabs}
\usepackage{natbib}
\usepackage{graphicx}

%% load any required packages here

\graphicspath{{/Users/nilsmetternich/Dropbox/elements/gallium/graphs/}{../graphs/}}

\begin{document}

\title{Splitting it up: A split-duration package}
%\author{by Beger, Andreas and Chiba, Daina and Hill, Daniel W. and Metternich, Nils W. and Ward, Michael D.}

\maketitle

\begin{abstract}
In this paper, we present a split-population R package and an application to data on military coups. The statistical model accounts for units that are immune to a certain outcome and not part of the duration process the researcher is primarily interested in. We provide insights that if immune units exist, we can significantly increase the predictive performance compared to standard duration model. The package includes estimation and forecasting methods for split-population Weibull and Loglogistic Models.
\end{abstract}

\section{Introduction}
Duration models are an important class of statistical estimators that take into account the duration dependency of outcomes we are interested in. For example, the risk of dying is not independent of time. Newborns are at a great risk of dying, but as they grow older this risk declines and then gradually starts to increase again after the age of 9-10. This risk can be increased or decreased by behavioral (smoking, exercise, diet, etc) and structural factors (health care, regulations, urban vs rural, etc), but there is an underlying risk that is time dependent and impacts on all units.

However, sometimes we are interested in situations in which not all units are at risk of dying, failing, or acquiring some new characteristic. Imagine we are interested in the risk of acquiring the flu. First, we might think that everyone is at risk of getting the flu and it just depends on individual behavior (e.g. good hygiene) and structural factors (e.g. workplace) whether individuals will acquire the flu. If this would be the case we could use a standard duration model and estimate how different factors impact on the baseline risk. But there might also be individuals who are immune to the flu, because they received a flu shot, had the same flu virus in the past year, or some other factors that makes it impossible for them to get sick. Given that this is the case we have two underlying populations: An at risk population and an immune one. If the immune population is relatively large, estimates using a standard duration model will be biased and predictions resulting from such a model inaccurate.

\section{Immune populations and inference of duration processes}

Duration models, where the variable to be modeled describes a distribution of times until some event of interest, were originally developed in health and demographic research, where they grew naturally from life tables and survival records for medical patients. Basic formulations of such models, like the parametric exponential or Weibull regressions or semi-paramaetric Cox regression, implicitly assume that all subjects or units under observation, including right-censored observations, will eventually experience the event of interest. This assumption is outright wrong or at least disadvantageous in many substantive areas and empirical applications, where there instead is a sub-population of units or subjects that will never experience an event, and thus are effectively ``cured''.

This insight was realized as early as 1949 by \citet{boag1949maximum} and \citet{berkson1952survival}, who were researching survival rates in cancer patients following treatment, and where quite obviously some fraction of patients survived because their cancer was cured, while others relapsed after apparent remission due to levels of disease below detectable thresholds. As with conventional duration models, their origin in health and medicine has shaped the terminology conventionally used with such model, e.g. split-population duration models are sometimes also referred to in such contexts as cure rate models, and the basic concepts like survival and failure rates reference the survival of humans.

Yet the intuition underlying split-population duration models has led to applications in a broad range of subject areas outside demographics and medicine. In an early and foundational application that reached beyond criminal science, \citet{schmidt1989predicting} examined criminal recidivism using data on close to 10,000 prisoners from the North Carolina prison system in the late 1970's and early 1980's to identify factors that influence whether a criminal lapses at all, and if so which factors are related to the amount of time between prison stints. This work already includes a full formulation of the model with independent covariates for both the duration equation and the risk or cure equation, although only with subject-specific covariates rather than time-varying covariates with multiple data points per subject.

In public health, \citet{douglas1994hazard} use data from the US to model factors related to the age at which smokers started the habit, and \citet{forster2001role} examines the impact of tobacco taxes on smoking and quitting decisions. \citet{deyoung2003failure} models the failure of new commercial banks in the US during the 1980's. Lastly, in our own domain, political science, \citet{svolik2008authoritarian} used a split-population duration framework to study democratic consolidation, i.e. whether democratic regimes persist or slide back to authoritarianism, and when. Building on this effort, split-population duration models have also been used, along with other models, to produce regular predictions for five different forms of political conflict for the Integrated Crisis and Early Warning System (ICEWS) project \citep{ward2013learning} and to model irregular leadership changes for the Political Instability Task Force \citep[PITF; ][]{beger2014ensemble}.

\subsection{Model development}

Conventional duration models assume that all subjects will eventually fail, if we can only observe them long enough. The likelihood for a data point with survival time $t$ is thus the failure rate at that time or the probability of survival beyond $t$, depending on whether the observation is right-censored ($1-\delta_i$) or not:
\begin{eqnarray}
\mathcal{L} = \prod_{i=1}^N  \left( f(t_i)\right)^{\delta_i} \times \left( S(t_i) \right)^{1-\delta_i}
\end{eqnarray}

The major modeling question in this setting, which we will return to below, is the choice of a function to describe the evolution of the hazard rate $h(t) = \frac{f(t)}{S(t)}$ over time, e.g. with exponential, Weibull, or log-logistic densities.

The cumulative failure rate ($F(t) = 1 - S(t)$) over time converges to 1, meaning all subjects fail eventually. This assumption is untenable in many situations. Some cancer patients are cured after treatment while others relapse, most young people do not start smoking even though most smokers started at a young age, and many states will not experience the kind of violence that persists in some parts of the world. A better assumption in such contexts is that there is a subpopulation of subjects who are at risk of experiencing an event at some point, and another subpopulation who will never experience the event.

The presence of a large sub-population which is not at risk for an event in practice will inflate estimates of the survival fraction, and reduce hazard estimates for all subjects by essentially, and falsely, distributing risk among subjects that genuinely will fail and those that are cured. A model will thus over predict hazard for subjects that are not at risk (cured), and under predict for those who are at risk of eventually experiencing the event of interest.

We can incorporate the presence of a sub-population, where we label the subpopulation at risk with $\pi$, by rewriting the likelihood as:\footnote{Usual presentation of the split-population duration framework in medical contexts focus on the ``cured'' subpopulation. In our applications events are typically rare and it thus is easier to emphasize the ``risk'' subpopulation. As risk $= 1 - $ cured, this difference is trivial.}

\begin{align}
\mathcal{L}\{\theta|(t_{1}, \dots, t_{n})\} &= \prod_{i=1}^{N} \left(\pi_i f(t_i)\right)^{\delta_i} \times  \left((1-\pi_i) + \pi_i S(t_i)\right)^{1-\delta_i}
\end{align}

Crucially, this split-population framework is primarily useful in contexts where sub-populations are not clearly or easily identifiable. Even though there is a clear sub-populations in a model of the age at first pregnancy for humans---men---no researcher in their right mind would think to include male subjects in their data at all, for example. On the other hand, whether a cancer patient is cured or not cured given that they have no visible signs of cancer following treatment or have hit the 5-year disease free survival mark is much less clear and in fact the question that initially led to the idea for split-population duration modeling \citep{boag1949maximum, berkson1952survival}.

Early efforts focused only on the cure rate ($1 - \pi$) and treated it as a constant, but we can model membership in the subpopulation with its own covariates through a logistic link function:

\begin{align}
\pi_i &= \frac{1}{1 + e^{-z_i \gamma}}
\end{align}

Where $z_i$ is a vector of covariates for a subject at a given time. For interpretation, it is important to note that with time-varying covariates, the risk (or cured) estimate for a subject is particular to a given time point rather than constant for that subject over all time periods in the spell.\footnote{We use ``spell'' to designate all time periods observed for a subject up to the failure time. Subjects can theoretically have multiple spells, e.g. cancer patients who go into remission and relapse more than once, or states that experience multiple civil war onsets over their history.} Depending on the covariates, the risk estimate for a subject can thus fluctuate rapidly over time. To ease interpretation, it might be convenient to restrict covariates in the logit risk model to slow-moving, stable covariates in order to produce stable risk estimates for subjects.

The last component to complete the likelihood is the choice of a distribution for the shape of the hazard rate. The \texttt{spduration} package implements two hazard rate shapes, Weibull\footnote{In the code, $f(t) = \alpha \lambda^{\alpha} t^{\alpha - 1} e^{-(\lambda t)^\alpha}$.} and log-logistic:

\begin{eqnarray*}
\textrm{Weibull} \\
 f(t) & = & \alpha \lambda (\lambda t)^{\alpha - 1} e^{-(\lambda t)^\alpha} \\
  S(t) & = & e^{ -(\lambda t )^\alpha } \\
 h(t) & = & \alpha \lambda (\lambda t)^{\alpha-1} \\
\textrm{Log-logistic} \\
 f(t) & = & \frac{ \alpha \lambda (\lambda t)^{\alpha-1} }{ (1 + (\lambda t)^\alpha)^2 } \\
 S(t) & = & \frac{1}{ 1+  (\lambda t)^\alpha }  \\
 h(t) & = & \frac{ \alpha \lambda (\lambda t)^{\alpha-1} }{ 1+  (\lambda t)^\alpha }
\end{eqnarray*}

Where $\lambda = e^{-x_i\beta}$ is a parameter of covariates.

With these, the main quantity of interest is the conditional hazard $h(t, \pi)$, where both the risk/cure probabilities and hazard are conditional on survival to time $t$:

\begin{eqnarray}
h(t, \pi) = \frac{f(t, \pi)}{S(t, \pi)} & = & \frac{ \pi(t) \times f(t) }{ (1-\pi(t)) + \pi(t) \times S(t) } \\
 \pi(t) & = & \frac{ 1-\pi }{ S(t) + (1-\pi) (1 - S(t)) }
\end{eqnarray}

For a given unconditional risk rate $\pi$, the probability that a particular case with survival time $t$ is in the risk set decreases over time because an increasing number of surviving cases consist of immune or cured ($1-\pi$) cases that will never fail. In the hazard rate, the failure rate in the numerator is conditional on the probability that a case is in the risk set, give survival up to time $t$, and the numerator is an adjusted survivor function that accounts for the fraction of cured cases by time $t$, which is $1-\pi(t)$.

%
%		Need to add MCMC results and writeup here
%
%
\section{MCMC simulations}

\section{Fit a split-population model on coups data}

The \texttt{spduration} package implements a split-population duration regression for data with time-varying covariates and with two options for the hazard rate shape, the Weibull and log-logistic.\footnote{The Weibull allows for monotonically increasing or decreasing hazard rates, while the log-logistic allows for rates that first increase and then decrease.} The primary function, \texttt{spdur}, produces a regression model object of class \texttt{spdur} which can be used with the generic \texttt{summary} and \texttt{plot} functions to produce test statistics and graphical goodness-of-fit displays. The package also contains a second regression function, \texttt{spdurCrisp}, which can be used to automatically validate a model using held-out data, and can produce out-of-sample forecasts using the stored estimates.

To illustrate the usefulness of the package and the broad applicability of the methods outlined above, we employ an example from political science. We replicate and extend the analysis presented in \citet{belkin2003toward}, which examines factors that lead to military coups. Belkin and Schofer's distinction between ``structural'' vs. ``triggering'' causes of coups strongly suggests that a split-population approach to this question will be useful. They argue that many countries never experience coups because coups are effectvely impossible due to structural factors, while others that never experience coups are nevertheless at risk due to a different configuration of those same factors. Using language which fits nicely with the class of models described above, they say ``Triggers are not the source of the original risk, and in the absence of structural causes, the presence of triggering factors alone cannot lead to a coup. Hence, triggers should not be equated with coup risk. Rather, they are factors that may determine the exact timing of a coup in regimes that suffer from high coup risk'' \citep[p.\ 598][] {belkin2003toward}. Belkin and Schofer develop a measure of ``structural coup risk'' which incorporates factors such as the strength of democratic institutions and civil society, and a recent history of successful coups. They employ a logistic regression model which does not adequately capture the process described in the quote above, since it implicitly assumes that all observations are at risk for a coup (i.e., the probability of a coup is non-zero for all observations). Their structural coup risk indicator is developed precisely to distinguish between cases where coups are possible and cases where they are not, so the split-population model allows one to examine whether the indicator effectively separates at-risk cases from cases where coups are impossible.

We begin by loading the package and formatting the data appropriately. 
\small
\begin{verbatim}
> library(spduration)
> new.coup.data <- add_duration(coup.data, "coup", unitID="countryid", tID="year", 
+   freq="year")
\end{verbatim}
\normalsize
The \texttt{add}\_\texttt{duration} function takes as input a data frame with a binary response variable (in this example, ``coup'') that measures the occurrence of the event, or failure, recorded over discrete time periods, and produces a data frame with a duration variable (named ``duration'') for each spell as well as a binary variable (``atrisk'') that is constant across the spell and indicates whether the spell ended in failure. These are then used as response variables for the formula in the \texttt{spdur} function. Though the data used in this example are recorded annually, the function supports annual, monthly, or daily data. We begin by fitting first a Weibull and then a log-logistic split-population duration model using the coups data, including the measure of coup risk in the logit (risk) equation. 
\small
\begin{verbatim}
> weib.model <- spdur(duration~Military.Regime+Instability+Recent.War+Regional.Conflict,
+ 	 	atrisk~Coup.Risk+GDP.cap.+Military.Regime+
+ 	 	Recent.War+Regional.Conflict+
+ 	 	South.Am.+Central.Am.,data=new.coup.data)
> loglog.model <- spdur(duration~Military.Regime+Instability+Recent.War+Regional.Conflict,
+  	 	atrisk~Coup.Risk+GDP.cap.+Military.Regime+
+  	 	Recent.War+Regional.Conflict+
+ 	 	South.Am.+Central.Am.,data=new.coup.data,distr="loglog")
\end{verbatim}
\normalsize  
Using the \texttt{summary} function on either model object will produce standard output showing the model formula, estimates for the duration and risk equations, and test statistics with $p$-values. One may also use the \texttt{AIC} and \texttt{BIC} function to calculate the information criterion statistics for \texttt{spdur} objects.  
\small
\begin{verbatim}
> AIC(weib.model)
[1] 1120.774
> AIC(loglog.model)
[1] 1111.908
> BIC(weib.model)
[1] 1209.577
> BIC(loglog.model)
[1] 1200.711
\end{verbatim}
\normalsize
The objects produced by \texttt{spdur} can also be used with functions in the \texttt{xtable} package to produce \LaTeX\ code for tables of results. The \texttt{xtable} function used on an \texttt{spdur} object will produce code such as that used to create Table \ref{loglog_table}. 
\small
\begin{verbatim}
> xtable(loglog.model,caption="Coup model with log-logistic hazard")
\end{verbatim}
\begin{table}[ht]
\centering
\begin{tabular}{rlrrr}
  \hline
 & Parameter & Estimate & StdErr & p \\ 
  \hline
1 & (Dur. Intercept) & 3.29 & 0.14 & 0.00 \\ 
  2 & Military.Regime & -1.51 & 0.16 & 0.00 \\ 
  3 & Instability & -0.21 & 0.04 & 0.00 \\ 
  4 & Recent.War & -0.20 & 0.28 & 0.48 \\ 
  5 & Regional.Conflict & 2.64 & 2.56 & 0.30 \\ 
  6 & log(alpha) & -0.53 & 0.07 & 0.00 \\ 
  7 & (Risk Intercept) & -13.10 & 10.03 & 0.19 \\ 
  8 & Coup.Risk & 7.32 & 3.23 & 0.02 \\ 
  9 & GDP.cap. & 3.11 & 1.82 & 0.09 \\ 
  10 & Military.Regime.1 & 25.41 & 0.01 & 0.00 \\ 
  11 & Recent.War.1 & -4.34 & 2.70 & 0.11 \\ 
  12 & Regional.Conflict.1 & -39.74 & 23.96 & 0.10 \\ 
  13 & South.Am. & -3.72 & 3.19 & 0.24 \\ 
  14 & Central.Am. & -6.08 & 2.45 & 0.01 \\ 
   \hline
\end{tabular}
\caption{Coup model with log-logistic hazard} \label{loglog_table}
\end{table}
\normalsize
The table shows estimates from the duration equation, beginning with the intercept, and then estimates from the risk equation. Interestingly, the coup risk variable has a rather large and statistically significant coefficient in the risk equation.

The package also includes a function, \texttt{plot.hazard}, that can be used with an \texttt{spdur} object to plot the estimated hazard rate. \texttt{plot.hazard} plots the conditional hazard, which is the probability of survival conditional on the covariates in the risk and duration equations, and conditional on survival up to time $t$. The function calculates the estimated hazard rate as well as 90\% confidence intervals, which are produced by simulating values from the estimated sampling distributions of the model parameters. By default the function uses the mean values of the covariates during the simulations, but users can choose specific covariate values by entering them as vectors in the arguments \texttt{xvals} and \texttt{zvals}, which correspond to the covariates in the duration and risk equations, respectively. The command below creates Figure \ref{hazard}.
\begin{verbatim}
> plot.hazard(loglog.model)
\end{verbatim}
\begin{figure}[htbp!]
\centering
\includegraphics[width=4in]{loglog_hazard.pdf}
\caption{{\em Conditional hazard rate from log-logistic model.}} \label{hazard}
\end{figure}

Importantly, the package also includes functions to evaluate model predictions. The generic function \texttt{predict} can be used on an object of class \texttt{spdur} to generate several kinds of predictions, including the probability that an observation is ``at-risk'' and the probability of failure for a given time period. In addition, the generic \texttt{plot} function can be used on \texttt{spdur} objects to produce a separation plot \citep{greenhill2011separation}, which is a graphical display for evaluating model predictions. The code below produces Figure \ref{insamp}.
\small
\begin{verbatim}
> dev.new(width=9,height=3)
> par(mfrow=c(2,1),mar=c(2,2,2,2))
> plot(weib.model,endSpellOnly=F)
> plot(loglog.model,endSpellOnly=F)
\end{verbatim}
\normalsize
The option \texttt{endSpellOnly} is set to \texttt{FALSE} so that every observation, not only those at the end of a spell, is used in the plot. By default the \texttt{plot} function will calculate the conditional hazard for each observation. The separation plot sorts observations from left to right according to the predicted probability assigned by the model (higher values to the right), and shows each event/failure as red line, with non-events shown in beige. This makes it easy to see whether the model is assigning high probabilities of failure to actual cases of failure, and low probabilities to non-failures. 
\begin{figure}[htbp!]
\centering
\includegraphics[width=5in]{in-sample.pdf}
\caption{{\em In-sample separation plots for Weibull and log-logistic models.}} \label{insamp}
\end{figure}

Finally, we demonstrate use of the \texttt{spdurCrisp} function for evaluating out-of-sample predictions. We begin by splitting the data into training and test sets, using the last year of data (1999) as the test set and omitting incomplete observations from the test set. 
\small
\begin{verbatim}
> coup.train<-new.coup.data[coup.data$year<1999,]
> coup.test<-new.coup.data[new.coup.data$year==1999,]
> coup.test<-na.omit(coup.test)
\end{verbatim}
\normalsize
The \texttt{spdurCrisp} function can then be used to fit a model on the training set and generate predictions for the test set. 
\small
\begin{verbatim}
> weib.model2<-spdurCrisp(duration~Military.Regime+Instability+Recent.War+Regional.Conflict,
+ 		atrisk~Coup.Risk+GDP.cap.+Military.Regime+Recent.War+Regional.Conflict+
+ 		South.Am.+Central.Am.,
+ 		last='end.spell',train=coup.train,test=coup.test,pred=coup.test)
> weib.preds<-as.numeric(weib.model2$test.p)
>
> loglog.model2<-spdurCrisp(duration~Military.Regime+Instability+Recent.War+Regional.Conflict,
+ 		atrisk~Coup.Risk+GDP.cap.+Military.Regime+Recent.War+Regional.Conflict+
+ 		South.Am.,
+ 		last='end.spell',train=coup.train,test=coup.test,pred=coup.test,distr="loglog")
> loglog.preds<-as.numeric(loglog.model2$test.p)
\end{verbatim}
\normalsize
The \texttt{train},\texttt{test}, and \texttt{pred} arguments are used to specify data frames for the training set, test set, and forecasting set. We do not wish to generate forecasts here, so we simply specify the same data frame for the test and forecasting set. Here we use \texttt{spdurCrisp} to calculate the conditional risk, which is the default calculation for this function and is the probability that an observation is in the at-risk population. The test set predictions are stored in the model object as \texttt{test.p}, and we extract them to produce the plot shown below in Figure \ref{outsamp}, which was created using the \texttt{separationplot} package. 
\small
\begin{verbatim}
> dev.new(width=9,height=3)
> par(mfrow=c(2,1),mar=c(2,2,2,2))
> separationplot(weib.preds,coup.test$Coup,newplot=F,show.expected=T,lwd1=5,lwd2=2)
> separationplot(loglog.preds,coup.test$Coup,newplot=F,show.expected=T,lwd1=5,lwd2=2)
\end{verbatim}
\normalsize
\begin{figure}[htbp!]
\centering
\includegraphics[width=5in]{out-of-sample.pdf}
\caption{{\em Out-of-sample separation plots for Weibull and log-logistic models. }} \label{outsamp}
\end{figure}

\section{Censoring considerations}

Truncation and censoring are problematic for split-population duration models as they are for standard duration regression, but also pose some additional considerations. In left-truncation, we do not observe data for a spell prior to some date, and thus have incomplete and inaccurate values for the duration or time to failure for a spell. Since immune spells in the sample are over time going to distinguish themselves with exceptionally long survival times compared to spells at risk which fail periodically, left-censoring also makes it more difficult to distinguish the immune and at risk subpopulations.

Sometimes information about previous failures in the data is available beyond the time period over which covariates are observed, making it possible to ameliorate or eliminate left-censoring by using the information of previous failures when constructing the necessary duration variables with \texttt{add\_duration()}.

If information on previous failures is not available, it might be beneficial to drop some fraction of early observations for the same effect. For example, if the average failure times in the at risk sub-population are short relative to the time period over which data are available, it might be possible to eliminate most if not all left-censored spells without a large loss to the overall quantity of data.

Right-censoring, where spells end before outcomes are observed, also pose a unique problem in the split-population framework. Although right-censored spells themselves are accommodated in the modeling function, they impact the coding of at risk vs. immune spells. The \texttt{add\_duration()} function codes all observations in a spell as at risk if the spell itself ended in failure. Right-censored spells are coded as immune over their entire duration. This is problematic since some right-censored spells will fail outside the observation window, meaning they have been misclassified as immune spells. To our knowledge there has been no systematic investigation of this issue and the extent to which it is problematic. The only, untested, recommendation we can provide is that in some circumstances it might be helpful to add additional criteria before considering a spell immune, e.g. a survival time threshold.

The risk coding for right-censored spells is an important consideration for out-of-sample testing. A purely random sample of observations would include risk coding that is informed by observations that are potentially outside of the sample, and sampling, for example for cross-validation, should thus occur by spell. For block-wise testing, i.e. where some earlier data are used to train models and a later block of time is held back for out-of-sample testing, the \texttt{add\_duration()} function should be applied after partitioning the data, rather than to the complete data set before partitioning, in which case the risk coding for some spells in the training set might have been informed by outcomes in the test data.

\section{Conclusion}

\bibliography{spdur}

\end{document}

### 2015-10-15 pa-letter.tex
catching up, PA letter version

--- 2015-10-15 PA-letter/manuscript.tex
@@
-% PA Letter
-% http://www.oxfordjournals.org/our_journals/polana/for_authors/general.html
-%
-% 2-4 pages
-% 1500-3000 words
-% 1-3 tables/figures
-%
-% word count, cut past latex code here
-% http://app.uio.no/ifi/texcount/online.php
+\documentclass[]{article}
+\usepackage{lmodern}
+\usepackage{amssymb,amsmath}
+\usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+\else % if luatex or xelatex
+  \ifxetex
+    \usepackage{mathspec}
+    \usepackage{xltxtra,xunicode}
+  \else
+    \usepackage{fontspec}
+  \fi
+  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
+  \newcommand{\euro}{€}
+\fi
+% use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+% use microtype if available
+\IfFileExists{microtype.sty}{%
+\usepackage{microtype}
+\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\usepackage[margin=1.5in]{geometry}
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\ifxetex
+  \usepackage[setpagesize=false, % page size defined by xetex
+              unicode=false, % unicode breaks when used with xetex
+              xetex]{hyperref}
+\else
+  \usepackage[unicode=true]{hyperref}
+\fi
+\hypersetup{breaklinks=true,
+            bookmarks=true,
+            pdfauthor={Daniel W. Hill, Nils W. Metternich, Andreas Beger, and Michael D. Ward},
+            pdftitle={Splitting it up: A split-duration package},
+            colorlinks=true,
+            citecolor=blue,
+            urlcolor=blue,
+            linkcolor=magenta,
+            pdfborder={0 0 0}}
+\urlstyle{same}  % don't use monospace font for urls
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+\setlength{\emergencystretch}{3em}  % prevent overfull lines
+\setcounter{secnumdepth}{0}
 
-\documentclass[letter]{article}
-\usepackage[utf8x]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{amsmath,amssymb,array}
-\usepackage{booktabs}
-\usepackage{natbib}
-\usepackage{graphicx}
+%%% Use protect on footnotes to avoid problems with footnotes in titles
+\let\rmarkdownfootnote\footnote%
+\def\footnote{\protect\rmarkdownfootnote}
 
-%% load any required packages here
+%%% Change title format to be more compact
+\usepackage{titling}
 
-\graphicspath{{/Users/nilsmetternich/Dropbox/elements/gallium/graphs/}{../graphs/}}
+% Create subtitle command for use in maketitle
+\newcommand{\subtitle}[1]{
+  \posttitle{
+    \begin{center}\large#1\end{center}
+    }
+}
+
+\setlength{\droptitle}{-2em}
+  \title{Splitting it up: A split-duration package}
+  \pretitle{\vspace{\droptitle}\centering\huge}
+  \posttitle{\par}
+  \author{Daniel W. Hill, Nils W. Metternich, Andreas Beger, and Michael D. Ward}
+  \preauthor{\centering\large\emph}
+  \postauthor{\par}
+  \predate{\centering\large\emph}
+  \postdate{\par}
+  \date{June 8, 2015}
 
 \begin{document}
-
-\title{Splitting it up: A split-duration package}
-%\author{by Beger, Andreas and Chiba, Daina and Hill, Daniel W. and Metternich, Nils W. and Ward, Michael D.}
 
@@
 \begin{abstract}
-In this paper, we present a split-population R package and an application to data on military coups.
+We present an implementation of split-population duration regression in the \texttt{spduration} R package and an application to data on military coups.
 The statistical model accounts for units that are immune to a certain outcome and not part of the duration process the researcher is primarily interested in.
 We provide insights that if immune units exist, we can significantly increase the predictive performance compared to standard duration model.
-The package includes estimation and forecasting methods for split-population Weibull and Loglogistic Models.
+The package includes estimation and forecasting methods for split-population Weibull and Loglogistic models.
 \end{abstract}
 
-\section{Introduction}
-Duration models are an important class of statistical estimators that take into account the duration dependency of outcomes we are interested in.
-For example, the risk of dying is not independent of time.
-Newborns are at a great risk of dying, but as they grow older this risk declines and then gradually starts to increase again after the age of 9-10.
-This risk can be increased or decreased by behavioral (smoking, exercise, diet, etc) and structural factors (health care, regulations, urban vs rural, etc), but there is an underlying risk that is time dependent and impacts on all units.
+\section{Introduction}\label{introduction}
 
-However, sometimes we are interested in situations in which not all units are at risk of dying, failing, or acquiring some new characteristic.
-Imagine we are interested in the risk of acquiring the flu.
-First, we might think that everyone is at risk of getting the flu and it just depends on individual behavior (e.g.
-good hygiene) and structural factors (e.g.
-workplace) whether individuals will acquire the flu.
-If this would be the case we could use a standard duration model and estimate how different factors impact on the baseline risk.
-But there might also be individuals who are immune to the flu, because they received a flu shot, had the same flu virus in the past year, or some other factors that makes it impossible for them to get sick.
-Given that this is the case we have two underlying populations: An at risk population and an immune one.
-If the immune population is relatively large, estimates using a standard duration model will be biased and predictions resulting from such a model inaccurate.
+Duration models are an important class of statistical estimators that
+take into account the duration dependency of outcomes we are interested
+in. For example, the risk of dying is not independent of time. Newborns
+are at a great risk of dying, but as they grow older this risk declines
+and then gradually starts to increase again after the age of 9-10. This
+risk can be increased or decreased by behavioral (smoking, exercise,
+diet, etc) and structural factors (health care, regulations, urban vs
+rural, etc), but there is an underlying risk that is time dependent and
+impacts on all units.
 
-\section{Immune populations and inference of duration processes}
+However, sometimes we are interested in situations in which not all
+units are at risk of dying, failing, or acquiring some new
+characteristic. Imagine we are interested in the risk of acquiring the
+flu. First, we might think that everyone is at risk of getting the flu
+and it just depends on individual behavior (e.g.~good hygiene) and
+structural factors (e.g.~workplace) whether individuals will acquire the
+flu. If this would be the case we could use a standard duration model
+and estimate how different factors impact on the baseline risk. But
+there might also be individuals who are immune to the flu, because they
+received a flu shot, had the same flu virus in the past year, or some
+other factors that makes it impossible for them to get sick. Given that
+this is the case we have two underlying populations: An at risk
+population and an immune one. If the immune population is relatively
+large, estimates using a standard duration model will be biased and
+predictions resulting from such a model inaccurate.
 
-Duration models, where the variable to be modeled describes a distribution of times until some event of interest, were originally developed in health and demographic research, where they grew naturally from life tables and survival records for medical patients.
-Basic formulations of such models, like the parametric exponential or Weibull regressions or semi-paramaetric Cox regression, implicitly assume that all subjects or units under observation, including right-censored observations, will eventually experience the event of interest.
-This assumption is outright wrong or at least disadvantageous in many substantive areas and empirical applications, where there instead is a sub-population of units or subjects that will never experience an event, and thus are effectively ``cured''.
+\section{Immune populations and inference of duration
+processes}\label{immune-populations-and-inference-of-duration-processes}
 
-This insight was realized as early as 1949 by \citet{boag1949maximum} and \citet{berkson1952survival}, who were researching survival rates in cancer patients following treatment, and where quite obviously some fraction of patients survived because their cancer was cured, while others relapsed after apparent remission due to levels of disease below detectable thresholds.
-As with conventional duration models, their origin in health and medicine has shaped the terminology conventionally used with such model, e.g.
-split-population duration models are sometimes also referred to in such contexts as cure rate models, and the basic concepts like survival and failure rates reference the survival of humans.
+Duration models, where the variable to be modeled describes a
+distribution of times until some event of interest, were originally
+developed in health and demographic research, where they grew naturally
+from life tables and survival records for medical patients. Basic
+formulations of such models, like the parametric exponential or Weibull
+regressions or semi-paramaetric Cox regression, implicitly assume that
+all subjects or units under observation, including right-censored
+observations, will eventually experience the event of interest. This
+assumption is outright wrong or at least disadvantageous in many
+substantive areas and empirical applications, where there instead is a
+sub-population of units or subjects that will never experience an event,
+and thus are effectively ``cured''.
 
-Yet the intuition underlying split-population duration models has led to applications in a broad range of subject areas outside demographics and medicine.
-In an early and foundational application that reached beyond criminal science, \citet{schmidt1989predicting} examined criminal recidivism using data on close to 10,000 prisoners from the North Carolina prison system in the late 1970's and early 1980's to identify factors that influence whether a criminal lapses at all, and if so which factors are related to the amount of time between prison stints.
-This work already includes a full formulation of the model with independent covariates for both the duration equation and the risk or cure equation, although only with subject-specific covariates rather than time-varying covariates with multiple data points per subject.
+This insight was realized as early as 1949 by Boag (1949) and Berkson
+and Gage (1952), who were researching survival rates in cancer patients
+following treatment, and where quite obviously some fraction of patients
+survived because their cancer was cured, while others relapsed after
+apparent remission due to levels of disease below detectable thresholds.
+As with conventional duration models, their origin in health and
+medicine has shaped the terminology conventionally used with such model,
+e.g.~split-population duration models are sometimes also referred to in
+such contexts as cure rate models, and the basic concepts like survival
+and failure rates reference the survival of humans.
 
-In public health, \citet{douglas1994hazard} use data from the US to model factors related to the age at which smokers started the habit, and \citet{forster2001role} examines the impact of tobacco taxes on smoking and quitting decisions.
-\citet{deyoung2003failure} models the failure of new commercial banks in the US during the 1980's.
-Lastly, in our own domain, political science, \citet{svolik2008authoritarian} used a split-population duration framework to study democratic consolidation, i.e.
-whether democratic regimes persist or slide back to authoritarianism, and when.
-Building on this effort, split-population duration models have also been used, along with other models, to produce regular predictions for five different forms of political conflict for the Integrated Crisis and Early Warning System (ICEWS) project \citep{ward2013learning} and to model irregular leadership changes for the Political Instability Task Force \citep[PITF; ][]{beger2014ensemble}.
+Yet the intuition underlying split-population duration models has led to
+applications in a broad range of subject areas outside demographics and
+medicine. In an early and foundational application that reached beyond
+criminal science, Schmidt and Witte (1989) examined criminal recidivism
+using data on close to 10,000 prisoners from the North Carolina prison
+system in the late 1970's and early 1980's to identify factors that
+influence whether a criminal lapses at all, and if so which factors are
+related to the amount of time between prison stints. This work already
+includes a full formulation of the model with independent covariates for
+both the duration equation and the risk or cure equation, although only
+with subject-specific covariates rather than time-varying covariates
+with multiple data points per subject.
 
-\subsection{Model development}
+In public health, Douglas and Hariharan (1994) use data from the US to
+model factors related to the age at which smokers started the habit, and
+Forster and Jones (2001) examines the impact of tobacco taxes on smoking
+and quitting decisions. DeYoung (2003) models the failure of new
+commercial banks in the US during the 1980's. Lastly, in our own domain,
+political science, Svolik (2008) used a split-population duration
+framework to study democratic consolidation, i.e.~whether democratic
+regimes persist or slide back to authoritarianism, and when. Building on
+this effort, split-population duration models have also been used, along
+with other models, to produce regular predictions for five different
+forms of political conflict for the Integrated Crisis and Early Warning
+System (ICEWS) project (M. D. Ward et al. 2013) and to model irregular
+leadership changes for the Political Instability Task Force (PITF;
+Beger, Dorff, and Ward 2014).
 
-Conventional duration models assume that all subjects will eventually fail, if we can only observe them long enough.
-The likelihood for a data point with survival time $t$ is thus the failure rate at that time or the probability of survival beyond $t$, depending on whether the observation is right-censored ($1-\delta_i$) or not:
+\section{Model development}\label{model-development}
+
+Conventional duration models assume that all subjects will eventually
+fail, if we can only observe them long enough. The likelihood for a data
+point with survival time \(t\) is thus the failure rate at that time or
+the probability of survival beyond \(t\), depending on whether the
+observation is right-censored (\(1-\delta_i\)) or not:
+
 \begin{eqnarray}
@@
 
-The major modeling question in this setting, which we will return to below, is the choice of a function to describe the evolution of the hazard rate $h(t) = \frac{f(t)}{S(t)}$ over time, e.g.
-with exponential, Weibull, or log-logistic densities.
+The major modeling question in this setting, which we will return to
+below, is the choice of a function to describe the evolution of the
+hazard rate \(h(t) = \frac{f(t)}{S(t)}\) over time, e.g.~with
+exponential, Weibull, or log-logistic densities.
 
-The cumulative failure rate ($F(t) = 1 - S(t)$) over time converges to 1, meaning all subjects fail eventually.
-This assumption is untenable in many situations.
-Some cancer patients are cured after treatment while others relapse, most young people do not start smoking even though most smokers started at a young age, and many states will not experience the kind of violence that persists in some parts of the world.
-A better assumption in such contexts is that there is a subpopulation of subjects who are at risk of experiencing an event at some point, and another subpopulation who will never experience the event.
+The cumulative failure rate (\(F(t) = 1 - S(t)\)) over time converges to
+1, meaning all subjects fail eventually. This assumption is untenable in
+many situations. Some cancer patients are cured after treatment while
+others relapse, most young people do not start smoking even though most
+smokers started at a young age, and many states will not experience the
+kind of violence that persists in some parts of the world. A better
+assumption in such contexts is that there is a subpopulation of subjects
+who are at risk of experiencing an event at some point, and another
+subpopulation who will never experience the event.
 
-The presence of a large sub-population which is not at risk for an event in practice will inflate estimates of the survival fraction, and reduce hazard estimates for all subjects by essentially, and falsely, distributing risk among subjects that genuinely will fail and those that are cured.
-A model will thus over predict hazard for subjects that are not at risk (cured), and under predict for those who are at risk of eventually experiencing the event of interest.
+The presence of a large sub-population which is not at risk for an event
+in practice will inflate estimates of the survival fraction, and reduce
+hazard estimates for all subjects by essentially, and falsely,
+distributing risk among subjects that genuinely will fail and those that
+are cured. A model will thus over predict hazard for subjects that are
+not at risk (cured), and under predict for those who are at risk of
+eventually experiencing the event of interest.
 
-We can incorporate the presence of a sub-population, where we label the subpopulation at risk with $\pi$, by rewriting the likelihood as:\footnote{Usual presentation of the split-population duration framework in medical contexts focus on the ``cured'' subpopulation.
+We can incorporate the presence of a sub-population, where we label the
+subpopulation at risk with \(\pi\), by rewriting the likelihood
+as:\footnote{Usual presentation of the split-population duration framework in medical contexts focus on the ``cured'' subpopulation.
 In our applications events are typically rare and it thus is easier to emphasize the ``risk'' subpopulation.
@@
 
-Crucially, this split-population framework is primarily useful in contexts where sub-populations are not clearly or easily identifiable.
-Even though there is a clear sub-populations in a model of the age at first pregnancy for humans---men---no researcher in their right mind would think to include male subjects in their data at all, for example.
-On the other hand, whether a cancer patient is cured or not cured given that they have no visible signs of cancer following treatment or have hit the 5-year disease free survival mark is much less clear and in fact the question that initially led to the idea for split-population duration modeling \citep{boag1949maximum, berkson1952survival}.
+Crucially, this split-population framework is primarily useful in
+contexts where sub-populations are not clearly or easily identifiable.
+Even though there is a clear sub-populations in a model of the age at
+first pregnancy for humans---men---no researcher in their right mind
+would think to include male subjects in their data at all, for example.
+On the other hand, whether a cancer patient is cured or not cured given
+that they have no visible signs of cancer following treatment or have
+hit the 5-year disease free survival mark is much less clear and in fact
+the question that initially led to the idea for split-population
+duration modeling (Boag 1949, Berkson and Gage (1952)).
 
-Early efforts focused only on the cure rate ($1 - \pi$) and treated it as a constant, but we can model membership in the subpopulation with its own covariates through a logistic link function:
+Early efforts focused only on the cure rate (\(1 - \pi\)) and treated it
+as a constant, but we can model membership in the subpopulation with its
+own covariates through a logistic link function:
 
@@
 
-Where $z_i$ is a vector of covariates for a subject at a given time.
-For interpretation, it is important to note that with time-varying covariates, the risk (or cured) estimate for a subject is particular to a given time point rather than constant for that subject over all time periods in the spell.\footnote{We use ``spell'' to designate all time periods observed for a subject up to the failure time.
+Where \(z_i\) is a vector of covariates for a subject at a given time.
+For interpretation, it is important to note that with time-varying
+covariates, the risk (or cured) estimate for a subject is particular to
+a given time point rather than constant for that subject over all time
+periods in the
+spell.\footnote{We use ``spell'' to designate all time periods observed for a subject up to the failure time.
 Subjects can theoretically have multiple spells, e.g.
-cancer patients who go into remission and relapse more than once, or states that experience multiple civil war onsets over their history.} Depending on the covariates, the risk estimate for a subject can thus fluctuate rapidly over time.
-To ease interpretation, it might be convenient to restrict covariates in the logit risk model to slow-moving, stable covariates in order to produce stable risk estimates for subjects.
+cancer patients who go into remission and relapse more than once, or states that experience multiple civil war onsets over their history.}
+Depending on the covariates, the risk estimate for a subject can thus
+fluctuate rapidly over time. To ease interpretation, it might be
+convenient to restrict covariates in the logit risk model to
+slow-moving, stable covariates in order to produce stable risk estimates
+for subjects.
 
-The last component to complete the likelihood is the choice of a distribution for the shape of the hazard rate.
-The \texttt{spduration} package implements two hazard rate shapes, Weibull\footnote{In the code, $f(t) = \alpha \lambda^{\alpha} t^{\alpha - 1} e^{-(\lambda t)^\alpha}$.} and log-logistic:
+The last component to complete the likelihood is the choice of a
+distribution for the shape of the hazard rate. The \texttt{spduration}
+package implements two hazard rate shapes,
+Weibull\footnote{In the code, $f(t) = \alpha \lambda^{\alpha} t^{\alpha - 1} e^{-(\lambda t)^\alpha}$.}
+and log-logistic:
 
@@
 
-Where $\lambda = e^{-x_i\beta}$ is a parameter of covariates.
+Where \(\lambda = e^{-x_i\beta}\) is a parameter of covariates.
 
-With these, the main quantity of interest is the conditional hazard $h(t, \pi)$, where both the risk/cure probabilities and hazard are conditional on survival to time $t$:
+With these, the main quantity of interest is the conditional hazard
+\(h(t, \pi)\), where both the risk/cure probabilities and hazard are
+conditional on survival to time \(t\):
 
@@
 
-For a given unconditional risk rate $\pi$, the probability that a particular case with survival time $t$ is in the risk set decreases over time because an increasing number of surviving cases consist of immune or cured ($1-\pi$) cases that will never fail.
-In the hazard rate, the failure rate in the numerator is conditional on the probability that a case is in the risk set, give survival up to time $t$, and the numerator is an adjusted survivor function that accounts for the fraction of cured cases by time $t$, which is $1-\pi(t)$.
+For a given unconditional risk rate \(\pi\), the probability that a
+particular case with survival time \(t\) is in the risk set decreases
+over time because an increasing number of surviving cases consist of
+immune or cured (\(1-\pi\)) cases that will never fail. In the hazard
+rate, the failure rate in the numerator is conditional on the
+probability that a case is in the risk set, give survival up to time
+\(t\), and the numerator is an adjusted survivor function that accounts
+for the fraction of cured cases by time \(t\), which is \(1-\pi(t)\).
 
-%
-%		Need to add MCMC results and writeup here
-%
-%
-\section{MCMC simulations}
+\section{Fit a split-population model on coups
+data}\label{fit-a-split-population-model-on-coups-data}
 
-\section{Fit a split-population model on coups data}
+The \texttt{spduration} package implements a split-population duration
+regression for data with time-varying covariates and with two options
+for the hazard rate shape, the Weibull and
+log-logistic.\footnote{The Weibull allows for monotonically increasing or decreasing hazard rates, while the log-logistic allows for rates that first increase and then decrease.}
+The primary function, \texttt{spdur}, produces a regression model object
+of class \texttt{spdur} which can then be used with further methods.
 
-The \texttt{spduration} package implements a split-population duration regression for data with time-varying covariates and with two options for the hazard rate shape, the Weibull and log-logistic.\footnote{The Weibull allows for monotonically increasing or decreasing hazard rates, while the log-logistic allows for rates that first increase and then decrease.} The primary function, \texttt{spdur}, produces a regression model object of class \texttt{spdur} which can be used with the generic \texttt{summary} and \texttt{plot} functions to produce test statistics and graphical goodness-of-fit displays.
-The package also contains a second regression function, \texttt{spdurCrisp}, which can be used to automatically validate a model using held-out data, and can produce out-of-sample forecasts using the stored estimates.
-
-To illustrate the usefulness of the package and the broad applicability of the methods outlined above, we employ an example from political science.
-We replicate and extend the analysis presented in \citet{belkin2003toward}, which examines factors that lead to military coups.
+To illustrate the usefulness of the package and the broad applicability
+of the methods outlined above, we employ an example from political
+science. We replicate and extend the analysis presented in Belkin and
+Schofer (2003), which examines factors that lead to military coups.
 Belkin and Schofer's distinction between ``structural'' vs.
-``triggering'' causes of coups strongly suggests that a split-population approach to this question will be useful.
-They argue that many countries never experience coups because coups are effectvely impossible due to structural factors, while others that never experience coups are nevertheless at risk due to a different configuration of those same factors.
-Using language which fits nicely with the class of models described above, they say ``Triggers are not the source of the original risk, and in the absence of structural causes, the presence of triggering factors alone cannot lead to a coup.
-Hence, triggers should not be equated with coup risk.
-Rather, they are factors that may determine the exact timing of a coup in regimes that suffer from high coup risk'' \citep[p.\ 598][] {belkin2003toward}.
-Belkin and Schofer develop a measure of ``structural coup risk'' which incorporates factors such as the strength of democratic institutions and civil society, and a recent history of successful coups.
-They employ a logistic regression model which does not adequately capture the process described in the quote above, since it implicitly assumes that all observations are at risk for a coup (i.e., the probability of a coup is non-zero for all observations).
-Their structural coup risk indicator is developed precisely to distinguish between cases where coups are possible and cases where they are not, so the split-population model allows one to examine whether the indicator effectively separates at-risk cases from cases where coups are impossible.
+``triggering''" causes of coups strongly suggests that a
+split-population approach to this question will be useful. They argue
+that many countries never experience coups because coups are effectvely
+impossible due to structural factors, while others that never experience
+coups are nevertheless at risk due to a different configuration of those
+same factors. Using language which fits nicely with the class of models
+described above, they say ``Triggers are not the source of the original
+risk, and in the absence of structural causes, the presence of
+triggering factors alone cannot lead to a coup. Hence, triggers should
+not be equated with coup risk. Rather, they are factors that may
+determine the exact timing of a coup in regimes that suffer from high
+coup risk'' (p.~598 Belkin and Schofer 2003). Belkin and Schofer develop
+a measure of ``structural coup risk'' which incorporates factors such as
+the strength of democratic institutions and civil society, and a recent
+history of successful coups. They employ a logistic regression model
+which does not adequately capture the process described in the quote
+above, since it implicitly assumes that all observations are at risk for
+a coup (i.e., the probability of a coup is non-zero for all
+observations). Their structural coup risk indicator is developed
+precisely to distinguish between cases where coups are possible and
+cases where they are not, so the split-population model allows one to
+examine whether the indicator effectively separates at-risk cases from
+cases where coups are impossible.
 
@@
 \small
+
 \begin{verbatim}
@@
 \end{verbatim}
+
 \normalsize
-The \texttt{add}\_\texttt{duration} function takes as input a data frame with a binary response variable (in this example, ``coup'') that measures the occurrence of the event, or failure, recorded over discrete time periods, and produces a data frame with a duration variable (named ``duration'') for each spell as well as a binary variable (``atrisk'') that is constant across the spell and indicates whether the spell ended in failure.
-These are then used as response variables for the formula in the \texttt{spdur} function.
-Though the data used in this example are recorded annually, the function supports annual, monthly, or daily data.
-We begin by fitting first a Weibull and then a log-logistic split-population duration model using the coups data, including the measure of coup risk in the logit (risk) equation.
-\small
+The \texttt{add}\_\texttt{duration} function takes as input a data frame
+with a binary response variable (in this example, ``coup''``) that
+measures the occurrence of the event, or failure, recorded over discrete
+time periods, and produces a data frame with a duration variable
+(named''duration``) for each spell as well as a binary variable
+(''atrisk``) that is constant across the spell and indicates whether the
+spell ended in failure. These are then used as response variables for
+the formula in the \texttt{spdur} function. Though the data used in this
+example are recorded annually, the function supports annual, monthly, or
+daily data. We begin by fitting first a Weibull and then a log-logistic
+split-population duration model using the coups data, including the
+measure of coup risk in the logit (risk) equation. \small
+
 \begin{verbatim}
@@
 \end{verbatim}
+
 \normalsize  
-Using the \texttt{summary} function on either model object will produce standard output showing the model formula, estimates for the duration and risk equations, and test statistics with $p$-values.
-One may also use the \texttt{AIC} and \texttt{BIC} function to calculate the information criterion statistics for \texttt{spdur} objects.
-\small
+Using the \texttt{summary} function on either model object will produce
+standard output showing the model formula, estimates for the duration
+and risk equations, and test statistics with \(p\)-values. One may also
+use the \texttt{AIC} and \texttt{BIC} function to calculate the
+information criterion statistics for \texttt{spdur} objects.\\\small
+
 \begin{verbatim}
@@
 \end{verbatim}
+
 \normalsize
-The objects produced by \texttt{spdur} can also be used with functions in the \texttt{xtable} package to produce \LaTeX\ code for tables of results.
-The \texttt{xtable} function used on an \texttt{spdur} object will produce code such as that used to create Table \ref{loglog_table}.
+The objects produced by \texttt{spdur} can also be used with functions
+in the \texttt{xtable} package to produce \LaTeX~code for tables of
+results. The \texttt{xtable} function used on an \texttt{spdur} object
+will produce code such as that used to create Table \ref{loglog_table}.
 \small
+
 \begin{verbatim}
@@
 \end{verbatim}
+
 \begin{table}[ht]
 \centering
+\caption{Coup model with log-logistic hazard} \label{loglog_table}
 \begin{tabular}{rlrrr}
@@
 \end{tabular}
-\caption{Coup model with log-logistic hazard} \label{loglog_table}
 \end{table}
+
 \normalsize
-The table shows estimates from the duration equation, beginning with the intercept, and then estimates from the risk equation.
-Interestingly, the coup risk variable has a rather large and statistically significant coefficient in the risk equation.
+The table shows estimates from the duration equation, beginning with the
+intercept, and then estimates from the risk equation. Interestingly, the
+coup risk variable has a rather large and statistically significant
+coefficient in the risk equation.
 
-The package also includes a function, \texttt{plot.hazard}, that can be used with an \texttt{spdur} object to plot the estimated hazard rate.
-\texttt{plot.hazard} plots the conditional hazard, which is the probability of survival conditional on the covariates in the risk and duration equations, and conditional on survival up to time $t$.
-The function calculates the estimated hazard rate as well as 90\% confidence intervals, which are produced by simulating values from the estimated sampling distributions of the model parameters.
-By default the function uses the mean values of the covariates during the simulations, but users can choose specific covariate values by entering them as vectors in the arguments \texttt{xvals} and \texttt{zvals}, which correspond to the covariates in the duration and risk equations, respectively.
-The command below creates Figure \ref{hazard}.
+The package also includes a function, \texttt{plot.hazard}, that can be
+used with an \texttt{spdur} object to plot the estimated hazard rate.
+\texttt{plot.hazard} plots the conditional hazard, which is the
+probability of survival conditional on the covariates in the risk and
+duration equations, and conditional on survival up to time \(t\). The
+function calculates the estimated hazard rate as well as 90\% confidence
+intervals, which are produced by simulating values from the estimated
+sampling distributions of the model parameters. By default the function
+uses the mean values of the covariates during the simulations, but users
+can choose specific covariate values by entering them as vectors in the
+arguments \texttt{xvals} and \texttt{zvals}, which correspond to the
+covariates in the duration and risk equations, respectively. The command
+below creates Figure \ref{hazard}.
+
 \begin{verbatim}
@@
 \end{verbatim}
+
 \begin{figure}[htbp!]
@@
 \includegraphics[width=4in]{loglog_hazard.pdf}
-\caption{{\em Conditional hazard rate from log-logistic model.}} \label{hazard}
+\caption{Conditional hazard rate from log-logistic model.} \label{hazard}
 \end{figure}
 
-Importantly, the package also includes functions to evaluate model predictions.
-The generic function \texttt{predict} can be used on an object of class \texttt{spdur} to generate several kinds of predictions, including the probability that an observation is ``at-risk'' and the probability of failure for a given time period.
-In addition, the generic \texttt{plot} function can be used on \texttt{spdur} objects to produce a separation plot \citep{greenhill2011separation}, which is a graphical display for evaluating model predictions.
-The code below produces Figure \ref{insamp}.
-\small
+Importantly, the package also includes functions to evaluate model
+predictions. The generic function \texttt{predict} can be used on an
+object of class \texttt{spdur} to generate several kinds of predictions,
+including the probability that an observation is ``at-risk'' and the
+probability of failure for a given time period. In addition, the generic
+\texttt{plot} function can be used on \texttt{spdur} objects to produce
+a separation plot (Greenhill, Ward, and Sacks 2011), which is a
+graphical display for evaluating model predictions. The code below
+produces Figure \ref{insamp}. \small
+
 \begin{verbatim}
@@
 \end{verbatim}
+
 \normalsize
-The option \texttt{endSpellOnly} is set to \texttt{FALSE} so that every observation, not only those at the end of a spell, is used in the plot.
-By default the \texttt{plot} function will calculate the conditional hazard for each observation.
-The separation plot sorts observations from left to right according to the predicted probability assigned by the model (higher values to the right), and shows each event/failure as red line, with non-events shown in beige.
-This makes it easy to see whether the model is assigning high probabilities of failure to actual cases of failure, and low probabilities to non-failures.
+The option \texttt{endSpellOnly} is set to \texttt{FALSE} so that every
+observation, not only those at the end of a spell, is used in the plot.
+By default the \texttt{plot} function will calculate the conditional
+hazard for each observation. The separation plot sorts observations from
+left to right according to the predicted probability assigned by the
+model (higher values to the right), and shows each event/failure as red
+line, with non-events shown in beige. This makes it easy to see whether
+the model is assigning high probabilities of failure to actual cases of
+failure, and low probabilities to non-failures.
+
 \begin{figure}[htbp!]
@@
 
-Finally, we demonstrate use of the \texttt{spdurCrisp} function for evaluating out-of-sample predictions.
-We begin by splitting the data into training and test sets, using the last year of data (1999) as the test set and omitting incomplete observations from the test set.
-\small
+Finally, we demonstrate use of the \texttt{spdurCrisp} function for
+evaluating out-of-sample predictions. We begin by splitting the data
+into training and test sets, using the last year of data (1999) as the
+test set and omitting incomplete observations from the test set. \small
+
 \begin{verbatim}
@@
 \end{verbatim}
+
 \normalsize
-The \texttt{spdurCrisp} function can then be used to fit a model on the training set and generate predictions for the test set. 
-\small
+The \texttt{spdurCrisp} function can then be used to fit a model on the
+training set and generate predictions for the test set. \small
+
 \begin{verbatim}
@@
 \end{verbatim}
+
 \normalsize
-The \texttt{train},\texttt{test}, and \texttt{pred} arguments are used to specify data frames for the training set, test set, and forecasting set.
-We do not wish to generate forecasts here, so we simply specify the same data frame for the test and forecasting set.
-Here we use \texttt{spdurCrisp} to calculate the conditional risk, which is the default calculation for this function and is the probability that an observation is in the at-risk population.
-The test set predictions are stored in the model object as \texttt{test.p}, and we extract them to produce the plot shown below in Figure \ref{outsamp}, which was created using the \texttt{separationplot} package.
-\small
+The \texttt{train},\texttt{test}, and \texttt{pred} arguments are used
+to specify data frames for the training set, test set, and forecasting
+set. We do not wish to generate forecasts here, so we simply specify the
+same data frame for the test and forecasting set. Here we use
+\texttt{spdurCrisp} to calculate the conditional risk, which is the
+default calculation for this function and is the probability that an
+observation is in the at-risk population. The test set predictions are
+stored in the model object as \texttt{test.p}, and we extract them to
+produce the plot shown below in Figure \ref{outsamp}, which was created
+using the \texttt{separationplot} package. \small
+
 \begin{verbatim}
@@
 \end{verbatim}
+
+\includegraphics{out-of-sample.pdf}
+
 \normalsize
+
 \begin{figure}[htbp!]
@@
 
-Truncation and censoring are problematic for split-population duration models as they are for standard duration regression, but also pose some additional considerations.
-In left-truncation, we do not observe data for a spell prior to some date, and thus have incomplete and inaccurate values for the duration or time to failure for a spell.
-Since immune spells in the sample are over time going to distinguish themselves with exceptionally long survival times compared to spells at risk which fail periodically, left-censoring also makes it more difficult to distinguish the immune and at risk subpopulations.
+Truncation and censoring are problematic for split-population duration
+models as they are for standard duration regression, but also pose some
+additional considerations. In left-truncation, we do not observe data
+for a spell prior to some date, and thus have incomplete and inaccurate
+values for the duration or time to failure for a spell. Since immune
+spells in the sample are over time going to distinguish themselves with
+exceptionally long survival times compared to spells at risk which fail
+periodically, left-censoring also makes it more difficult to distinguish
+the immune and at risk subpopulations.
 
-Sometimes information about previous failures in the data is available beyond the time period over which covariates are observed, making it possible to ameliorate or eliminate left-censoring by using the information of previous failures when constructing the necessary duration variables with \texttt{add\_duration()}.
+Sometimes information about previous failures in the data is available
+beyond the time period over which covariates are observed, making it
+possible to ameliorate or eliminate left-censoring by using the
+information of previous failures when constructing the necessary
+duration variables with \texttt{add\_duration()}.
 
-If information on previous failures is not available, it might be beneficial to drop some fraction of early observations for the same effect.
-For example, if the average failure times in the at risk sub-population are short relative to the time period over which data are available, it might be possible to eliminate most if not all left-censored spells without a large loss to the overall quantity of data.
+Right-censoring, where spells end before outcomes are observed, also
+pose a unique problem in the split-population framework. Although
+right-censored spells themselves are accommodated in the modeling
+function, they impact the coding of at risk vs.~immune spells. The
+\texttt{add\_duration()} function retroactively codes all observations
+in a spell as at risk if the spell itself ended in failure.
+Right-censored spells are coded as immune over their entire duration.
+This can lead to some misclassification of observations as immune even
+though they experience failure at some point in the unobserved future.
 
-Right-censoring, where spells end before outcomes are observed, also pose a unique problem in the split-population framework.
-Although right-censored spells themselves are accommodated in the modeling function, they impact the coding of at risk vs.
-immune spells.
-The \texttt{add\_duration()} function codes all observations in a spell as at risk if the spell itself ended in failure.
-Right-censored spells are coded as immune over their entire duration.
-This is problematic since some right-censored spells will fail outside the observation window, meaning they have been misclassified as immune spells.
-To our knowledge there has been no systematic investigation of this issue and the extent to which it is problematic.
-The only, untested, recommendation we can provide is that in some circumstances it might be helpful to add additional criteria before considering a spell immune, e.g.
-a survival time threshold.
+Furthermore, in out-of-sample testing based on some kind of data
+partitioning scheme, this coding scheme can lead to unintentioned
+contamination of in-sample cases with knowledge of out-of-sample
+failures through the risk coding for failed spells. This leads to two
+recommendations. First, the duration data (\texttt{add\_duration})
+should not be built until after partitioning the data. Secondly, and as
+a result of the first point, data should be partitioned in a scheme that
+samples either by spell or block-wise, e.g.~by witholding the last
+\textbf{x} years of data, in order to be able to proberly build the
+duration variables.
 
-The risk coding for right-censored spells is an important consideration for out-of-sample testing.
-A purely random sample of observations would include risk coding that is informed by observations that are potentially outside of the sample, and sampling, for example for cross-validation, should thus occur by spell.
-For block-wise testing, i.e.
-where some earlier data are used to train models and a later block of time is held back for out-of-sample testing, the \texttt{add\_duration()} function should be applied after partitioning the data, rather than to the complete data set before partitioning, in which case the risk coding for some spells in the training set might have been informed by outcomes in the test data.
+\section{Conclusion}\label{conclusion}
 
-\section{Conclusion}
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width = 4in]{graphics/rates.pdf}
+\caption{Rates of positive outcomes in select publications with binary outcomes}
+\label{rates}
+\end{figure}
 
-\bibliography{spdur}
+Data and outcomes in international relations, and to a lesser extent in
+other fields as well, are typically characterized by temporal dependence
+and, where a binary, ordinal, or count outcome is modelled, rare events
+or severely imbalanced classes with excessive numbers of zeros. Figure
+\ref{rates} shows a few examples for published research that models
+binary outcomes.
+
+This is a well-recognized problem and has led to the development or use
+of several specialized mixture models like zero-inflated Poisson and
+negative binomial regression for count data, and a zero-inflated ordered
+probit for ordinal outcomes (Bagozzi et al. 2015). Split-population
+duration regression provides another principled solution to the
+challenges posed by data in this domain, but, unlike other solutions to
+the sparse outcome problem, also addresses underlying temporal dynamics.
+
+In addition to these technical features, split-population duration
+models are also appealing in the context of theories or arguments that
+explicitly distinguish long-term structural causes from short-term
+triggering events, as Belkin and Schofer (2003) and others do, as this
+conceptual distinction can be accomodated through the seperate risk and
+duration equations.
+
+\section*{References}\label{references}
+\addcontentsline{toc}{section}{References}
+
+Bagozzi, Benjamin E, Daniel W Hill, Will H Moore, and Bumba Mukherjee.
+2015. ``Modeling Two Types of Peace the Zero-Inflated Ordered Probit
+(ZiOP) Model in Conflict Research.'' \emph{Journal of Conflict
+Resolution} 59 (4). SAGE Publications: 728--52.
+
+Beger, Andreas, Cassy L. Dorff, and Michael D. Ward. 2014. ``Ensemble
+Forecasting of Irregular Leadership Change.'' \emph{Research \&
+Politics} 1 (3).
+
+Belkin, Aaron, and Evan Schofer. 2003. ``Toward a Structural
+Understanding of Coup Risk.'' \emph{Journal of Conflict Resolution} 47
+(5). Sage Publications: 594--620.
+
+Berkson, Joseph, and Robert P Gage. 1952. ``Survival Curve for Cancer
+Patients Following Treatment.'' \emph{Journal of the American
+Statistical Association} 47 (259). Taylor \& Francis Group: 501--15.
+
+Boag, John W. 1949. ``Maximum Likelihood Estimates of the Proportion of
+Patients Cured by Cancer Therapy.'' \emph{Journal of the Royal
+Statistical Society. Series B (Methodological)} 11 (1). JSTOR: 15--53.
+
+DeYoung, Robert. 2003. ``The Failure of New Entrants in Commercial
+Banking Markets: A Split-Population Duration Analysis.'' \emph{Review of
+Financial Economics} 12 (1). Elsevier: 7--33.
+
+Douglas, Stratford, and Govind Hariharan. 1994. ``The Hazard of Starting
+Smoking: Estimates from a Split Population Duration Model.''
+\emph{Journal of Health Economics} 13 (2). Elsevier: 213--30.
+
+Forster, Martin, and Andrew M Jones. 2001. ``The Role of Tobacco Taxes
+in Starting and Quitting Smoking: Duration Analysis of British Data.''
+\emph{Journal of the Royal Statistical Society: Series A (Statistics in
+Society)} 164 (3). Wiley Online Library: 517--47.
+
+Greenhill, Brian, Michael D Ward, and Audrey Sacks. 2011. ``The
+Separation Plot: A New Visual Method for Evaluating the Fit of Binary
+Models.'' \emph{American Journal of Political Science} 55 (4). Wiley
+Online Library: 991--1002.
+
+Schmidt, Peter, and Ann Dryden Witte. 1989. ``Predicting Criminal
+Recidivism Using Split Population Survival Time Models.'' \emph{Journal
+of Econometrics} 40 (1). Elsevier: 141--59.
+
+Svolik, Milan. 2008. ``Authoritarian Reversals and Democratic
+Consolidation.'' \emph{American Political Science Review} 102 (02).
+Cambridge Univ Press: 153--68.
+
+Ward, Michael D, Nils W Metternich, Cassy L Dorff, Max Gallop, Florian M
+Hollenbach, Anna Schultz, and Simon Weschle. 2013. ``Learning from the
+Past and Stepping into the Future: Toward a New Generation of Conflict
+Prediction.'' \emph{International Studies Review} 15 (4). Wiley Online
+Library: 473--90.
 

### 2015-10-15 belkin_schofer_redo.R

@@ -164,6 +164,7 @@ plot.hazard(loglog.model,xvals=c(1, 2, 1, 0),zvals=c(1,7,1,1,0,0,1))
 
 # playing around
 shape <- exp(weib.model$coefficients["log(alpha)"])
+shape <- exp(0.788)
 t <- seq(1, 35, length.out=200)
 
 hist(new.coup.data$duration[new.coup.data$failure==1], breaks=35, freq=F)

### 2015-10-15 ch-greater-1.r

@@ -1,0 +1,43 @@
+# Why is CH greater than 1 sometimes?
+
+user <- Sys.info()[["user"]]
+wd <- switch(user,
+             andybega="~/Work/spduration-paper",
+             danielhill="/Users/danielhill/Documents/spduration")
+dropbox <- switch(user,
+                  andybega="~/Dropbox/Work/gallium/data",
+                  danielhill="")
+setwd(wd)
+
+
+#library(spduration)
+#library(survival)
+library(foreign)
+
+coup.data <- read.dta(file="BelkinSchoferTable4.dta")
+coup.data$coup<-as.numeric(coup.data$coup)-1
+
+### split-pop models
+new.coup.data<-add_duration(coup.data, "coup", unitID="countryid", tID="year", freq="year")
+colnames(new.coup.data)[3:14]<-c("Coup.Risk","Recent.Coups","Recent.War","Military.Regime","GDP.cap.","Instability","Coup","Africa","Europe.North.Am.","South.Am.","Central.Am.","Regional.Conflict")
+
+summary(weib.model<-spdur(duration~Military.Regime+Instability+Recent.War+Regional.Conflict,
+                          atrisk~Coup.Risk+GDP.cap.+Military.Regime+
+                            Recent.War+Regional.Conflict+
+                            South.Am.+Central.Am.,data=new.coup.data))
+
+weib.preds <- predict(weib.model, truncate=FALSE, newdata=new.coup.data)
+range(weib.preds)
+
+# Flow:
+# change p_min in predict.spdur
+# run packagebuilder.r through load_all
+# run this script through here
+
+# p_min     max P
+# 1e-10     1.517623
+# 1e-15     1.517623
+# 1e-16     1.517623
+
+predict.spdur
+

### 2015-10-15 manuscript.Rmd
paper is in JSS style now

--- 2015-10-15 PA-letter/manuscript.tex
@@
-% PA Letter
-% http://www.oxfordjournals.org/our_journals/polana/for_authors/general.html
-%
-% 2-4 pages
-% 1500-3000 words
-% 1-3 tables/figures
-%
-% word count, cut past latex code here
-% http://app.uio.no/ifi/texcount/online.php
-
-\documentclass[letter]{article}
-\usepackage[utf8x]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{amsmath,amssymb,array}
-\usepackage{booktabs}
-\usepackage{natbib}
-\usepackage{graphicx}
-
-%% load any required packages here
-
-\graphicspath{{/Users/nilsmetternich/Dropbox/elements/gallium/graphs/}{../graphs/}}
-
-\begin{document}
-
-\title{Splitting it up: A split-duration package}
-%\author{by Beger, Andreas and Chiba, Daina and Hill, Daniel W. and Metternich, Nils W. and Ward, Michael D.}
-
-\maketitle
-
-\begin{abstract}
-In this paper, we present a split-population R package and an application to data on military coups.
+---
+author:
+  - name: Daniel W. Hill
+    affiliation: University of Georgia
+    address: >
+      First line
+      Second line
+    email: name@company.com
+    url: http://rstudio.com
+  - name: Nils W. Metternich
+    affiliation: University College London
+  - name: Andreas Beger
+    affiliation: who knows
+  - name: Michael D. Ward
+    affiliation: Duke University
+title:
+  formatted: "Splitting it up: the \\pkg{spduration} split-population duration regression package"
+  # If you use tex in the formatted title, also supply version without
+  plain:     "Splitting it up: the spduration split-population duration regression package"
+  # For running headers, if needed
+  short:     "\\pkg{spduration}: Split-population duration regression"
+abstract: >
+  We present an implementation of split-population duration regression in the   \pkg{spduration} R package and an application to data on military coups.
 The statistical model accounts for units that are immune to a certain outcome and not part of the duration process the researcher is primarily interested in.
 We provide insights that if immune units exist, we can significantly increase the predictive performance compared to standard duration model.
-The package includes estimation and forecasting methods for split-population Weibull and Loglogistic Models.
-\end{abstract}
+The package includes estimation and forecasting methods for split-population Weibull and Loglogistic models.
+keywords:
+  # at least one keyword must be supplied
+  formatted: [keywords, not capitalized, "\\proglang{R}"]
+  plain:     [keywords, not capitalized, R]
+preamble: >
+  \usepackage{amsmath}
+output: 
+  rticles::jss_article
+bibliography: spdur.bib
+---
 
-\section{Introduction}
+# Introduction
+
+This template demonstrates some of the basic latex you'll need to know to create a JSS article.
+
+## Code formatting
+
+Don't use markdown, instead use the more precise latex commands:
+
+* \proglang{Java}
+* \pkg{plyr}
+* \code{print("abc")}
+
+# Introduction
+
 Duration models are an important class of statistical estimators that take into account the duration dependency of outcomes we are interested in.
@@
 
-\section{Immune populations and inference of duration processes}
+# Immune populations and inference of duration processes
 
@@
 
-This insight was realized as early as 1949 by \citet{boag1949maximum} and \citet{berkson1952survival}, who were researching survival rates in cancer patients following treatment, and where quite obviously some fraction of patients survived because their cancer was cured, while others relapsed after apparent remission due to levels of disease below detectable thresholds.
+This insight was realized as early as 1949 by @boag1949maximum and @berkson1952survival, who were researching survival rates in cancer patients following treatment, and where quite obviously some fraction of patients survived because their cancer was cured, while others relapsed after apparent remission due to levels of disease below detectable thresholds.
 As with conventional duration models, their origin in health and medicine has shaped the terminology conventionally used with such model, e.g.
@@
 Yet the intuition underlying split-population duration models has led to applications in a broad range of subject areas outside demographics and medicine.
-In an early and foundational application that reached beyond criminal science, \citet{schmidt1989predicting} examined criminal recidivism using data on close to 10,000 prisoners from the North Carolina prison system in the late 1970's and early 1980's to identify factors that influence whether a criminal lapses at all, and if so which factors are related to the amount of time between prison stints.
+In an early and foundational application that reached beyond criminal science, @schmidt1989predicting examined criminal recidivism using data on close to 10,000 prisoners from the North Carolina prison system in the late 1970's and early 1980's to identify factors that influence whether a criminal lapses at all, and if so which factors are related to the amount of time between prison stints.
 This work already includes a full formulation of the model with independent covariates for both the duration equation and the risk or cure equation, although only with subject-specific covariates rather than time-varying covariates with multiple data points per subject.
 
-In public health, \citet{douglas1994hazard} use data from the US to model factors related to the age at which smokers started the habit, and \citet{forster2001role} examines the impact of tobacco taxes on smoking and quitting decisions.
-\citet{deyoung2003failure} models the failure of new commercial banks in the US during the 1980's.
-Lastly, in our own domain, political science, \citet{svolik2008authoritarian} used a split-population duration framework to study democratic consolidation, i.e.
+In public health, @douglas1994hazard use data from the US to model factors related to the age at which smokers started the habit, and @forster2001role examines the impact of tobacco taxes on smoking and quitting decisions.
+@deyoung2003failure models the failure of new commercial banks in the US during the 1980's.
+Lastly, in our own domain, political science, @svolik2008authoritarian used a split-population duration framework to study democratic consolidation, i.e.
 whether democratic regimes persist or slide back to authoritarianism, and when.
-Building on this effort, split-population duration models have also been used, along with other models, to produce regular predictions for five different forms of political conflict for the Integrated Crisis and Early Warning System (ICEWS) project \citep{ward2013learning} and to model irregular leadership changes for the Political Instability Task Force \citep[PITF; ][]{beger2014ensemble}.
+Building on this effort, split-population duration models have also been used, along with other models, to produce regular predictions for five different forms of political conflict for the Integrated Crisis and Early Warning System (ICEWS) project [@ward2013learning] and to model irregular leadership changes for the Political Instability Task Force [PITF; @beger2014ensemble].
 
-\subsection{Model development}
+# Model development
 
@@
 Even though there is a clear sub-populations in a model of the age at first pregnancy for humans---men---no researcher in their right mind would think to include male subjects in their data at all, for example.
-On the other hand, whether a cancer patient is cured or not cured given that they have no visible signs of cancer following treatment or have hit the 5-year disease free survival mark is much less clear and in fact the question that initially led to the idea for split-population duration modeling \citep{boag1949maximum, berkson1952survival}.
+On the other hand, whether a cancer patient is cured or not cured given that they have no visible signs of cancer following treatment or have hit the 5-year disease free survival mark is much less clear and in fact the question that initially led to the idea for split-population duration modeling [@boag1949maximum, @berkson1952survival].
 
@@
 The last component to complete the likelihood is the choice of a distribution for the shape of the hazard rate.
-The \texttt{spduration} package implements two hazard rate shapes, Weibull\footnote{In the code, $f(t) = \alpha \lambda^{\alpha} t^{\alpha - 1} e^{-(\lambda t)^\alpha}$.} and log-logistic:
+The \pkg{spduration} package implements two hazard rate shapes, Weibull\footnote{In the code, $f(t) = \alpha \lambda^{\alpha} t^{\alpha - 1} e^{-(\lambda t)^\alpha}$.} and log-logistic:
 
@@
 
-%
-%		Need to add MCMC results and writeup here
-%
-%
-\section{MCMC simulations}
+# Fit a split-population model on coups data
 
-\section{Fit a split-population model on coups data}
-
-The \texttt{spduration} package implements a split-population duration regression for data with time-varying covariates and with two options for the hazard rate shape, the Weibull and log-logistic.\footnote{The Weibull allows for monotonically increasing or decreasing hazard rates, while the log-logistic allows for rates that first increase and then decrease.} The primary function, \texttt{spdur}, produces a regression model object of class \texttt{spdur} which can be used with the generic \texttt{summary} and \texttt{plot} functions to produce test statistics and graphical goodness-of-fit displays.
-The package also contains a second regression function, \texttt{spdurCrisp}, which can be used to automatically validate a model using held-out data, and can produce out-of-sample forecasts using the stored estimates.
+The `spduration` package implements a split-population duration regression for data with time-varying covariates and with two options for the hazard rate shape, the Weibull and log-logistic.\footnote{The Weibull allows for monotonically increasing or decreasing hazard rates, while the log-logistic allows for rates that first increase and then decrease.} The primary function, \texttt{spdur}, produces a regression model object of class \texttt{spdur} which can then be used with further methods.
 
 To illustrate the usefulness of the package and the broad applicability of the methods outlined above, we employ an example from political science.
-We replicate and extend the analysis presented in \citet{belkin2003toward}, which examines factors that lead to military coups.
-Belkin and Schofer's distinction between ``structural'' vs.
-``triggering'' causes of coups strongly suggests that a split-population approach to this question will be useful.
+We replicate and extend the analysis presented in @belkin2003toward, which examines factors that lead to military coups.
+Belkin and Schofer's distinction between "structural" vs.
+"triggering"" causes of coups strongly suggests that a split-population approach to this question will be useful.
 They argue that many countries never experience coups because coups are effectvely impossible due to structural factors, while others that never experience coups are nevertheless at risk due to a different configuration of those same factors.
-Using language which fits nicely with the class of models described above, they say ``Triggers are not the source of the original risk, and in the absence of structural causes, the presence of triggering factors alone cannot lead to a coup.
+Using language which fits nicely with the class of models described above, they say "Triggers are not the source of the original risk, and in the absence of structural causes, the presence of triggering factors alone cannot lead to a coup.
 Hence, triggers should not be equated with coup risk.
-Rather, they are factors that may determine the exact timing of a coup in regimes that suffer from high coup risk'' \citep[p.\ 598][] {belkin2003toward}.
+Rather, they are factors that may determine the exact timing of a coup in regimes that suffer from high coup risk" [p.
+598 @belkin2003toward].
 Belkin and Schofer develop a measure of ``structural coup risk'' which incorporates factors such as the strength of democratic institutions and civil society, and a recent history of successful coups.
@@
 \end{verbatim}
+
 \normalsize
-The \texttt{add}\_\texttt{duration} function takes as input a data frame with a binary response variable (in this example, ``coup'') that measures the occurrence of the event, or failure, recorded over discrete time periods, and produces a data frame with a duration variable (named ``duration'') for each spell as well as a binary variable (``atrisk'') that is constant across the spell and indicates whether the spell ended in failure.
-These are then used as response variables for the formula in the \texttt{spdur} function.
+The \texttt{add}\_\texttt{duration} function takes as input a data frame with a binary response variable (in this example, "coup"") that measures the occurrence of the event, or failure, recorded over discrete time periods, and produces a data frame with a duration variable (named "duration") for each spell as well as a binary variable ("atrisk") that is constant across the spell and indicates whether the spell ended in failure.
+These are then used as response variables for the formula in the `spdur` function.
 Though the data used in this example are recorded annually, the function supports annual, monthly, or daily data.
@@
 \end{verbatim}
+
 \normalsize  
@@
 \end{verbatim}
+
 \normalsize
@@
 \end{tabular}
-\caption{Coup model with log-logistic hazard} \label{loglog_table}
 \end{table}
+
 \normalsize
@@
 \end{verbatim}
+
 \begin{figure}[htbp!]
 \centering
-\includegraphics[width=4in]{loglog_hazard.pdf}
-\caption{{\em Conditional hazard rate from log-logistic model.}} \label{hazard}
+\includegraphics[width=4in]{graphics/loglog_hazard.pdf}
+\caption{Conditional hazard rate from log-logistic model.} \label{hazard}
 \end{figure}
@@
 The generic function \texttt{predict} can be used on an object of class \texttt{spdur} to generate several kinds of predictions, including the probability that an observation is ``at-risk'' and the probability of failure for a given time period.
-In addition, the generic \texttt{plot} function can be used on \texttt{spdur} objects to produce a separation plot \citep{greenhill2011separation}, which is a graphical display for evaluating model predictions.
+In addition, the generic \texttt{plot} function can be used on \texttt{spdur} objects to produce a separation plot [@greenhill2011separation], which is a graphical display for evaluating model predictions.
 The code below produces Figure \ref{insamp}.
@@
 \centering
-\includegraphics[width=5in]{in-sample.pdf}
+\includegraphics[width=5in]{graphics/in-sample.pdf}
 \caption{{\em In-sample separation plots for Weibull and log-logistic models.}} \label{insamp}
@@
 \end{verbatim}
+
+![](out-of-sample.pdf)
+
 \normalsize
@@
 
-If information on previous failures is not available, it might be beneficial to drop some fraction of early observations for the same effect.
-For example, if the average failure times in the at risk sub-population are short relative to the time period over which data are available, it might be possible to eliminate most if not all left-censored spells without a large loss to the overall quantity of data.
-
 Right-censoring, where spells end before outcomes are observed, also pose a unique problem in the split-population framework.
@@
 immune spells.
-The \texttt{add\_duration()} function codes all observations in a spell as at risk if the spell itself ended in failure.
+The \texttt{add\_duration()} function retroactively codes all observations in a spell as at risk if the spell itself ended in failure.
 Right-censored spells are coded as immune over their entire duration.
-This is problematic since some right-censored spells will fail outside the observation window, meaning they have been misclassified as immune spells.
-To our knowledge there has been no systematic investigation of this issue and the extent to which it is problematic.
-The only, untested, recommendation we can provide is that in some circumstances it might be helpful to add additional criteria before considering a spell immune, e.g.
-a survival time threshold.
+This can lead to some misclassification of observations as immune even though they experience failure at some point in the unobserved future.
 
-The risk coding for right-censored spells is an important consideration for out-of-sample testing.
-A purely random sample of observations would include risk coding that is informed by observations that are potentially outside of the sample, and sampling, for example for cross-validation, should thus occur by spell.
-For block-wise testing, i.e.
-where some earlier data are used to train models and a later block of time is held back for out-of-sample testing, the \texttt{add\_duration()} function should be applied after partitioning the data, rather than to the complete data set before partitioning, in which case the risk coding for some spells in the training set might have been informed by outcomes in the test data.
+Furthermore, in out-of-sample testing based on some kind of data partitioning scheme, this coding scheme can lead to unintentioned contamination of in-sample cases with knowledge of out-of-sample failures through the risk coding for failed spells.
+This leads to two recommendations.
+First, the duration data (`add_duration`) should not be built until after partitioning the data.
+Secondly, and as a result of the first point, data should be partitioned in a scheme that samples either by spell or block-wise, e.g.
+by witholding the last **x** years of data, in order to be able to proberly build the duration variables.
 
-\section{Conclusion}
+# Conclusion
 
-\bibliography{spdur}
+```{r, include = FALSE}
+library(foreign)
+library(magrittr)
+library(ggplot2)
+library(xtable)
+library(scales)
 
-\end{document}
+fl2003 <- read.dta("data/repdata.dta")
+rates <- data.frame(
+  data = "Fearon and Laitin 2003",
+  rate = mean(fl2003$onset > 0)
+)
+
+load("data/irc_data_mod.rda")
+rates <- rbind(rates, data.frame(
+  data = "Beger et al 2015",
+  rate = mean(irc.data$irr.t)
+))
+
+belkin <- read.dta("data/BelkinSchoferTable4.dta")
+belkin$coup <- ifelse(belkin$coup=="yes", 1, 0)
+rates <- rbind(rates, data.frame(
+  data = "Belkin & Schofer 2003",
+  rate = mean(belkin$coup)
+))
+
+rates$rate <- rates$rate * 1000
+
+# Convert data to factor ordered by rate, so plot will sort by rate
+rates$data <- with(rates, factor(data, levels = data[order(rate)]))
+
+#xtable(rates, digits = 2)
+
+p <- ggplot(rates, aes(x = data, y = rate)) + 
+  geom_bar(stat = "identity", fill = muted("blue"), width = 0.6) +
+  coord_flip() +
+  theme_minimal() +
+  labs(y = "Positives per 1,000", x = "")
+
+ggsave(plot = p, file = "graphics/rates.pdf", height = 1.3, width = 5, scale = 1)
+```
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width = 4in]{graphics/rates.pdf}
+\caption{Rates of positive outcomes in select publications with binary outcomes}
+\label{rates}
+\end{figure}
+
+Data and outcomes in international relations, and to a lesser extent in other fields as well, are typically characterized by temporal dependence and, where a binary, ordinal, or count outcome is modelled, rare events or severely imbalanced classes with excessive numbers of zeros.
+Figure \ref{rates} shows a few examples for published research that models binary outcomes.
+
+This is a well-recognized problem and has led to the development or use of several specialized mixture models like zero-inflated Poisson and negative binomial regression for count data, and a zero-inflated ordered probit for ordinal outcomes [@bagozzi2015modeling].
+Split-population duration regression provides another principled solution to the challenges posed by data in this domain, but, unlike other solutions to the sparse outcome problem, also addresses underlying temporal dynamics.
+
+In addition to these technical features, split-population duration models are also appealing in the context of theories or arguments that explicitly distinguish long-term structural causes from short-term triggering events, as @belkin2003toward and others do, as this conceptual distinction can be accomodated through the seperate risk and duration equations.
+
+# References

### 2015-10-15 manuscript.tex
paper is in JSS style now

--- 2015-10-15 pa-letter.tex
@@
-\documentclass[]{article}
-\usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
+\documentclass[article]{jss}
 \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-    \usepackage{xltxtra,xunicode}
-  \else
-    \usepackage{fontspec}
-  \fi
-  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
-  \newcommand{\euro}{€}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
-\usepackage{microtype}
-\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\usepackage[margin=1.5in]{geometry}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\ifxetex
-  \usepackage[setpagesize=false, % page size defined by xetex
-              unicode=false, % unicode breaks when used with xetex
-              xetex]{hyperref}
-\else
-  \usepackage[unicode=true]{hyperref}
-\fi
-\hypersetup{breaklinks=true,
-            bookmarks=true,
-            pdfauthor={Daniel W. Hill, Nils W. Metternich, Andreas Beger, and Michael D. Ward},
-            pdftitle={Splitting it up: A split-duration package},
-            colorlinks=true,
-            citecolor=blue,
-            urlcolor=blue,
-            linkcolor=magenta,
-            pdfborder={0 0 0}}
-\urlstyle{same}  % don't use monospace font for urls
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-\setlength{\emergencystretch}{3em}  % prevent overfull lines
-\setcounter{secnumdepth}{0}
 
-%%% Use protect on footnotes to avoid problems with footnotes in titles
-\let\rmarkdownfootnote\footnote%
-\def\footnote{\protect\rmarkdownfootnote}
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
-%%% Change title format to be more compact
-\usepackage{titling}
+\author{
+Daniel W. Hill\\University of Georgia \And Nils W. Metternich\\University College London \And Andreas Beger\\who knows \And Michael D. Ward\\Duke University
+}
+\title{Splitting it up: the \pkg{spduration} split-population duration
+regression package}
+\Keywords{keywords, not capitalized, \proglang{R}}
 
-% Create subtitle command for use in maketitle
-\newcommand{\subtitle}[1]{
-  \posttitle{
-    \begin{center}\large#1\end{center}
-    }
+\Abstract{
+We present an implementation of split-population duration regression in
+the \pkg{spduration} R package and an application to data on military
+coups. The statistical model accounts for units that are immune to a
+certain outcome and not part of the duration process the researcher is
+primarily interested in. We provide insights that if immune units exist,
+we can significantly increase the predictive performance compared to
+standard duration model. The package includes estimation and forecasting
+methods for split-population Weibull and Loglogistic models.
 }
 
-\setlength{\droptitle}{-2em}
-  \title{Splitting it up: A split-duration package}
-  \pretitle{\vspace{\droptitle}\centering\huge}
-  \posttitle{\par}
-  \author{Daniel W. Hill, Nils W. Metternich, Andreas Beger, and Michael D. Ward}
-  \preauthor{\centering\large\emph}
-  \postauthor{\par}
-  \predate{\centering\large\emph}
-  \postdate{\par}
-  \date{June 8, 2015}
+\Plainauthor{Daniel W. Hill, Nils W. Metternich, Andreas Beger, Michael D. Ward}
+\Plaintitle{Splitting it up: the spduration split-population duration regression
+package}
+\Shorttitle{\pkg{spduration}: Split-population duration regression}
+\Plainkeywords{keywords, not capitalized, R}
+
+%% publication information
+%% \Volume{50}
+%% \Issue{9}
+%% \Month{June}
+%% \Year{2012}
+\Submitdate{}
+%% \Acceptdate{2012-06-04}
+
+\Address{
+    Daniel W. Hill\\
+  University of Georgia\\
+  First line Second line\\
+  E-mail: \href{mailto:name@company.com}{\nolinkurl{name@company.com}}\\
+  URL: \url{http://rstudio.com}\\~\\
+        }
+
+\usepackage{amsmath}
 
@@
 
-\maketitle
+\section{Introduction}\label{introduction}
 
-\begin{abstract}
-We present an implementation of split-population duration regression in the \texttt{spduration} R package and an application to data on military coups.
-The statistical model accounts for units that are immune to a certain outcome and not part of the duration process the researcher is primarily interested in.
-We provide insights that if immune units exist, we can significantly increase the predictive performance compared to standard duration model.
-The package includes estimation and forecasting methods for split-population Weibull and Loglogistic models.
-\end{abstract}
+This template demonstrates some of the basic latex you'll need to know
+to create a JSS article.
 
-\section{Introduction}\label{introduction}
+\subsection{Code formatting}\label{code-formatting}
+
+Don't use markdown, instead use the more precise latex commands:
+
+\begin{itemize}
+\itemsep1pt\parskip0pt\parsep0pt
+\item
+  \proglang{Java}
+\item
+  \pkg{plyr}
+\item
+  \code{print("abc")}
+\end{itemize}
+
+\section{Introduction}\label{introduction-1}
 
@@
 The last component to complete the likelihood is the choice of a
-distribution for the shape of the hazard rate. The \texttt{spduration}
+distribution for the shape of the hazard rate. The \pkg{spduration}
 package implements two hazard rate shapes,
@@
 \centering
-\includegraphics[width=4in]{loglog_hazard.pdf}
+\includegraphics[width=4in]{graphics/loglog_hazard.pdf}
 \caption{Conditional hazard rate from log-logistic model.} \label{hazard}

### 2015-10-15 old-stuff/pa-letter.Rmd
paper is in JSS style now

--- 2015-10-15 manuscript.Rmd
@@
 ---
-author:
-  - name: Daniel W. Hill
-    affiliation: University of Georgia
-    address: >
-      First line
-      Second line
-    email: name@company.com
-    url: http://rstudio.com
-  - name: Nils W. Metternich
-    affiliation: University College London
-  - name: Andreas Beger
-    affiliation: who knows
-  - name: Michael D. Ward
-    affiliation: Duke University
-title:
-  formatted: "Splitting it up: the \\pkg{spduration} split-population duration regression package"
-  # If you use tex in the formatted title, also supply version without
-  plain:     "Splitting it up: the spduration split-population duration regression package"
-  # For running headers, if needed
-  short:     "\\pkg{spduration}: Split-population duration regression"
-abstract: >
-  We present an implementation of split-population duration regression in the   \pkg{spduration} R package and an application to data on military coups.
+title: "Splitting it up: A split-duration package"
+author: "Daniel W. Hill, Nils W. Metternich, Andreas Beger, and Michael D. Ward"
+date: "June 8, 2015"
+output: 
+  pdf_document:
+    keep_tex: true
+bibliography: spdur.bib
+geometry: margin=1.5in
+---
+
+\begin{abstract}
+We present an implementation of split-population duration regression in the \texttt{spduration} R package and an application to data on military coups.
 The statistical model accounts for units that are immune to a certain outcome and not part of the duration process the researcher is primarily interested in.
@@
 The package includes estimation and forecasting methods for split-population Weibull and Loglogistic models.
-keywords:
-  # at least one keyword must be supplied
-  formatted: [keywords, not capitalized, "\\proglang{R}"]
-  plain:     [keywords, not capitalized, R]
-preamble: >
-  \usepackage{amsmath}
-output: 
-  rticles::jss_article
-bibliography: spdur.bib
----
-
-# Introduction
-
-This template demonstrates some of the basic latex you'll need to know to create a JSS article.
-
-## Code formatting
-
-Don't use markdown, instead use the more precise latex commands:
-
-* \proglang{Java}
-* \pkg{plyr}
-* \code{print("abc")}
+\end{abstract}
 
@@
 The last component to complete the likelihood is the choice of a distribution for the shape of the hazard rate.
-The \pkg{spduration} package implements two hazard rate shapes, Weibull\footnote{In the code, $f(t) = \alpha \lambda^{\alpha} t^{\alpha - 1} e^{-(\lambda t)^\alpha}$.} and log-logistic:
+The \texttt{spduration} package implements two hazard rate shapes, Weibull\footnote{In the code, $f(t) = \alpha \lambda^{\alpha} t^{\alpha - 1} e^{-(\lambda t)^\alpha}$.} and log-logistic:
 
@@
 \centering
-\includegraphics[width=4in]{graphics/loglog_hazard.pdf}
+\includegraphics[width=4in]{loglog_hazard.pdf}
 \caption{Conditional hazard rate from log-logistic model.} \label{hazard}
@@
 \centering
-\includegraphics[width=5in]{graphics/in-sample.pdf}
+\includegraphics[width=5in]{in-sample.pdf}
 \caption{{\em In-sample separation plots for Weibull and log-logistic models.}} \label{insamp}

### 2015-10-15 manuscript.Rmd
started working on replication script

@@
 We begin by loading the package and formatting the data appropriately. 
-\small
-\begin{verbatim}
-> library(spduration)
-> new.coup.data <- add_duration(coup.data, "coup", unitID="countryid", tID="year", 
-+   freq="year")
-\end{verbatim}
+```{r, eval=FALSE}
+library(spduration)
+data(coup.data)
+new.coup.data <- add_duration(coup.data, "coup", unitID="countryid", tID="year",
+                              freq="year")
+```
 
@@
 We begin by fitting first a Weibull and then a log-logistic split-population duration model using the coups data, including the measure of coup risk in the logit (risk) equation.
-\small
-\begin{verbatim}
-> weib.model <- spdur(duration~Military.Regime+Instability+Recent.War+Regional.Conflict,
-+ 	 	atrisk~Coup.Risk+GDP.cap.+Military.Regime+
-+ 	 	Recent.War+Regional.Conflict+
-+ 	 	South.Am.+Central.Am.,data=new.coup.data)
-> loglog.model <- spdur(duration~Military.Regime+Instability+Recent.War+Regional.Conflict,
-+  	 	atrisk~Coup.Risk+GDP.cap.+Military.Regime+
-+  	 	Recent.War+Regional.Conflict+
-+ 	 	South.Am.+Central.Am.,data=new.coup.data,distr="loglog")
-\end{verbatim}
 
-\normalsize  
-Using the \texttt{summary} function on either model object will produce standard output showing the model formula, estimates for the duration and risk equations, and test statistics with $p$-values.
-One may also use the \texttt{AIC} and \texttt{BIC} function to calculate the information criterion statistics for \texttt{spdur} objects.
+```{r models, eval=FALSE}
+weib_model <- spdur(
+  duration ~ milreg + instab + rwar + regconf,
+  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
+  data=bscoup)
+
+loglog_model <- spdur(
+  duration~Military.Regime+Instability+Recent.War+Regional.Conflict,
+  atrisk~Coup.Risk+GDP.cap.+Military.Regime+Recent.War+Regional.Conflict+
+South.Am.+Central.Am.,
+  data=bscoup, distr="loglog")
+```
+
+Using the \code{summary} function on either model object will produce standard output showing the model formula, estimates for the duration and risk equations, and test statistics with $p$-values.
+One may also use the \code{AIC} and \code{BIC} function to calculate the information criterion statistics for \code{spdur} objects.
 \small
@@
 
-```{r, include = FALSE}
+```{r rates, include = FALSE}
 library(foreign)

### 2015-10-15 manuscript.tex
started working on replication script

@@
 We begin by loading the package and formatting the data appropriately.
-\small
 
-\begin{verbatim}
-> library(spduration)
-> new.coup.data <- add_duration(coup.data, "coup", unitID="countryid", tID="year", 
-+   freq="year")
-\end{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
+library(spduration)
+data(coup.data)
+new.coup.data <- add_duration(coup.data, "coup", unitID="countryid", tID="year",
+                              freq="year")
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 split-population duration model using the coups data, including the
-measure of coup risk in the logit (risk) equation. \small
+measure of coup risk in the logit (risk) equation.
 
-\begin{verbatim}
-> weib.model <- spdur(duration~Military.Regime+Instability+Recent.War+Regional.Conflict,
-+       atrisk~Coup.Risk+GDP.cap.+Military.Regime+
-+       Recent.War+Regional.Conflict+
-+       South.Am.+Central.Am.,data=new.coup.data)
-> loglog.model <- spdur(duration~Military.Regime+Instability+Recent.War+Regional.Conflict,
-+       atrisk~Coup.Risk+GDP.cap.+Military.Regime+
-+       Recent.War+Regional.Conflict+
-+       South.Am.+Central.Am.,data=new.coup.data,distr="loglog")
-\end{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
+weib_model <- spdur(
+  duration ~ milreg + instab + rwar + regconf,
+  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
+  data=bscoup)
 
-\normalsize  
-Using the \texttt{summary} function on either model object will produce
+loglog_model <- spdur(
+  duration~Military.Regime+Instability+Recent.War+Regional.Conflict,
+  atrisk~Coup.Risk+GDP.cap.+Military.Regime+Recent.War+Regional.Conflict+
+South.Am.+Central.Am.,
+  data=bscoup, distr="loglog")
+\end{CodeInput}
+\end{CodeChunk}
+
+Using the \code{summary} function on either model object will produce
 standard output showing the model formula, estimates for the duration
 and risk equations, and test statistics with \(p\)-values. One may also
-use the \texttt{AIC} and \texttt{BIC} function to calculate the
-information criterion statistics for \texttt{spdur} objects.\\\small
+use the \code{AIC} and \code{BIC} function to calculate the information
+criterion statistics for \code{spdur} objects.\\\small
 

### 2015-10-15 replication.R

@@ -1,0 +1,9 @@
+#  Replication script for JSS manuscript examples
+
+
+library(spduration)
+#data(bscoup)
+bscoup <- read.dta("data/BelkinSchoferTable4.dta")
+bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
+bscoup <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
+                       freq="year")

### 2015-10-27 manuscript.Rmd
new graphics, writing is next

@@
 We begin by loading the package and formatting the data appropriately. 
-```{r, eval=FALSE}
-library(spduration)
-data(coup.data)
-new.coup.data <- add_duration(coup.data, "coup", unitID="countryid", tID="year",
+```{r data}
+library("spduration")
+#data(bscoup)
+library("foreign")
+bscoup <- read.dta("data/BelkinSchoferTable4.dta")
+bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
+bscoup      <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
                             freq="year")
@@
 
-```{r models, eval=FALSE}
+```{r models, include = FALSE}
 weib_model <- spdur(
-  duration ~ milreg + instab + rwar + regconf,
+  duration ~ milreg + instab + regconf,
   atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
-  data=bscoup)
+  data=bscoup, silent = TRUE)
 
 loglog_model <- spdur(
-  duration~Military.Regime+Instability+Recent.War+Regional.Conflict,
-  atrisk~Coup.Risk+GDP.cap.+Military.Regime+Recent.War+Regional.Conflict+
-South.Am.+Central.Am.,
-  data=bscoup, distr="loglog")
+  duration ~ milreg + instab + regconf,
+  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
+  data=bscoup, distr="loglog", silent = TRUE)
 ```
@@
 One may also use the \code{AIC} and \code{BIC} function to calculate the information criterion statistics for \code{spdur} objects.
-\small
-\begin{verbatim}
-> AIC(weib.model)
-[1] 1120.774
-> AIC(loglog.model)
-[1] 1111.908
-> BIC(weib.model)
-[1] 1209.577
-> BIC(loglog.model)
-[1] 1200.711
-\end{verbatim}
+```{r fit-stats}
+AIC(weib_model)
+AIC(loglog_model)
+BIC(weib_model)
+BIC(loglog_model)
+```
 
-\normalsize
 The objects produced by \texttt{spdur} can also be used with functions in the \texttt{xtable} package to produce \LaTeX\ code for tables of results.
 The \texttt{xtable} function used on an \texttt{spdur} object will produce code such as that used to create Table \ref{loglog_table}.
-\small
-\begin{verbatim}
-> xtable(loglog.model,caption="Coup model with log-logistic hazard")
-\end{verbatim}
+
+```{r summary-table, include=FALSE}
+library("xtable")
+tbl <- xtable(loglog_model, caption="Coup model with log-logistic hazard",
+              label="loglog_table")
+print(tbl, caption.placement="top", comment=FALSE, include.rownames=FALSE)
+```
 
@@
 \centering
-\caption{Coup model with log-logistic hazard} \label{loglog_table}
-\begin{tabular}{rlrrr}
+\caption{Coup model with log-logistic hazard} 
+\label{loglog_table}
+\begin{tabular}{lrrr}
   \hline
- & Parameter & Estimate & StdErr & p \\ 
+Parameter & Estimate & StdErr & p \\ 
   \hline
-1 & (Dur. Intercept) & 3.29 & 0.14 & 0.00 \\ 
-  2 & Military.Regime & -1.51 & 0.16 & 0.00 \\ 
-  3 & Instability & -0.21 & 0.04 & 0.00 \\ 
-  4 & Recent.War & -0.20 & 0.28 & 0.48 \\ 
-  5 & Regional.Conflict & 2.64 & 2.56 & 0.30 \\ 
-  6 & log(alpha) & -0.53 & 0.07 & 0.00 \\ 
-  7 & (Risk Intercept) & -13.10 & 10.03 & 0.19 \\ 
-  8 & Coup.Risk & 7.32 & 3.23 & 0.02 \\ 
-  9 & GDP.cap. & 3.11 & 1.82 & 0.09 \\ 
-  10 & Military.Regime.1 & 25.41 & 0.01 & 0.00 \\ 
-  11 & Recent.War.1 & -4.34 & 2.70 & 0.11 \\ 
-  12 & Regional.Conflict.1 & -39.74 & 23.96 & 0.10 \\ 
-  13 & South.Am. & -3.72 & 3.19 & 0.24 \\ 
-  14 & Central.Am. & -6.08 & 2.45 & 0.01 \\ 
+(Dur. Intercept) & 3.30 & 0.14 & 0.00 \\ 
+  milreg & -1.53 & 0.15 & 0.00 \\ 
+  instab & -0.22 & 0.03 & 0.00 \\ 
+  regconf & 3.04 & 2.22 & 0.17 \\ 
+  log(alpha) & -0.52 & 0.07 & 0.00 \\ 
+  (Risk Intercept) & -14.61 & 9.02 & 0.11 \\ 
+  couprisk & 9.11 & 4.37 & 0.04 \\ 
+  wealth & 3.51 & 1.75 & 0.05 \\ 
+  milreg.1 & 21.76 & 0.24 & 0.00 \\ 
+  rwar & -6.37 & 5.38 & 0.24 \\ 
+  regconf.1 & -20.22 & 54.04 & 0.71 \\ 
+  samerica & -3.31 & 3.28 & 0.31 \\ 
+  camerica & -6.46 & 2.57 & 0.01 \\ 
    \hline
@@
 
-\normalsize
 The table shows estimates from the duration equation, beginning with the intercept, and then estimates from the risk equation.
@@
 By default the function uses the mean values of the covariates during the simulations, but users can choose specific covariate values by entering them as vectors in the arguments \texttt{xvals} and \texttt{zvals}, which correspond to the covariates in the duration and risk equations, respectively.
-The command below creates Figure \ref{hazard}.
-\begin{verbatim}
-> plot.hazard(loglog.model)
-\end{verbatim}
+The command below creates Figure \ref{plot-hazard2}.
 
-\begin{figure}[htbp!]
-\centering
-\includegraphics[width=4in]{graphics/loglog_hazard.pdf}
-\caption{Conditional hazard rate from log-logistic model.} \label{hazard}
+```{r hazard-ex-1, eval = FALSE}
+plot(loglog_model, type="hazard", main = "A")
+```
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.8\textwidth]{graphics/hazard-ex.pdf}
+\end{center}
 \end{figure}
+
+```{r hazard-ex-2, eval = FALSE}
+plot(loglog_model, type="hazard", 
+     xvals = c(1, 1, 10, 0.05), 
+     zvals = c(1, 7, 8.64, 1, 1, 0.05, 0, 0),
+     main = "B")
+```
 
@@
 The code below produces Figure \ref{insamp}.
-\small
-\begin{verbatim}
-> dev.new(width=9,height=3)
-> par(mfrow=c(2,1),mar=c(2,2,2,2))
-> plot(weib.model,endSpellOnly=F)
-> plot(loglog.model,endSpellOnly=F)
-\end{verbatim}
 
-\normalsize
+```{r, eval = FALSE}
+plot(weib_model)
+plot(loglog_model)
+```
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.8\textwidth]{graphics/sepplots.pdf}
+\end{center}
+\end{figure}
+
 The option \texttt{endSpellOnly} is set to \texttt{FALSE} so that every observation, not only those at the end of a spell, is used in the plot.
@@
 This makes it easy to see whether the model is assigning high probabilities of failure to actual cases of failure, and low probabilities to non-failures.
-\begin{figure}[htbp!]
-\centering
-\includegraphics[width=5in]{graphics/in-sample.pdf}
-\caption{{\em In-sample separation plots for Weibull and log-logistic models.}} \label{insamp}
-\end{figure}
 
-Finally, we demonstrate use of the \texttt{spdurCrisp} function for evaluating out-of-sample predictions.
+Finally, we demonstrate how to evaluate a model's out of sample predictions.
 We begin by splitting the data into training and test sets, using the last year of data (1999) as the test set and omitting incomplete observations from the test set.
-\small
-\begin{verbatim}
-> coup.train<-new.coup.data[coup.data$year<1999,]
-> coup.test<-new.coup.data[new.coup.data$year==1999,]
-> coup.test<-na.omit(coup.test)
-\end{verbatim}
+```{r}
+data(bscoup)
+bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
+coup_train <- bscoup[bscoup$year < 1996, ]
+coup_train <- add_duration(coup_train, "coup", unitID="countryid", tID="year",
+                          freq="year")
+```
+
+```{r}
+coup_test  <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
+                           freq="year")
+coup_test  <- coup_test[coup_test$year >= 1996, ]
+```
 
@@
 The \texttt{spdurCrisp} function can then be used to fit a model on the training set and generate predictions for the test set.
-\small
-\begin{verbatim}
-> weib.model2<-spdurCrisp(duration~Military.Regime+Instability+Recent.War+Regional.Conflict,
-+ 		atrisk~Coup.Risk+GDP.cap.+Military.Regime+Recent.War+Regional.Conflict+
-+ 		South.Am.+Central.Am.,
-+ 		last='end.spell',train=coup.train,test=coup.test,pred=coup.test)
-> weib.preds<-as.numeric(weib.model2$test.p)
->
-> loglog.model2<-spdurCrisp(duration~Military.Regime+Instability+Recent.War+Regional.Conflict,
-+ 		atrisk~Coup.Risk+GDP.cap.+Military.Regime+Recent.War+Regional.Conflict+
-+ 		South.Am.,
-+ 		last='end.spell',train=coup.train,test=coup.test,pred=coup.test,distr="loglog")
-> loglog.preds<-as.numeric(loglog.model2$test.p)
-\end{verbatim}
+
+```{r}
+weib_model2   <- spdur(
+  duration ~ milreg + instab + regconf,
+  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
+  data = coup_train, silent = TRUE)
+
+loglog_model2 <- spdur(
+  duration ~ milreg + instab + regconf,
+  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
+  data = coup_train, distr="loglog", silent = TRUE) 
+```
+
+Foo
+
+```{r}
+weib2_test_p   <- predict(weib_model2, newdata = coup_test)
+loglog2_test_p <- predict(loglog_model2, newdata = coup_test)
+```
 
@@
 The test set predictions are stored in the model object as \texttt{test.p}, and we extract them to produce the plot shown below in Figure \ref{outsamp}, which was created using the \texttt{separationplot} package.
-\small
-\begin{verbatim}
-> dev.new(width=9,height=3)
-> par(mfrow=c(2,1),mar=c(2,2,2,2))
-> separationplot(weib.preds,coup.test$Coup,newplot=F,show.expected=T,lwd1=5,lwd2=2)
-> separationplot(loglog.preds,coup.test$Coup,newplot=F,show.expected=T,lwd1=5,lwd2=2)
-\end{verbatim}
 
-![](out-of-sample.pdf)
+```{r oos-sepplots}
+library("separationplot")
 
-\normalsize
-\begin{figure}[htbp!]
-\centering
-\includegraphics[width=5in]{out-of-sample.pdf}
-\caption{{\em Out-of-sample separation plots for Weibull and log-logistic models. }} \label{outsamp}
+obs_y <- coup_test$coup[complete.cases(coup_test)]
+
+par(mfrow=c(2,1),mar=c(2,2,2,2))
+separationplot(weib2_test_p,   obs_y)
+separationplot(loglog2_test_p, obs_y)
+```
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.8\textwidth]{graphics/oos-sepplots.pdf}
+\end{center}
 \end{figure}
+
+```{r}
+glm_model2    <- glm(
+  coup ~ milreg + instab + regconf + couprisk + wealth + rwar + regconf + 
+    samerica + camerica,
+  data = coup_train, family = "binomial"
+)
+
+c(Weibull = AIC(weib_model2), 
+  Loglog = AIC(loglog_model2), 
+  Logistic = AIC(glm_model2))
+```
 
@@
 
-```{r rates, include = FALSE}
-library(foreign)
-library(magrittr)
-library(ggplot2)
-library(xtable)
-library(scales)
-
-fl2003 <- read.dta("data/repdata.dta")
-rates <- data.frame(
-  data = "Fearon and Laitin 2003",
-  rate = mean(fl2003$onset > 0)
-)
-
-load("data/irc_data_mod.rda")
-rates <- rbind(rates, data.frame(
-  data = "Beger et al 2015",
-  rate = mean(irc.data$irr.t)
-))
-
-belkin <- read.dta("data/BelkinSchoferTable4.dta")
-belkin$coup <- ifelse(belkin$coup=="yes", 1, 0)
-rates <- rbind(rates, data.frame(
-  data = "Belkin & Schofer 2003",
-  rate = mean(belkin$coup)
-))
-
-rates$rate <- rates$rate * 1000
-
-# Convert data to factor ordered by rate, so plot will sort by rate
-rates$data <- with(rates, factor(data, levels = data[order(rate)]))
-
-#xtable(rates, digits = 2)
-
-p <- ggplot(rates, aes(x = data, y = rate)) + 
-  geom_bar(stat = "identity", fill = muted("blue"), width = 0.6) +
-  coord_flip() +
-  theme_minimal() +
-  labs(y = "Positives per 1,000", x = "")
-
-ggsave(plot = p, file = "graphics/rates.pdf", height = 1.3, width = 5, scale = 1)
-```
-
 \begin{figure}[htbp!]

### 2015-10-27 manuscript.tex
new graphics, writing is next

@@
 \begin{CodeInput}
-library(spduration)
-data(coup.data)
-new.coup.data <- add_duration(coup.data, "coup", unitID="countryid", tID="year",
+library("spduration")
+#data(bscoup)
+library("foreign")
+bscoup <- read.dta("data/BelkinSchoferTable4.dta")
+bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
+bscoup      <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
                             freq="year")
 \end{CodeInput}
+\begin{CodeOutput}
+Warning in attempt_date(data[, tID], freq): Converting to 'Date' class with
+yyyy-06-30
+\end{CodeOutput}
 \end{CodeChunk}
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-weib_model <- spdur(
-  duration ~ milreg + instab + rwar + regconf,
-  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
-  data=bscoup)
-
-loglog_model <- spdur(
-  duration~Military.Regime+Instability+Recent.War+Regional.Conflict,
-  atrisk~Coup.Risk+GDP.cap.+Military.Regime+Recent.War+Regional.Conflict+
-South.Am.+Central.Am.,
-  data=bscoup, distr="loglog")
-\end{CodeInput}
-\end{CodeChunk}
-
 Using the \code{summary} function on either model object will produce
@@
 use the \code{AIC} and \code{BIC} function to calculate the information
-criterion statistics for \code{spdur} objects.\\\small
+criterion statistics for \code{spdur} objects.
 
-\begin{verbatim}
-> AIC(weib.model)
-[1] 1120.774
-> AIC(loglog.model)
-[1] 1111.908
-> BIC(weib.model)
-[1] 1209.577
-> BIC(loglog.model)
-[1] 1200.711
-\end{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
+AIC(weib_model)
+\end{CodeInput}
+\begin{CodeOutput}
+[1] 1207.992
+\end{CodeOutput}
+\begin{CodeInput}
+AIC(loglog_model)
+\end{CodeInput}
+\begin{CodeOutput}
+[1] 1110.829
+\end{CodeOutput}
+\begin{CodeInput}
+BIC(weib_model)
+\end{CodeInput}
+\begin{CodeOutput}
+[1] 1290.452
+\end{CodeOutput}
+\begin{CodeInput}
+BIC(loglog_model)
+\end{CodeInput}
+\begin{CodeOutput}
+[1] 1193.289
+\end{CodeOutput}
+\end{CodeChunk}
 
-\normalsize
 The objects produced by \texttt{spdur} can also be used with functions
@@
 will produce code such as that used to create Table \ref{loglog_table}.
-\small
-
-\begin{verbatim}
-> xtable(loglog.model,caption="Coup model with log-logistic hazard")
-\end{verbatim}
 
@@
 
-\normalsize
 The table shows estimates from the duration equation, beginning with the
@@
 covariates in the duration and risk equations, respectively. The command
-below creates Figure \ref{hazard}.
+below creates Figure \ref{plot-hazard2}.
 
-\begin{verbatim}
-> plot.hazard(loglog.model)
-\end{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
+plot(loglog_model, type="hazard", main = "A")
+\end{CodeInput}
+\end{CodeChunk}
 
-\begin{figure}[htbp!]
-\centering
-\includegraphics[width=4in]{graphics/loglog_hazard.pdf}
-\caption{Conditional hazard rate from log-logistic model.} \label{hazard}
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.8\textwidth]{graphics/hazard-ex.pdf}
+\end{center}
 \end{figure}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+plot(loglog_model, type="hazard", 
+     xvals = c(1, 1, 10, 0.05), 
+     zvals = c(1, 7, 8.64, 1, 1, 0.05, 0, 0),
+     main = "B")
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 graphical display for evaluating model predictions. The code below
-produces Figure \ref{insamp}. \small
+produces Figure \ref{insamp}.
 
-\begin{verbatim}
-> dev.new(width=9,height=3)
-> par(mfrow=c(2,1),mar=c(2,2,2,2))
-> plot(weib.model,endSpellOnly=F)
-> plot(loglog.model,endSpellOnly=F)
-\end{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
+plot(weib_model)
+plot(loglog_model)
+\end{CodeInput}
+\end{CodeChunk}
 
-\normalsize
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.8\textwidth]{graphics/sepplots.pdf}
+\end{center}
+\end{figure}
+
 The option \texttt{endSpellOnly} is set to \texttt{FALSE} so that every
@@
 
-\begin{figure}[htbp!]
-\centering
-\includegraphics[width=5in]{graphics/in-sample.pdf}
-\caption{{\em In-sample separation plots for Weibull and log-logistic models.}} \label{insamp}
-\end{figure}
+Finally, we demonstrate how to evaluate a model's out of sample
+predictions. We begin by splitting the data into training and test sets,
+using the last year of data (1999) as the test set and omitting
+incomplete observations from the test set.
 
-Finally, we demonstrate use of the \texttt{spdurCrisp} function for
-evaluating out-of-sample predictions. We begin by splitting the data
-into training and test sets, using the last year of data (1999) as the
-test set and omitting incomplete observations from the test set. \small
+\begin{CodeChunk}
+\begin{CodeInput}
+data(bscoup)
+bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
+coup_train <- bscoup[bscoup$year < 1996, ]
+coup_train <- add_duration(coup_train, "coup", unitID="countryid", tID="year",
+                          freq="year")
+\end{CodeInput}
+\begin{CodeOutput}
+Warning in attempt_date(data[, tID], freq): Converting to 'Date' class with
+yyyy-06-30
+\end{CodeOutput}
+\end{CodeChunk}
 
-\begin{verbatim}
-> coup.train<-new.coup.data[coup.data$year<1999,]
-> coup.test<-new.coup.data[new.coup.data$year==1999,]
-> coup.test<-na.omit(coup.test)
-\end{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
+coup_test  <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
+                           freq="year")
+\end{CodeInput}
+\begin{CodeOutput}
+Warning in attempt_date(data[, tID], freq): Converting to 'Date' class with
+yyyy-06-30
+\end{CodeOutput}
+\begin{CodeInput}
+coup_test  <- coup_test[coup_test$year >= 1996, ]
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 The \texttt{spdurCrisp} function can then be used to fit a model on the
-training set and generate predictions for the test set. \small
+training set and generate predictions for the test set.
 
-\begin{verbatim}
-> weib.model2<-spdurCrisp(duration~Military.Regime+Instability+Recent.War+Regional.Conflict,
-+       atrisk~Coup.Risk+GDP.cap.+Military.Regime+Recent.War+Regional.Conflict+
-+       South.Am.+Central.Am.,
-+       last='end.spell',train=coup.train,test=coup.test,pred=coup.test)
-> weib.preds<-as.numeric(weib.model2$test.p)
->
-> loglog.model2<-spdurCrisp(duration~Military.Regime+Instability+Recent.War+Regional.Conflict,
-+       atrisk~Coup.Risk+GDP.cap.+Military.Regime+Recent.War+Regional.Conflict+
-+       South.Am.,
-+       last='end.spell',train=coup.train,test=coup.test,pred=coup.test,distr="loglog")
-> loglog.preds<-as.numeric(loglog.model2$test.p)
-\end{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
+weib_model2   <- spdur(
+  duration ~ milreg + instab + regconf,
+  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
+  data = coup_train, silent = TRUE)
+
+loglog_model2 <- spdur(
+  duration ~ milreg + instab + regconf,
+  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
+  data = coup_train, distr="loglog", silent = TRUE) 
+\end{CodeInput}
+\end{CodeChunk}
+
+Foo
+
+\begin{CodeChunk}
+\begin{CodeInput}
+weib2_test_p   <- predict(weib_model2, newdata = coup_test)
+loglog2_test_p <- predict(loglog_model2, newdata = coup_test)
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 produce the plot shown below in Figure \ref{outsamp}, which was created
-using the \texttt{separationplot} package. \small
+using the \texttt{separationplot} package.
 
-\begin{verbatim}
-> dev.new(width=9,height=3)
-> par(mfrow=c(2,1),mar=c(2,2,2,2))
-> separationplot(weib.preds,coup.test$Coup,newplot=F,show.expected=T,lwd1=5,lwd2=2)
-> separationplot(loglog.preds,coup.test$Coup,newplot=F,show.expected=T,lwd1=5,lwd2=2)
-\end{verbatim}
+\textbackslash{}begin\{CodeChunk\} \textbackslash{}begin\{CodeInput\}
+library(``separationplot'')
 
-\includegraphics{out-of-sample.pdf}
+obs\_y \textless{}- coup\_test\$coup{[}complete.cases(coup\_test){]}
 
-\normalsize
+par(mfrow=c(2,1),mar=c(2,2,2,2)) separationplot(weib2\_test\_p, obs\_y)
+separationplot(loglog2\_test\_p, obs\_y)
+\textbackslash{}end\{CodeInput\} \textbackslash{}end\{CodeChunk\}
 
-\begin{figure}[htbp!]
-\centering
-\includegraphics[width=5in]{out-of-sample.pdf}
-\caption{{\em Out-of-sample separation plots for Weibull and log-logistic models. }} \label{outsamp}
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.8\textwidth]{graphics/oos-sepplots.pdf}
+\end{center}
 \end{figure}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+glm_model2    <- glm(
+  coup ~ milreg + instab + regconf + couprisk + wealth + rwar + regconf + 
+    samerica + camerica,
+  data = coup_train, family = "binomial"
+)
+
+c(Weibull = AIC(weib_model2), 
+  Loglog = AIC(loglog_model2), 
+  Logistic = AIC(glm_model2))
+\end{CodeInput}
+\begin{CodeOutput}
+  Weibull    Loglog  Logistic 
+ 990.5818  985.2633 1171.3556 
+\end{CodeOutput}
+\end{CodeChunk}
 

### 2015-10-27 replication.R

@@ -1,10 +1,151 @@
 #  Replication script for JSS manuscript examples
 
+library("spduration")
+library(foreign)
+library(magrittr)
+library(ggplot2)
+library(xtable)
+library(scales)
 
-library(spduration)
-#data(bscoup)
-bscoup <- read.dta("data/BelkinSchoferTable4.dta")
+
+data(bscoup)
 bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
 bscoup <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
                        freq="year")
 
+weib_model <- spdur(
+  duration ~ milreg + instab + regconf,
+  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
+  data=bscoup)
+
+loglog_model <- spdur(
+  duration ~ milreg + instab + regconf,
+  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
+  data=bscoup, distr="loglog")
+
+AIC(weib_model)
+AIC(loglog_model)
+BIC(weib_model)
+BIC(loglog_model)
+
+library("xtable")
+tbl <- xtable(loglog_model, caption="Coup model with log-logistic hazard",
+              label="loglog_table")
+print(tbl, caption.placement="top", comment=FALSE, include.rownames=FALSE)
+
+#   Plotting functions
+#   ____________________
+
+pdf("graphics/hazard-ex.pdf", height=4, width=8)
+par(mfrow = c(1, 2))
+plot(loglog_model, type="hazard", main = "A")
+plot(loglog_model, type="hazard", 
+     xvals = c(1, 1, 10, 0.05), 
+     zvals = c(1, 7, 8.64, 1, 1, 0.05, 0, 0),
+     main = "B")
+dev.off()
+
+pdf("graphics/sepplots.pdf", height=4, width=10)
+par(mfrow=c(2,1),mar=c(2,2,2,2))
+plot(weib_model, type="sepplot", lwd1 = 1, lwd2 = 1)
+title(main="Weibull")
+plot(loglog_model, type="sepplot", lwd1 = 1, lwd2 = 1)
+title(main="Loglog")
+dev.off()
+
+pdf("graphics/hazard-rates.pdf", height=4, width=8)
+par(mfrow=c(1, 2))
+plot(weib_model,   type="hazard", main="Weibull")
+plot(loglog_model, type="hazard", main="Loglog")
+dev.off()
+
+
+#   Out-of-sample testing
[…]

### 2015-10-28 manuscript.Rmd
writinating the manuscripter

@@
       Second line
-    email: name@company.com
-    url: http://rstudio.com
+    email: dwhill@uga.edu
+    url: http://spia.uga.edu/directory/faculty_staff/hill-daniel
   - name: Nils W. Metternich
@@
   - name: Andreas Beger
-    affiliation: who knows
+    affiliation: Duke University
   - name: Michael D. Ward
@@
 
-The `spduration` package implements a split-population duration regression for data with time-varying covariates and with two options for the hazard rate shape, the Weibull and log-logistic.\footnote{The Weibull allows for monotonically increasing or decreasing hazard rates, while the log-logistic allows for rates that first increase and then decrease.} The primary function, \texttt{spdur}, produces a regression model object of class \texttt{spdur} which can then be used with further methods.
+The \pkg{spduration} package for \proglang{R} implements a split-population duration regression for data with time-varying covariates and with two options for the hazard rate shape, Weibull and log-logistic.
+The Weibull density allows for hazard rates that are increasing, constant, or decreasing over survival time, while the log-logistic density also can fit rates that have a peak at a particular survival time.
 
-To illustrate the usefulness of the package and the broad applicability of the methods outlined above, we employ an example from political science.
-We replicate and extend the analysis presented in @belkin2003toward, which examines factors that lead to military coups.
-Belkin and Schofer's distinction between "structural" vs.
-"triggering"" causes of coups strongly suggests that a split-population approach to this question will be useful.
+As an example to illustrate the package functionality, we will replicate and extend the model of coup onsets in @belkin2003toward.
+Belkin and Schofer's paper lends itself to re-analysis with a split-populuation duration model because they explicitly distinguish long-term structural risk factors for coups from more short-term triggering causes that can explain the timing of a coup in a regime at risk.
 They argue that many countries never experience coups because coups are effectvely impossible due to structural factors, while others that never experience coups are nevertheless at risk due to a different configuration of those same factors.
@@
 Rather, they are factors that may determine the exact timing of a coup in regimes that suffer from high coup risk" [p.
-598 @belkin2003toward].
+598 @belkin2003toward]
+
 Belkin and Schofer develop a measure of ``structural coup risk'' which incorporates factors such as the strength of democratic institutions and civil society, and a recent history of successful coups.
@@
 
-We begin by loading the package and formatting the data appropriately. 
+We begin by loading the package and the Belkin and Schofer replication data, and formatting the data to add several variables needed by the split-population duration model.
+
 ```{r data}
@@
 \normalsize
-The \texttt{add}\_\texttt{duration} function takes as input a data frame with a binary response variable (in this example, "coup"") that measures the occurrence of the event, or failure, recorded over discrete time periods, and produces a data frame with a duration variable (named "duration") for each spell as well as a binary variable ("atrisk") that is constant across the spell and indicates whether the spell ended in failure.
+The \code{add\_duration} function takes as input a data frame with a binary response variable (in this example, "coup"") that measures the occurrence of the event, or failure, recorded over discrete time periods, and produces a data frame with a duration variable (named "duration") for each spell as well as a binary variable ("atrisk") that is constant across the spell and indicates whether the spell ended in failure.
 These are then used as response variables for the formula in the `spdur` function.
+The \code{spdur} function is the primary function in the package and produces a regression model object of class \code{spdur} which can then be used with further methods.
+
 Though the data used in this example are recorded annually, the function supports annual, monthly, or daily data.
@@
 
-The objects produced by \texttt{spdur} can also be used with functions in the \texttt{xtable} package to produce \LaTeX\ code for tables of results.
-The \texttt{xtable} function used on an \texttt{spdur} object will produce code such as that used to create Table \ref{loglog_table}.
+The objects produced by \code{spdur} can also be used with functions in the \code{xtable} package to produce \LaTeX\ code for tables of results.
+The \code{xtable} function used on an \code{spdur} object will produce code such as that used to create Table \ref{loglog_table}.
 
@@
 
-The package also includes a function, \texttt{plot.hazard}, that can be used with an \texttt{spdur} object to plot the estimated hazard rate.
-\texttt{plot.hazard} plots the conditional hazard, which is the probability of survival conditional on the covariates in the risk and duration equations, and conditional on survival up to time $t$.
+The package also includes a function, \code{plot_hazard}, that can be used with an \code{spdur} object to plot the estimated hazard rate.
+\code{plot_hazard} plots the conditional hazard, which is the probability of survival conditional on the covariates in the risk and duration equations, and conditional on survival up to time $t$.
 The function calculates the estimated hazard rate as well as 90\% confidence intervals, which are produced by simulating values from the estimated sampling distributions of the model parameters.
-By default the function uses the mean values of the covariates during the simulations, but users can choose specific covariate values by entering them as vectors in the arguments \texttt{xvals} and \texttt{zvals}, which correspond to the covariates in the duration and risk equations, respectively.
-The command below creates Figure \ref{plot-hazard2}.
+By default the function uses the mean values of the covariates during the simulations, but users can choose specific covariate values by entering them as vectors in the arguments \code{xvals} and \code{zvals}, which correspond to the covariates in the duration and risk equations, respectively.
+The command below creates the graph A in Figure \ref{hazard-ex}.
 
@@
 \includegraphics[width=0.8\textwidth]{graphics/hazard-ex.pdf}
+\caption{Plots of the hazard rate for the log-logistic model of coups.
+The left graph uses the default mean values for covariates, while graph B uses user-specified variable values for a high-risk military regime.}
+\label{hazard-ex}
 \end{center}
@@
 Importantly, the package also includes functions to evaluate model predictions.
-The generic function \texttt{predict} can be used on an object of class \texttt{spdur} to generate several kinds of predictions, including the probability that an observation is ``at-risk'' and the probability of failure for a given time period.
-In addition, the generic \texttt{plot} function can be used on \texttt{spdur} objects to produce a separation plot [@greenhill2011separation], which is a graphical display for evaluating model predictions.
+The generic function \code{predict} can be used on an object of class \code{spdur} to generate several kinds of predictions, including the probability that an observation is ``at-risk'' and the probability of failure for a given time period.
+In addition, the generic \code{plot} function can be used on \code{spdur} objects to produce a separation plot [@greenhill2011separation], which is a graphical display for evaluating model predictions.
 The code below produces Figure \ref{insamp}.
@@
 \includegraphics[width=0.8\textwidth]{graphics/sepplots.pdf}
+\caption{In-sample separation plots of Weibull and log-logistic model conditional hazard predictions}
+\label{insamp}
 \end{center}
@@
 
-The option \texttt{endSpellOnly} is set to \texttt{FALSE} so that every observation, not only those at the end of a spell, is used in the plot.
-By default the \texttt{plot} function will calculate the conditional hazard for each observation.
+The option \code{endSpellOnly} is set to \code{FALSE} so that every observation, not only those at the end of a spell, is used in the plot.
+By default the \code{plot} function will calculate the conditional hazard for each observation.
 The separation plot sorts observations from left to right according to the predicted probability assigned by the model (higher values to the right), and shows each event/failure as red line, with non-events shown in beige.
@@
 \normalsize
-The \texttt{spdurCrisp} function can then be used to fit a model on the training set and generate predictions for the test set.
+The \code{spdurCrisp} function can then be used to fit a model on the training set and generate predictions for the test set.
 
@@
 \normalsize
-The \texttt{train},\texttt{test}, and \texttt{pred} arguments are used to specify data frames for the training set, test set, and forecasting set.
+The \code{train},\code{test}, and \code{pred} arguments are used to specify data frames for the training set, test set, and forecasting set.
 We do not wish to generate forecasts here, so we simply specify the same data frame for the test and forecasting set.
-Here we use \texttt{spdurCrisp} to calculate the conditional risk, which is the default calculation for this function and is the probability that an observation is in the at-risk population.
-The test set predictions are stored in the model object as \texttt{test.p}, and we extract them to produce the plot shown below in Figure \ref{outsamp}, which was created using the \texttt{separationplot} package.
+Here we use \code{spdurCrisp} to calculate the conditional risk, which is the default calculation for this function and is the probability that an observation is in the at-risk population.
+The test set predictions are stored in the model object as \code{test.p}, and we extract them to produce the plot shown below in Figure \ref{outsamp}, which was created using the \code{separationplot} package.
 
@@
 
-Sometimes information about previous failures in the data is available beyond the time period over which covariates are observed, making it possible to ameliorate or eliminate left-censoring by using the information of previous failures when constructing the necessary duration variables with \texttt{add\_duration()}.
+Sometimes information about previous failures in the data is available beyond the time period over which covariates are observed, making it possible to ameliorate or eliminate left-censoring by using the information of previous failures when constructing the necessary duration variables with \code{add\_duration()}.
 
@@
 immune spells.
-The \texttt{add\_duration()} function retroactively codes all observations in a spell as at risk if the spell itself ended in failure.
+The \code{add\_duration()} function retroactively codes all observations in a spell as at risk if the spell itself ended in failure.
 Right-censored spells are coded as immune over their entire duration.
@@
 
-Data and outcomes in international relations, and to a lesser extent in other fields as well, are typically characterized by temporal dependence and, where a binary, ordinal, or count outcome is modelled, rare events or severely imbalanced classes with excessive numbers of zeros.
+Many outcomes in the area that we work in, political science and political violence more specifically, are rare events.
+Coups, the onset of war and civil war, mass killings, and so on are exceedingly rare events when one considers that most countries most of the time are stable.
 Figure \ref{rates} shows a few examples for published research that models binary outcomes.
+@fearon2003ethnicity is a very widely cited study of civil war onset that uses yearly observations of all major countries, and @beger2015ensemble is an example of the positive rates when moving to monthly data for a similar set of countries.
 
+In the language of machine learning, we are dealing with highly imbalanced classes, although our data are also highly structured with clear dependence over space and time in countries and societies that may experience political violence.
 This is a well-recognized problem and has led to the development or use of several specialized mixture models like zero-inflated Poisson and negative binomial regression for count data, and a zero-inflated ordered probit for ordinal outcomes [@bagozzi2015modeling].
@@
 
-In addition to these technical features, split-population duration models are also appealing in the context of theories or arguments that explicitly distinguish long-term structural causes from short-term triggering events, as @belkin2003toward and others do, as this conceptual distinction can be accomodated through the seperate risk and duration equations.
+Split-population duration models are not only appealing in this technical sense, but they also match the logic or intuituion many analysts of politics use when they distinguish long-term risk factors from more fleeting triggering causes.
+The example we have used, @belkin2003toward, is a particularly clear illustration of how well the language of theorists maps onto the model intuition.
 

### 2015-10-28 manuscript.tex
writinating the manuscripter

@@
 \author{
-Daniel W. Hill\\University of Georgia \And Nils W. Metternich\\University College London \And Andreas Beger\\who knows \And Michael D. Ward\\Duke University
+Daniel W. Hill\\University of Georgia \And Nils W. Metternich\\University College London \And Andreas Beger\\Duke University \And Michael D. Ward\\Duke University
 }
@@
   First line Second line\\
-  E-mail: \href{mailto:name@company.com}{\nolinkurl{name@company.com}}\\
-  URL: \url{http://rstudio.com}\\~\\
+  E-mail: \href{mailto:dwhill@uga.edu}{\nolinkurl{dwhill@uga.edu}}\\
+  URL: \url{http://spia.uga.edu/directory/faculty_staff/hill-daniel}\\~\\
         }
@@
 
-The \texttt{spduration} package implements a split-population duration
-regression for data with time-varying covariates and with two options
-for the hazard rate shape, the Weibull and
-log-logistic.\footnote{The Weibull allows for monotonically increasing or decreasing hazard rates, while the log-logistic allows for rates that first increase and then decrease.}
-The primary function, \texttt{spdur}, produces a regression model object
-of class \texttt{spdur} which can then be used with further methods.
+The \pkg{spduration} package for \proglang{R} implements a
+split-population duration regression for data with time-varying
+covariates and with two options for the hazard rate shape, Weibull and
+log-logistic. The Weibull density allows for hazard rates that are
+increasing, constant, or decreasing over survival time, while the
+log-logistic density also can fit rates that have a peak at a particular
+survival time.
 
-To illustrate the usefulness of the package and the broad applicability
-of the methods outlined above, we employ an example from political
-science. We replicate and extend the analysis presented in Belkin and
-Schofer (2003), which examines factors that lead to military coups.
-Belkin and Schofer's distinction between ``structural'' vs.
-``triggering''" causes of coups strongly suggests that a
-split-population approach to this question will be useful. They argue
-that many countries never experience coups because coups are effectvely
-impossible due to structural factors, while others that never experience
-coups are nevertheless at risk due to a different configuration of those
-same factors. Using language which fits nicely with the class of models
+As an example to illustrate the package functionality, we will replicate
+and extend the model of coup onsets in Belkin and Schofer (2003). Belkin
+and Schofer's paper lends itself to re-analysis with a split-populuation
+duration model because they explicitly distinguish long-term structural
+risk factors for coups from more short-term triggering causes that can
+explain the timing of a coup in a regime at risk. They argue that many
+countries never experience coups because coups are effectvely impossible
+due to structural factors, while others that never experience coups are
+nevertheless at risk due to a different configuration of those same
+factors. Using language which fits nicely with the class of models
 described above, they say ``Triggers are not the source of the original
@@
 determine the exact timing of a coup in regimes that suffer from high
-coup risk'' (p.~598 Belkin and Schofer 2003). Belkin and Schofer develop
-a measure of ``structural coup risk'' which incorporates factors such as
-the strength of democratic institutions and civil society, and a recent
-history of successful coups. They employ a logistic regression model
-which does not adequately capture the process described in the quote
-above, since it implicitly assumes that all observations are at risk for
-a coup (i.e., the probability of a coup is non-zero for all
-observations). Their structural coup risk indicator is developed
-precisely to distinguish between cases where coups are possible and
-cases where they are not, so the split-population model allows one to
-examine whether the indicator effectively separates at-risk cases from
-cases where coups are impossible.
+coup risk'' (p.~598 Belkin and Schofer 2003)
 
-We begin by loading the package and formatting the data appropriately.
+Belkin and Schofer develop a measure of ``structural coup risk'' which
+incorporates factors such as the strength of democratic institutions and
+civil society, and a recent history of successful coups. They employ a
+logistic regression model which does not adequately capture the process
+described in the quote above, since it implicitly assumes that all
+observations are at risk for a coup (i.e., the probability of a coup is
+non-zero for all observations). Their structural coup risk indicator is
+developed precisely to distinguish between cases where coups are
+possible and cases where they are not, so the split-population model
+allows one to examine whether the indicator effectively separates
+at-risk cases from cases where coups are impossible.
+
+We begin by loading the package and the Belkin and Schofer replication
+data, and formatting the data to add several variables needed by the
+split-population duration model.
 
@@
 \normalsize
-The \texttt{add}\_\texttt{duration} function takes as input a data frame
-with a binary response variable (in this example, ``coup''``) that
-measures the occurrence of the event, or failure, recorded over discrete
-time periods, and produces a data frame with a duration variable
+The \code{add\_duration} function takes as input a data frame with a
+binary response variable (in this example, ``coup''``) that measures the
+occurrence of the event, or failure, recorded over discrete time
+periods, and produces a data frame with a duration variable
 (named''duration``) for each spell as well as a binary variable
@@
 spell ended in failure. These are then used as response variables for
-the formula in the \texttt{spdur} function. Though the data used in this
-example are recorded annually, the function supports annual, monthly, or
-daily data. We begin by fitting first a Weibull and then a log-logistic
-split-population duration model using the coups data, including the
-measure of coup risk in the logit (risk) equation.
+the formula in the \texttt{spdur} function. The \code{spdur} function is
+the primary function in the package and produces a regression model
+object of class \code{spdur} which can then be used with further
+methods.
+
+Though the data used in this example are recorded annually, the function
+supports annual, monthly, or daily data. We begin by fitting first a
+Weibull and then a log-logistic split-population duration model using
+the coups data, including the measure of coup risk in the logit (risk)
+equation.
 
@@
 
-The objects produced by \texttt{spdur} can also be used with functions
-in the \texttt{xtable} package to produce \LaTeX~code for tables of
-results. The \texttt{xtable} function used on an \texttt{spdur} object
-will produce code such as that used to create Table \ref{loglog_table}.
+The objects produced by \code{spdur} can also be used with functions in
+the \code{xtable} package to produce \LaTeX~code for tables of results.
+The \code{xtable} function used on an \code{spdur} object will produce
+code such as that used to create Table \ref{loglog_table}.
 
@@
 
-The package also includes a function, \texttt{plot.hazard}, that can be
-used with an \texttt{spdur} object to plot the estimated hazard rate.
-\texttt{plot.hazard} plots the conditional hazard, which is the
+The package also includes a function, \code{plot_hazard}, that can be
+used with an \code{spdur} object to plot the estimated hazard rate.
+\code{plot_hazard} plots the conditional hazard, which is the
 probability of survival conditional on the covariates in the risk and
@@
 can choose specific covariate values by entering them as vectors in the
-arguments \texttt{xvals} and \texttt{zvals}, which correspond to the
+arguments \code{xvals} and \code{zvals}, which correspond to the
 covariates in the duration and risk equations, respectively. The command
-below creates Figure \ref{plot-hazard2}.
+below creates the graph A in Figure \ref{hazard-ex}.
 
@@
 Importantly, the package also includes functions to evaluate model
-predictions. The generic function \texttt{predict} can be used on an
-object of class \texttt{spdur} to generate several kinds of predictions,
+predictions. The generic function \code{predict} can be used on an
+object of class \code{spdur} to generate several kinds of predictions,
 including the probability that an observation is ``at-risk'' and the
 probability of failure for a given time period. In addition, the generic
-\texttt{plot} function can be used on \texttt{spdur} objects to produce
-a separation plot (Greenhill, Ward, and Sacks 2011), which is a
-graphical display for evaluating model predictions. The code below
-produces Figure \ref{insamp}.
+\code{plot} function can be used on \code{spdur} objects to produce a
+separation plot (Greenhill, Ward, and Sacks 2011), which is a graphical
+display for evaluating model predictions. The code below produces Figure
+\ref{insamp}.
 
@@
 
-The option \texttt{endSpellOnly} is set to \texttt{FALSE} so that every
+The option \code{endSpellOnly} is set to \code{FALSE} so that every
 observation, not only those at the end of a spell, is used in the plot.
-By default the \texttt{plot} function will calculate the conditional
+By default the \code{plot} function will calculate the conditional
 hazard for each observation. The separation plot sorts observations from
@@
 \normalsize
-The \texttt{spdurCrisp} function can then be used to fit a model on the
+The \code{spdurCrisp} function can then be used to fit a model on the
 training set and generate predictions for the test set.
@@
 \normalsize
-The \texttt{train},\texttt{test}, and \texttt{pred} arguments are used
-to specify data frames for the training set, test set, and forecasting
-set. We do not wish to generate forecasts here, so we simply specify the
-same data frame for the test and forecasting set. Here we use
-\texttt{spdurCrisp} to calculate the conditional risk, which is the
+The \code{train},\code{test}, and \code{pred} arguments are used to
+specify data frames for the training set, test set, and forecasting set.
+We do not wish to generate forecasts here, so we simply specify the same
+data frame for the test and forecasting set. Here we use
+\code{spdurCrisp} to calculate the conditional risk, which is the
 default calculation for this function and is the probability that an
 observation is in the at-risk population. The test set predictions are
-stored in the model object as \texttt{test.p}, and we extract them to
+stored in the model object as \code{test.p}, and we extract them to
 produce the plot shown below in Figure \ref{outsamp}, which was created
-using the \texttt{separationplot} package.
+using the \code{separationplot} package.
 
@@
 information of previous failures when constructing the necessary
-duration variables with \texttt{add\_duration()}.
+duration variables with \code{add\_duration()}.
 
@@
 function, they impact the coding of at risk vs.~immune spells. The
-\texttt{add\_duration()} function retroactively codes all observations
-in a spell as at risk if the spell itself ended in failure.
-Right-censored spells are coded as immune over their entire duration.
-This can lead to some misclassification of observations as immune even
-though they experience failure at some point in the unobserved future.
+\code{add\_duration()} function retroactively codes all observations in
+a spell as at risk if the spell itself ended in failure. Right-censored
+spells are coded as immune over their entire duration. This can lead to
+some misclassification of observations as immune even though they
+experience failure at some point in the unobserved future.
 
@@
 
-Data and outcomes in international relations, and to a lesser extent in
-other fields as well, are typically characterized by temporal dependence
-and, where a binary, ordinal, or count outcome is modelled, rare events
-or severely imbalanced classes with excessive numbers of zeros. Figure
-\ref{rates} shows a few examples for published research that models
-binary outcomes.
+Many outcomes in the area that we work in, political science and
+political violence more specifically, are rare events. Coups, the onset
+of war and civil war, mass killings, and so on are exceedingly rare
+events when one considers that most countries most of the time are
+stable. Figure \ref{rates} shows a few examples for published research
+that models binary outcomes.
 
-This is a well-recognized problem and has led to the development or use
-of several specialized mixture models like zero-inflated Poisson and
-negative binomial regression for count data, and a zero-inflated ordered
-probit for ordinal outcomes (Bagozzi et al. 2015). Split-population
-duration regression provides another principled solution to the
-challenges posed by data in this domain, but, unlike other solutions to
-the sparse outcome problem, also addresses underlying temporal dynamics.
+In the language of machine learning, we are dealing with highly
+imbalanced classes, although our data are also highly structured with
+clear dependence over space and time in countries and societies that may
+experience political violence. This is a well-recognized problem and has
+led to the development or use of several specialized mixture models like
+zero-inflated Poisson and negative binomial regression for count data,
+and a zero-inflated ordered probit for ordinal outcomes (Bagozzi et al.
+2015). Split-population duration regression provides another principled
+solution to the challenges posed by data in this domain, but, unlike
+other solutions to the sparse outcome problem, also addresses underlying
+temporal dynamics.
 
-In addition to these technical features, split-population duration
-models are also appealing in the context of theories or arguments that
-explicitly distinguish long-term structural causes from short-term
-triggering events, as Belkin and Schofer (2003) and others do, as this
-conceptual distinction can be accomodated through the seperate risk and
-duration equations.
+Split-population duration models are not only appealing in this
+technical sense, but they also match the logic or intuituion many
+analysts of politics use when they distinguish long-term risk factors
+from more fleeting triggering causes. The example we have used, Belkin
+and Schofer (2003), is a particularly clear illustration of how well the
+language of theorists maps onto the model intuition.
 

### 2015-10-29 manuscript.Rmd
more writing

@@
   \usepackage{amsmath}
+  \usepackage{booktabs}
 output: 
@@
 library("spduration")
-#data(bscoup)
-library("foreign")
-bscoup <- read.dta("data/BelkinSchoferTable4.dta")
+data(bscoup)
 bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
 bscoup      <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
-                            freq="year")
+                            freq="year", ongoing = FALSE)
 ```
 
-\normalsize
-The \code{add\_duration} function takes as input a data frame with a binary response variable (in this example, "coup"") that measures the occurrence of the event, or failure, recorded over discrete time periods, and produces a data frame with a duration variable (named "duration") for each spell as well as a binary variable ("atrisk") that is constant across the spell and indicates whether the spell ended in failure.
-These are then used as response variables for the formula in the `spdur` function.
+The \code{add\_duration} function takes as input a data frame with a binary response variable--"coup"--that measures the occurrence of the event, or failure, recorded over discrete time periods.
+We assume that the data frame is in a cross-sectional time-series format that consists of multiple observations over time of a number of subjects, in our case countries.
+
+Within the framework of duration modeling, these data conceptually consist of "spells", which are repated observations of a unit from the time they enter the data (left-censoring) or after they experienced a failure event, until the next failure event or end of observation (right censoring).
+Table \ref{tab-ex} shows how a single country could have two spells that consist of the observations for that country from 1960, when our data start, to a failure event in say 1980, when a new spell starts and continues until the next failure event or data end.
+Country "B" on the other hand experienced no coups and all observation for that country from 1960 to 2000 make up one spell.
+
+\begin{table}
+\begin{center}
+\caption{Example data frame with select duration variables}
+\label{tab-ex}
+\begin{tabular}{lrrrrr} \toprule
+Country & Year & Coup & Spell ID & Duration \\ \midrule
+A & 1960 & 0 & 1 & 0  \\ 
+A & \ldots & 0 & 1 & \ldots \\
+A & 1980 & 1 & 1 & 20 \\ \midrule
+A & 1981 & 0 & 2 & 0 \\
+A & \ldots & 0 & 2 & \ldots \\
+A & 2000 & 0 & 2 & 19  \\ \midrule
+B & 1960 & 0 & 3 & 0 \\
+B & \ldots & \ldots & 3 & \ldots \\ 
+B & 2000 & 0 & 3 & 40 \\ \bottomrule
+\end{tabular}
+\end{center}
+\end{table}
+
+The \code{add_duration} functions identifies spells in the data and returns the data frame with several additional variables needed for estimation.
+The most important among these are a counter of elapsed time in a spell, "duration".
+It also codes that a spell was at risk ("atrisk==1") if it ended in a failure event, and not at risk otherwise.
+These are then used as response variables for the formula in the \code{spdur} function.
+
+Though the data used in this example are recorded annually, the function supports annual, monthly, or daily data and will try to convert the \code{tID} input to class "Date" given the provided argument in \code{freq}.
+Another important question is how to handle consecutive 1's in the response variable provided.
+This is controlled with the \code{ongoing} argument.
+In some cases like civil war occurence, the *y* variable records all years during which a country experienced a civil war.
+Here \code{ongoing} should be set to true so that the models try to predict the onset of the civil war, but disregard ongoing conflicts (\code{failure} is set to \code{NA} for these cases, dropping them from analysis).
+With ongoing set to false, successive 1's in *y* are treated as distinct and new failures, and kept in the analysis.
+This makes sense for discrete, short-term events like coups.
+Countries can experience distinct coups in successive years.
+
 The \code{spdur} function is the primary function in the package and produces a regression model object of class \code{spdur} which can then be used with further methods.
-
-Though the data used in this example are recorded annually, the function supports annual, monthly, or daily data.
 We begin by fitting first a Weibull and then a log-logistic split-population duration model using the coups data, including the measure of coup risk in the logit (risk) equation.
 
-```{r models, include = FALSE}
+```{r models, include = TRUE, cache = TRUE}
 weib_model <- spdur(
@@
 Using the \code{summary} function on either model object will produce standard output showing the model formula, estimates for the duration and risk equations, and test statistics with $p$-values.
-One may also use the \code{AIC} and \code{BIC} function to calculate the information criterion statistics for \code{spdur} objects.
-```{r fit-stats}
-AIC(weib_model)
-AIC(loglog_model)
-BIC(weib_model)
-BIC(loglog_model)
-```
+These can be exported by calling \code{xtable} directly on the "spdur" object to produce Table \ref{loglog-table}.
 
-The objects produced by \code{spdur} can also be used with functions in the \code{xtable} package to produce \LaTeX\ code for tables of results.
-The \code{xtable} function used on an \code{spdur} object will produce code such as that used to create Table \ref{loglog_table}.
-
-```{r summary-table, include=FALSE}
+```{r summary-table, include = TRUE, eval = FALSE}
 library("xtable")
@@
 print(tbl, caption.placement="top", comment=FALSE, include.rownames=FALSE)
+```
+
+One may also use the \code{AIC} and \code{BIC} function to calculate the information criterion statistics for \code{spdur} objects.
+
+```{r fit-stats}
+matrix(c(
+  AIC(weib_model), AIC(loglog_model), BIC(weib_model), BIC(loglog_model)
+  ), ncol = 2, dimnames = list(c("Weibull", "Loglog"), c("AIC", "BIC")))
 ```
@@
   \hline
-(Dur. Intercept) & 3.30 & 0.14 & 0.00 \\ 
-  milreg & -1.53 & 0.15 & 0.00 \\ 
-  instab & -0.22 & 0.03 & 0.00 \\ 
-  regconf & 3.04 & 2.22 & 0.17 \\ 
-  log(alpha) & -0.52 & 0.07 & 0.00 \\ 
-  (Risk Intercept) & -14.61 & 9.02 & 0.11 \\ 
-  couprisk & 9.11 & 4.37 & 0.04 \\ 
-  wealth & 3.51 & 1.75 & 0.05 \\ 
-  milreg.1 & 21.76 & 0.24 & 0.00 \\ 
-  rwar & -6.37 & 5.38 & 0.24 \\ 
-  regconf.1 & -20.22 & 54.04 & 0.71 \\ 
-  samerica & -3.31 & 3.28 & 0.31 \\ 
-  camerica & -6.46 & 2.57 & 0.01 \\ 
+(Dur. Intercept) & 2.40 & 0.21 & 0.00 \\ 
+  milreg & -1.13 & 0.21 & 0.00 \\ 
+  instab & -0.09 & 0.02 & 0.00 \\ 
+  regconf & -2.52 & 2.16 & 0.24 \\ 
+  log(alpha) & -0.45 & 0.06 & 0.00 \\ 
+  (Risk Intercept) & 2.93 & 1.87 & 0.12 \\ 
+  couprisk & 0.59 & 0.32 & 0.07 \\ 
+  wealth & -0.36 & 0.28 & 0.20 \\ 
+  milreg.1 & 10.82 & 9.31 & 0.25 \\ 
+  rwar & -0.52 & 0.94 & 0.58 \\ 
+  regconf.1 & -5.43 & 5.62 & 0.33 \\ 
+  samerica & 2.09 & 1.45 & 0.15 \\ 
+  camerica & -0.39 & 0.73 & 0.59 \\ 
    \hline
@@
 
+## Out-of-sample testing
+
 Finally, we demonstrate how to evaluate a model's out of sample predictions.
-We begin by splitting the data into training and test sets, using the last year of data (1999) as the test set and omitting incomplete observations from the test set.
-```{r}
+We will use data from 1996 on as the test set, and the prior data for training.
+The \code{add_duration} function retrospectively codes the risk variable based on how a spell ended, and we therefore need to take care in how we add the duration variables for each data set.
+For the training data we need to subset the training set first so that coups in the tests set don't influence the risk coding in the training data.
+
+```{r, eval = FALSE}
 data(bscoup)
@@
 coup_train <- add_duration(coup_train, "coup", unitID="countryid", tID="year",
-                          freq="year")
+                          freq="year", ongoing = FALSE)
 ```
 
-```{r}
+```{r, include = FALSE}
+data(bscoup)
+bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
+coup_train <- bscoup[bscoup$year < 1996, ]
+coup_train <- add_duration(coup_train, "coup", unitID="countryid", tID="year",
+                          freq="year", ongoing = FALSE)
+```
+
+For the test set it is better to add the duration variables and *then* subset the test set.
+Since the test set is later in time than the training set we do not have to worry about contamination of the risk coding, but if we subset the data before building the duration variables we will start all duration counters at 1996, when in fact we can safely use the previous historic coup information.
+To do this we need to build the duration variables first:
+
+```{r, include = FALSE}
 coup_test  <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
-                           freq="year")
+                           freq="year", ongoing = FALSE)
 coup_test  <- coup_test[coup_test$year >= 1996, ]
@@
 
-\normalsize
-The \code{spdurCrisp} function can then be used to fit a model on the training set and generate predictions for the test set.
+```{r, eval = FALSE}
+coup_test  <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
+                           freq="year", ongoing = FALSE)
+coup_test  <- coup_test[coup_test$year >= 1996, ]
+```
 
-```{r}
+Now we can fit new models using the training data, and calculate predictions from these models for the test data, using \code{predict(..., newdata = coup_test)}.
+
+```{r estimate train models, cache = TRUE}
 weib_model2   <- spdur(
@@
 
-Foo
-
 ```{r}
@@
 
-\normalsize
-The \code{train},\code{test}, and \code{pred} arguments are used to specify data frames for the training set, test set, and forecasting set.
-We do not wish to generate forecasts here, so we simply specify the same data frame for the test and forecasting set.
-Here we use \code{spdurCrisp} to calculate the conditional risk, which is the default calculation for this function and is the probability that an observation is in the at-risk population.
-The test set predictions are stored in the model object as \code{test.p}, and we extract them to produce the plot shown below in Figure \ref{outsamp}, which was created using the \code{separationplot} package.
+Since we are predicting for data that is not contained in the "spdur" model objects, we have to use the \code{separationplot} functions directly from the package to produce Figure \ref{oos-sepplots}.
 
-```{r oos-sepplots}
+```{r oos-sepplots, eval = FALSE}
 library("separationplot")
-
-obs_y <- coup_test$coup[complete.cases(coup_test)]
+obs_y <- coup_test[complete.cases(coup_test), "coup"]
 
@@
 \includegraphics[width=0.8\textwidth]{graphics/oos-sepplots.pdf}
+\caption{Out-of-sample separation plots}
+\label{oos-sepplots}
 \end{center}
@@
 
-```{r}
+Lastly, here is a comparision to regular logistic regression.
+
+```{r logistic regression}
 glm_model2    <- glm(
@@
   Logistic = AIC(glm_model2))
+```
+
+```{r, eval = FALSE, include = FALSE}
+glm2_test_p   <- predict(glm_model2, newdata = coup_test, type = "response")
+glm2_test_p   <- glm2_test_p[!is.na(glm2_test_p)]
+
+obs_y <- coup_test[complete.cases(coup_test), "coup"]
+
+library("separationplot")
+separationplot(glm2_test_p,   obs_y)
+
+brier(obs_y, glm2_test_p)
+auc_roc(obs_y, glm2_test_p)
+auc_roc(obs_y, weib2_test_p)
+auc_roc(obs_y, loglog2_test_p)
+auc_pr(obs_y, glm2_test_p)
+auc_pr(obs_y, weib2_test_p)
+auc_pr(obs_y, loglog2_test_p)
 ```
@@
 Figure \ref{rates} shows a few examples for published research that models binary outcomes.
-@fearon2003ethnicity is a very widely cited study of civil war onset that uses yearly observations of all major countries, and @beger2015ensemble is an example of the positive rates when moving to monthly data for a similar set of countries.
+@fearon2003ethnicity is a very widely cited study of civil war onset that uses yearly observations of all major countries, and @beger2014ensemble is an example of the positive rates when moving to monthly data for a similar set of countries.
+The positive rates range from 5 to less than 0.2\% of all data points.
 
-In the language of machine learning, we are dealing with highly imbalanced classes, although our data are also highly structured with clear dependence over space and time in countries and societies that may experience political violence.
-This is a well-recognized problem and has led to the development or use of several specialized mixture models like zero-inflated Poisson and negative binomial regression for count data, and a zero-inflated ordered probit for ordinal outcomes [@bagozzi2015modeling].
-Split-population duration regression provides another principled solution to the challenges posed by data in this domain, but, unlike other solutions to the sparse outcome problem, also addresses underlying temporal dynamics.
+In the language of machine learning, we are dealing with highly imbalanced classes, this is a well-recognized problem and has led to the development or use of several specialized mixture models like zero-inflated Poisson and negative binomial regression for count data, and a zero-inflated ordered probit for ordinal outcomes [@bagozzi2015modeling].
+Split-population duration regression provides another principled solution to the challenges posed by data in this domain, but, unlike other solutions to the sparse outcome problem, also addresses underlying temporal dynamics that are an important part of the non-independent data political scientists commonly use.
 
-Split-population duration models are not only appealing in this technical sense, but they also match the logic or intuituion many analysts of politics use when they distinguish long-term risk factors from more fleeting triggering causes.
+Split-population duration models are not only appealing in a technical sense, but they also match the logic or intuituion many analysts of politics use when they distinguish long-term risk factors from more fleeting triggering causes.
 The example we have used, @belkin2003toward, is a particularly clear illustration of how well the language of theorists maps onto the model intuition.

### 2015-10-29 replication.R

@@ -1,32 +1,32 @@
 #  Replication script for JSS manuscript examples
 
 library("spduration")
-library(foreign)
-library(magrittr)
-library(ggplot2)
-library(xtable)
-library(scales)
+library("foreign")
+library("magrittr")
+library("ggplot2")
+library("xtable")
+library("scales")
 
 
 data(bscoup)
 bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
-bscoup <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
-                       freq="year")
+bscoup      <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
+                            freq="year", ongoing = FALSE)
 
 weib_model <- spdur(
   duration ~ milreg + instab + regconf,
   atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
-  data=bscoup)
+  data=bscoup, silent = TRUE)
 
 loglog_model <- spdur(
   duration ~ milreg + instab + regconf,
   atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
-  data=bscoup, distr="loglog")
+  data=bscoup, distr="loglog", silent = TRUE)
 
-AIC(weib_model)
-AIC(loglog_model)
-BIC(weib_model)
-BIC(loglog_model)
+matrix(c(
+  AIC(weib_model), AIC(loglog_model), BIC(weib_model), BIC(loglog_model)
+  ), ncol = 2, dimnames = list(c("Weibull", "Loglog"), c("AIC", "BIC")))
+
 
 library("xtable")
 tbl <- xtable(loglog_model, caption="Coup model with log-logistic hazard",
@@ -67,10 +67,10 @@ data(bscoup)
 bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
 coup_train <- bscoup[bscoup$year < 1996, ]
 coup_train <- add_duration(coup_train, "coup", unitID="countryid", tID="year",
-                           freq="year")
+                           freq="year", ongoing = FALSE)
 
 coup_test  <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
-                           freq="year")
+                           freq="year", ongoing = FALSE)
 coup_test  <- coup_test[coup_test$year >= 1996, ]
 
 weib_model2   <- spdur(
[…]

### 2015-11-03 manuscript.Rmd
spell checking & bibs

@@
 The statistical model accounts for units that are immune to a certain outcome and not part of the duration process the researcher is primarily interested in.
-We provide insights that if immune units exist, we can significantly increase the predictive performance compared to standard duration model.
-The package includes estimation and forecasting methods for split-population Weibull and Loglogistic models.
+We provide insights that if immune units exist, we can significantly increase the predictive performance compared to standard duration models.
+The package includes estimation and several post-estimation methods for split-population Weibull and Loglogistic models.
 keywords:
   # at least one keyword must be supplied
-  formatted: [keywords, not capitalized, "\\proglang{R}"]
-  plain:     [keywords, not capitalized, R]
+  formatted: [regression, rare events, "\\proglang{R}"]
+  plain:     [regression, rare events, R]
 preamble: >
@@
 ---
-
-# Introduction
-
-This template demonstrates some of the basic latex you'll need to know to create a JSS article.
-
-## Code formatting
-
-Don't use markdown, instead use the more precise latex commands:
-
-* \proglang{Java}
-* \pkg{plyr}
-* \code{print("abc")}
 
@@
 
-This insight was realized as early as 1949 by @boag1949maximum and @berkson1952survival, who were researching survival rates in cancer patients following treatment, and where quite obviously some fraction of patients survived because their cancer was cured, while others relapsed after apparent remission due to levels of disease below detectable thresholds.
+This insight was realized as early as 1949 by @boag1949maximum and @berkson1952survival, who were researching survival rates in cancer patients following treatment, and were quite obviously some fraction of patients survived because their cancer was cured, while others relapsed after apparent remission due to levels of disease below detectable thresholds.
 As with conventional duration models, their origin in health and medicine has shaped the terminology conventionally used with such model, e.g.
@@
 Yet the intuition underlying split-population duration models has led to applications in a broad range of subject areas outside demographics and medicine.
-In an early and foundational application that reached beyond criminal science, @schmidt1989predicting examined criminal recidivism using data on close to 10,000 prisoners from the North Carolina prison system in the late 1970's and early 1980's to identify factors that influence whether a criminal lapses at all, and if so which factors are related to the amount of time between prison stints.
+In an early and foundational application that reached beyond medicine, @schmidt1989predicting examined criminal recidivism using data on close to 10,000 prisoners from the North Carolina prison system in the late 1970's and early 1980's to identify factors that influence whether a criminal lapses at all, and if so which factors are related to the amount of time between prison stints.
 This work already includes a full formulation of the model with independent covariates for both the duration equation and the risk or cure equation, although only with subject-specific covariates rather than time-varying covariates with multiple data points per subject.
 
-In public health, @douglas1994hazard use data from the US to model factors related to the age at which smokers started the habit, and @forster2001role examines the impact of tobacco taxes on smoking and quitting decisions.
-@deyoung2003failure models the failure of new commercial banks in the US during the 1980's.
-Lastly, in our own domain, political science, @svolik2008authoritarian used a split-population duration framework to study democratic consolidation, i.e.
-whether democratic regimes persist or slide back to authoritarianism, and when.
+In a complementary set of research from public health, @douglas1994hazard use data from the US to model the age at which smokers started the habit, and @forster2001role examine the impact of tobacco taxes on smoking and quitting decisions.
+@deyoung2003failure, an economist, models the failure of new commercial banks in the US during the 1980's.
+Lastly, in our own domain, political science, @svolik2008authoritarian used a split-population duration framework to whether democratic regimes persist or slide back to authoritarianism, and when.
 Building on this effort, split-population duration models have also been used, along with other models, to produce regular predictions for five different forms of political conflict for the Integrated Crisis and Early Warning System (ICEWS) project [@ward2013learning] and to model irregular leadership changes for the Political Instability Task Force [PITF; @beger2014ensemble].
@@
 The last component to complete the likelihood is the choice of a distribution for the shape of the hazard rate.
-The \pkg{spduration} package implements two hazard rate shapes, Weibull\footnote{In the code, $f(t) = \alpha \lambda^{\alpha} t^{\alpha - 1} e^{-(\lambda t)^\alpha}$.} and log-logistic:
+The \pkg{spduration} package implements two hazard rate shapes, Weibull and log-logistic:
 
@@
 Using the \code{summary} function on either model object will produce standard output showing the model formula, estimates for the duration and risk equations, and test statistics with $p$-values.
-These can be exported by calling \code{xtable} directly on the "spdur" object to produce Table \ref{loglog-table}.
+These can be exported by calling \code{xtable} directly on the "spdur" object to produce Table 2.
 
@@
 print(tbl, caption.placement="top", comment=FALSE, include.rownames=FALSE)
-```
-
-One may also use the \code{AIC} and \code{BIC} function to calculate the information criterion statistics for \code{spdur} objects.
-
-```{r fit-stats}
-matrix(c(
-  AIC(weib_model), AIC(loglog_model), BIC(weib_model), BIC(loglog_model)
-  ), ncol = 2, dimnames = list(c("Weibull", "Loglog"), c("AIC", "BIC")))
 ```
@@
 The table shows estimates from the duration equation, beginning with the intercept, and then estimates from the risk equation.
-Interestingly, the coup risk variable has a rather large and statistically significant coefficient in the risk equation.
+The duration component of the model is in accelerated failure time form and the effect of coefficient estimates is on the log of expected time to failure.
+The negative coefficient for military regimes for example shows that the expected time to a coup is shorter in military regimes than non-military regimes, holding all other factors constant.
+In the risk equation positive coefficients mean higher risk of coup, and military regimes thus also have a higher risk of experiencing any coup at all.
+
+One may also use the \code{AIC} and \code{BIC} function to calculate the information criterion statistics for \code{spdur} objects.
+In our example both models are close enough in both statistics to be indistinguishable, so we will continue to focus on the log-logistic form.
+
+```{r fit-stats}
+matrix(c(
+  AIC(weib_model), AIC(loglog_model), BIC(weib_model), BIC(loglog_model)
+  ), ncol = 2, dimnames = list(c("Weibull", "Loglog"), c("AIC", "BIC")))
+```
 
@@
 ```{r hazard-ex-1, eval = FALSE}
-plot(loglog_model, type="hazard", main = "A")
+plot_hazard(loglog_model, main = "A")
 ```
@@
 
+As mean values are maybe not that interesting, we can also supply our values for a scenario of choice.
+The values are used in the same order that variables are specified in the equations used to estimate the models, which also corresponds to the order of variables in Table 2.
+Including an intercept term.
+This produces plot B in Figure \ref{hazard-ex}
+
 ```{r hazard-ex-2, eval = FALSE}
-plot(loglog_model, type="hazard", 
+plot_hazard(loglog_model, 
      xvals = c(1, 1, 10, 0.05), 
@@
 
-Importantly, the package also includes functions to evaluate model predictions.
-The generic function \code{predict} can be used on an object of class \code{spdur} to generate several kinds of predictions, including the probability that an observation is ``at-risk'' and the probability of failure for a given time period.
-In addition, the generic \code{plot} function can be used on \code{spdur} objects to produce a separation plot [@greenhill2011separation], which is a graphical display for evaluating model predictions.
+The hazard plots can also be called with \code{plot(..., type = "hazard")}.
+Without a type argument, \code{plot.spdur} will produce a separation plot, which is a graphical display for evaluating model predictions [@greenhill2011separation].
 The code below produces Figure \ref{insamp}.
@@
 This makes it easy to see whether the model is assigning high probabilities of failure to actual cases of failure, and low probabilities to non-failures.
+
+Underlying both plotting functions is the \code{predict} function, which can be used on an object of class \code{spdur} to generate several kinds of predictions, including the probability that an observation is ``at-risk'' and the probability of failure for a given time period.
 
@@
 
-Lastly, here is a comparision to regular logistic regression.
-
-```{r logistic regression}
-glm_model2    <- glm(
-  coup ~ milreg + instab + regconf + couprisk + wealth + rwar + regconf + 
-    samerica + camerica,
-  data = coup_train, family = "binomial"
-)
-
-c(Weibull = AIC(weib_model2), 
-  Loglog = AIC(loglog_model2), 
-  Logistic = AIC(glm_model2))
-```
-
-```{r, eval = FALSE, include = FALSE}
-glm2_test_p   <- predict(glm_model2, newdata = coup_test, type = "response")
-glm2_test_p   <- glm2_test_p[!is.na(glm2_test_p)]
-
-obs_y <- coup_test[complete.cases(coup_test), "coup"]
-
-library("separationplot")
-separationplot(glm2_test_p,   obs_y)
-
-brier(obs_y, glm2_test_p)
-auc_roc(obs_y, glm2_test_p)
-auc_roc(obs_y, weib2_test_p)
-auc_roc(obs_y, loglog2_test_p)
-auc_pr(obs_y, glm2_test_p)
-auc_pr(obs_y, weib2_test_p)
-auc_pr(obs_y, loglog2_test_p)
-```
-
 \section{Censoring considerations}

### 2015-11-03 replication.R

@@ -38,8 +38,8 @@ print(tbl, caption.placement="top", comment=FALSE, include.rownames=FALSE)
 
 pdf("graphics/hazard-ex.pdf", height=4, width=8)
 par(mfrow = c(1, 2))
-plot(loglog_model, type="hazard", main = "A")
-plot(loglog_model, type="hazard", 
+plot_hazard(loglog_model, main = "A")
+plot_hazard(loglog_model, 
      xvals = c(1, 1, 10, 0.05), 
      zvals = c(1, 7, 8.64, 1, 1, 0.05, 0, 0),
      main = "B")

### 2015-11-04 manuscript.Rmd
fixes for Mike comments

@@
 author:
+  - name: Andreas Beger
+    affiliation: Duke University
+    address: >
+      Department of Political Science, Duke University
+      Durham, NC, USA
+    email: ab428@duke.edu
+    url: something
   - name: Daniel W. Hill
@@
       First line
-      Second line
+      Athens, GA, USA
     email: dwhill@uga.edu
-    url: http://spia.uga.edu/directory/faculty_staff/hill-daniel
   - name: Nils W. Metternich
     affiliation: University College London
-  - name: Andreas Beger
-    affiliation: Duke University
+    address: >
+      First line
+      London, UK
+    email: n.metternich@ucl.ac.uk
   - name: Michael D. Ward
     affiliation: Duke University
+    address: >
+      Department of Political Science, Duke University
+      Durham, NC, USA
+    email: mw160@duke.edu
 title:
@@
 Newborns are at a great risk of dying, but as they grow older this risk declines and then gradually starts to increase again after the age of 9-10.
-This risk can be increased or decreased by behavioral (smoking, exercise, diet, etc) and structural factors (health care, regulations, urban vs rural, etc), but there is an underlying risk that is time dependent and impacts on all units.
+This risk can be increased or decreased by behavioral (smoking, exercise, diet) and structural factors (health care, regulations, urban vs rural), but there is an underlying risk that is time dependent and impacts on all units.
 
@@
 Basic formulations of such models, like the parametric exponential or Weibull regressions or semi-paramaetric Cox regression, implicitly assume that all subjects or units under observation, including right-censored observations, will eventually experience the event of interest.
-This assumption is outright wrong or at least disadvantageous in many substantive areas and empirical applications, where there instead is a sub-population of units or subjects that will never experience an event, and thus are effectively ``cured''.
+This assumption is wrong or at least disadvantageous in many substantive areas and empirical applications, where there instead is a sub-population of units or subjects that will never experience an event, and thus are effectively ``cured''.
 
-This insight was realized as early as 1949 by @boag1949maximum and @berkson1952survival, who were researching survival rates in cancer patients following treatment, and were quite obviously some fraction of patients survived because their cancer was cured, while others relapsed after apparent remission due to levels of disease below detectable thresholds.
+This insight was realized as early as 1949 by @boag1949maximum and @berkson1952survival, who were researching survival rates in cancer patients, and where some fraction of patients survived because their cancer was cured, while others relapsed after apparent remission due to levels of disease below detectable thresholds.
 As with conventional duration models, their origin in health and medicine has shaped the terminology conventionally used with such model, e.g.
@@
 @deyoung2003failure, an economist, models the failure of new commercial banks in the US during the 1980's.
-Lastly, in our own domain, political science, @svolik2008authoritarian used a split-population duration framework to whether democratic regimes persist or slide back to authoritarianism, and when.
+@svolik2008authoritarian used a split-population duration framework to whether democratic regimes persist or slide back to authoritarianism, and when.
 Building on this effort, split-population duration models have also been used, along with other models, to produce regular predictions for five different forms of political conflict for the Integrated Crisis and Early Warning System (ICEWS) project [@ward2013learning] and to model irregular leadership changes for the Political Instability Task Force [PITF; @beger2014ensemble].
@@
 Conventional duration models assume that all subjects will eventually fail, if we can only observe them long enough.
-The likelihood for a data point with survival time $t$ is thus the failure rate at that time or the probability of survival beyond $t$, depending on whether the observation is right-censored ($1-\delta_i$) or not:
+The likelihood for a data point with survival time $t$ is thus the failure rate $f(t)$ at that time or the probability of survival beyond $t$, $S(t)$, depending on whether the subject has already experienced the event ($\delta_i$) or is right-censored ($1-\delta_i$):
 \begin{eqnarray}
@@
 As an example to illustrate the package functionality, we will replicate and extend the model of coup onsets in @belkin2003toward.
+
 Belkin and Schofer's paper lends itself to re-analysis with a split-populuation duration model because they explicitly distinguish long-term structural risk factors for coups from more short-term triggering causes that can explain the timing of a coup in a regime at risk.
 They argue that many countries never experience coups because coups are effectvely impossible due to structural factors, while others that never experience coups are nevertheless at risk due to a different configuration of those same factors.
-Using language which fits nicely with the class of models described above, they say "Triggers are not the source of the original risk, and in the absence of structural causes, the presence of triggering factors alone cannot lead to a coup.
+Using language which fits nicely with the class of models described above, they say "[t]riggers are not the source of the original risk, and in the absence of structural causes, the presence of triggering factors alone cannot lead to a coup.
 Hence, triggers should not be equated with coup risk.
-Rather, they are factors that may determine the exact timing of a coup in regimes that suffer from high coup risk" [p.
-598 @belkin2003toward]
+Rather, they are factors that may determine the exact timing of a coup in regimes that suffer from high coup risk" [@belkin2003toward, p.
+598].
 
-Belkin and Schofer develop a measure of ``structural coup risk'' which incorporates factors such as the strength of democratic institutions and civil society, and a recent history of successful coups.
-They employ a logistic regression model which does not adequately capture the process described in the quote above, since it implicitly assumes that all observations are at risk for a coup (i.e., the probability of a coup is non-zero for all observations).
+For their empirical test Belkin and Schofer develop a measure of ``structural coup risk'' which incorporates factors such as the strength of democratic institutions and civil society, and a recent history of successful coups, but the model itself is implemented using a logistic regression model.
+This does not adequately capture the process described in the quote above, since it implicitly assumes that all observations are at risk for a coup (i.e., the probability of a coup is non-zero for all observations).
 Their structural coup risk indicator is developed precisely to distinguish between cases where coups are possible and cases where they are not, so the split-population model allows one to examine whether the indicator effectively separates at-risk cases from cases where coups are impossible.
@@
 
-```{r data}
+```{r data, warning=FALSE}
 library("spduration")
@@
 Country & Year & Coup & Spell ID & Duration \\ \midrule
-A & 1960 & 0 & 1 & 0  \\ 
-A & \ldots & 0 & 1 & \ldots \\
-A & 1980 & 1 & 1 & 20 \\ \midrule
-A & 1981 & 0 & 2 & 0 \\
-A & \ldots & 0 & 2 & \ldots \\
-A & 2000 & 0 & 2 & 19  \\ \midrule
-B & 1960 & 0 & 3 & 0 \\
-B & \ldots & \ldots & 3 & \ldots \\ 
-B & 2000 & 0 & 3 & 40 \\ \bottomrule
+Portugal & 1960 & 0 & 1 & 1  \\ 
+\ldots & \ldots & 0 & 1 & \ldots \\
+\ldots & 1975 & 1 & 1 & 16 \\ \midrule
+Portugal & 1976 & 0 & 2 & 1 \\
+\ldots & \ldots & 0 & 2 & \ldots \\
+\ldots & 2000 & 0 & 2 & 20  \\ \midrule
+Canada & 1960 & 0 & 3 & 1 \\
+\ldots & \ldots & 0 & 3 & \ldots \\ 
+\ldots & 2000 & 0 & 3 & 41 \\ \bottomrule
 \end{tabular}
@@
 
-The table shows estimates from the duration equation, beginning with the intercept, and then estimates from the risk equation.
+Table \ref{loglog_table} shows estimates from the duration equation, beginning with the intercept, and then estimates from the risk equation.
 The duration component of the model is in accelerated failure time form and the effect of coefficient estimates is on the log of expected time to failure.
@@
 
-Many outcomes in the area that we work in, political science and political violence more specifically, are rare events.
+Many outcomes are rare events.
 Coups, the onset of war and civil war, mass killings, and so on are exceedingly rare events when one considers that most countries most of the time are stable.
 Figure \ref{rates} shows a few examples for published research that models binary outcomes.
-@fearon2003ethnicity is a very widely cited study of civil war onset that uses yearly observations of all major countries, and @beger2014ensemble is an example of the positive rates when moving to monthly data for a similar set of countries.
+@fearon2003ethnicity is a widely cited study of civil war onset that uses yearly observations of all major countries; @beger2014ensemble is an example of the positive rates when moving to monthly data for a similar set of countries.
 The positive rates range from 5 to less than 0.2\% of all data points.
 
-In the language of machine learning, we are dealing with highly imbalanced classes, this is a well-recognized problem and has led to the development or use of several specialized mixture models like zero-inflated Poisson and negative binomial regression for count data, and a zero-inflated ordered probit for ordinal outcomes [@bagozzi2015modeling].
+In the language of machine learning, we are dealing with highly imbalanced classes.
+It is a well-recognized problem and has led to the development or use of several specialized mixture models like zero-inflated Poisson and negative binomial regression for count data, and a zero-inflated ordered probit for ordinal outcomes [@bagozzi2015modeling].
 Split-population duration regression provides another principled solution to the challenges posed by data in this domain, but, unlike other solutions to the sparse outcome problem, also addresses underlying temporal dynamics that are an important part of the non-independent data political scientists commonly use.

### 2015-12-07 manuscript.Rmd
Danny's changes

@@
     url: something
-  - name: Daniel W. Hill
+  - name: Daniel W. Hill, Jr.
     affiliation: University of Georgia
@@
 Newborns are at a great risk of dying, but as they grow older this risk declines and then gradually starts to increase again after the age of 9-10.
-This risk can be increased or decreased by behavioral (smoking, exercise, diet) and structural factors (health care, regulations, urban vs rural), but there is an underlying risk that is time dependent and impacts on all units.
+Individual behaviors (smoking, exercise, diet) and structural factors (health care, regulations, urban vs rural) can increase or decrease this risk, but there is an underlying risk that is time dependent and impacts on all units.
 
-However, sometimes we are interested in situations in which not all units are at risk of dying, failing, or acquiring some new characteristic.
+However, sometimes we are interested in situations in which not all units are at risk of dying or experiencing some other event of interest.
 Imagine we are interested in the risk of acquiring the flu.
-First, we might think that everyone is at risk of getting the flu and it just depends on individual behavior (e.g.
+We might think that everyone is at risk of getting the flu, and individual behavior (e.g.
 good hygiene) and structural factors (e.g.
-workplace) whether individuals will acquire the flu.
-If this would be the case we could use a standard duration model and estimate how different factors impact on the baseline risk.
-But there might also be individuals who are immune to the flu, because they received a flu shot, had the same flu virus in the past year, or some other factors that makes it impossible for them to get sick.
+workplace) determine whether individuals catch the flu.
+If this were the case we could use a standard duration model and estimate how different factors impact on the baseline risk.
+But there might also be individuals who are immune to the flu, because they received a flu shot, had the same flu virus in the past year, or have some other characteristic that makes it impossible for them to get sick.
 Given that this is the case we have two underlying populations: An at risk population and an immune one.
@@
 Duration models, where the variable to be modeled describes a distribution of times until some event of interest, were originally developed in health and demographic research, where they grew naturally from life tables and survival records for medical patients.
-Basic formulations of such models, like the parametric exponential or Weibull regressions or semi-paramaetric Cox regression, implicitly assume that all subjects or units under observation, including right-censored observations, will eventually experience the event of interest.
+Basic formulations of such models, like the parametric exponential or Weibull regressions or semi-parametric Cox regression, implicitly assume that all subjects or units under observation, including right-censored observations, will eventually experience the event of interest.
 This assumption is wrong or at least disadvantageous in many substantive areas and empirical applications, where there instead is a sub-population of units or subjects that will never experience an event, and thus are effectively ``cured''.
@@
 Yet the intuition underlying split-population duration models has led to applications in a broad range of subject areas outside demographics and medicine.
-In an early and foundational application that reached beyond medicine, @schmidt1989predicting examined criminal recidivism using data on close to 10,000 prisoners from the North Carolina prison system in the late 1970's and early 1980's to identify factors that influence whether a criminal lapses at all, and if so which factors are related to the amount of time between prison stints.
+In an early and foundational application that reached beyond medicine, @schmidt1989predicting examined criminal recidivism using data on close to 10,000 prisoners from the North Carolina prison system in the late 1970's and early 1980's to identify factors that influence whether a criminal relapses at all, and if so which factors are related to the amount of time between prison stints.
 This work already includes a full formulation of the model with independent covariates for both the duration equation and the risk or cure equation, although only with subject-specific covariates rather than time-varying covariates with multiple data points per subject.
@@
 The cumulative failure rate ($F(t) = 1 - S(t)$) over time converges to 1, meaning all subjects fail eventually.
-This assumption is untenable in many situations.
-Some cancer patients are cured after treatment while others relapse, most young people do not start smoking even though most smokers started at a young age, and many states will not experience the kind of violence that persists in some parts of the world.
-A better assumption in such contexts is that there is a subpopulation of subjects who are at risk of experiencing an event at some point, and another subpopulation who will never experience the event.
-
+In the examples of applied research discussed above this assumption is untenable.
+Some cancer patients are cured after treatment, most young people never become regular smokers, and many contemporary states will not experience the kind of violence that persists in some parts of the world.
 The presence of a large sub-population which is not at risk for an event in practice will inflate estimates of the survival fraction, and reduce hazard estimates for all subjects by essentially, and falsely, distributing risk among subjects that genuinely will fail and those that are cured.
@@
 Crucially, this split-population framework is primarily useful in contexts where sub-populations are not clearly or easily identifiable.
-Even though there is a clear sub-populations in a model of the age at first pregnancy for humans---men---no researcher in their right mind would think to include male subjects in their data at all, for example.
-On the other hand, whether a cancer patient is cured or not cured given that they have no visible signs of cancer following treatment or have hit the 5-year disease free survival mark is much less clear and in fact the question that initially led to the idea for split-population duration modeling [@boag1949maximum, @berkson1952survival].
+For example, there is a clear sub-population in a model of the age at first pregnancy for humans---men---which researchers can easily identify and exclude from their data.
+Much less clear are answers to questions such as whether a cancer patient is cured or not cured given that they have no visible signs of cancer following treatment or have hit the 5-year disease free survival mark.
+In such situations split-population models provide a way to infer membership in sub-populations in a probabilistic fashion.
 
@@
 Where $\lambda = e^{-x_i\beta}$ is a parameter of covariates.
+The Weibull density allows for hazard rates that are increasing, constant, or decreasing over survival time, while the log-logistic density also can fit rates that have a peak at a particular survival time.
 
@@
 For a given unconditional risk rate $\pi$, the probability that a particular case with survival time $t$ is in the risk set decreases over time because an increasing number of surviving cases consist of immune or cured ($1-\pi$) cases that will never fail.
-In the hazard rate, the failure rate in the numerator is conditional on the probability that a case is in the risk set, give survival up to time $t$, and the numerator is an adjusted survivor function that accounts for the fraction of cured cases by time $t$, which is $1-\pi(t)$.
+In the hazard rate, the failure rate in the numerator is conditional on the probability that a case is in the risk set, given survival up to time $t$, and the numerator is an adjusted survivor function that accounts for the fraction of cured cases by time $t$, which is $1-\pi(t)$.
 
@@
 
-The \pkg{spduration} package for \proglang{R} implements a split-population duration regression for data with time-varying covariates and with two options for the hazard rate shape, Weibull and log-logistic.
-The Weibull density allows for hazard rates that are increasing, constant, or decreasing over survival time, while the log-logistic density also can fit rates that have a peak at a particular survival time.
-
-As an example to illustrate the package functionality, we will replicate and extend the model of coup onsets in @belkin2003toward.
-
-Belkin and Schofer's paper lends itself to re-analysis with a split-populuation duration model because they explicitly distinguish long-term structural risk factors for coups from more short-term triggering causes that can explain the timing of a coup in a regime at risk.
+As an example to illustrate the package functionality, we will examine the model of coups d`etat in @belkin2003toward.
+Belkin and Schofer's paper lends itself to re-analysis with a split-populuation duration model because they explicitly distinguish long-term structural risk factors for coups from more short-term triggering causes that can explain the timing of a coup in an at-risk regime.
 They argue that many countries never experience coups because coups are effectvely impossible due to structural factors, while others that never experience coups are nevertheless at risk due to a different configuration of those same factors.
@@
 
-Within the framework of duration modeling, these data conceptually consist of "spells", which are repated observations of a unit from the time they enter the data (left-censoring) or after they experienced a failure event, until the next failure event or end of observation (right censoring).
-Table \ref{tab-ex} shows how a single country could have two spells that consist of the observations for that country from 1960, when our data start, to a failure event in say 1980, when a new spell starts and continues until the next failure event or data end.
-Country "B" on the other hand experienced no coups and all observation for that country from 1960 to 2000 make up one spell.
+Within the framework of duration modeling, these data conceptually consist of "spells", which are repated observations of a unit from the time they enter the data (left-censoring) or after they experienced the event of interest, until the next event or the end of observation (right censoring).
+Table \ref{tab-ex} shows how a single country could have two spells during the observation period.
+Observations for Portugal make up two spells, one that begins in 1960, when our data start, to a coup event in 1975, when a second spell starts and continues until either the next coup or the data end.
+Canada on the other hand experienced no coups and so observations for that country from 1960 to 2000 make up one spell.
 
@@
 The duration component of the model is in accelerated failure time form and the effect of coefficient estimates is on the log of expected time to failure.
-The negative coefficient for military regimes for example shows that the expected time to a coup is shorter in military regimes than non-military regimes, holding all other factors constant.
+The negative coefficient for military regimes for example means that the expected time to a coup is shorter in military regimes than non-military regimes, holding all other factors constant.
 In the risk equation positive coefficients mean higher risk of coup, and military regimes thus also have a higher risk of experiencing any coup at all.
@@
 
-Many outcomes are rare events.
+Many outcomes of interest are rare events.
 Coups, the onset of war and civil war, mass killings, and so on are exceedingly rare events when one considers that most countries most of the time are stable.

### 2015-12-11 manuscript.Rmd
Intro update

@@
 
-Duration models are an important class of statistical estimators that take into account the duration dependency of outcomes we are interested in.
-For example, the risk of dying is not independent of time.
-Newborns are at a great risk of dying, but as they grow older this risk declines and then gradually starts to increase again after the age of 9-10.
-Individual behaviors (smoking, exercise, diet) and structural factors (health care, regulations, urban vs rural) can increase or decrease this risk, but there is an underlying risk that is time dependent and impacts on all units.
+Duration models are an important class of statistical estimators that take into account the duration dependency of specific outcomes.
+A prominent example is that the risk of dying depends on the age of an individual.
+Newborns are at a greater risk of dying, but as they grow older this risk quickly declines and then gradually starts to increase again after the age of 9-10.
+While individual behavior (smoking, exercising, diet) and structural factors (health care, health regulations, urban vs rural) can increase or decrease the probability of dying over time, the underlying risk is time dependent and impacts on all individuals.
 
-However, sometimes we are interested in situations in which not all units are at risk of dying or experiencing some other event of interest.
-Imagine we are interested in the risk of acquiring the flu.
-We might think that everyone is at risk of getting the flu, and individual behavior (e.g.
+However, there are conditions under which not all individuals have the same underlying risk to experience a specific outcome and might not even be at risk at all.
+Consider the risk of acquiring a viral infection like the common flu.
+Let us initially assume that everyone is at risk of infection, and individual behavior (e.g.
 good hygiene) and structural factors (e.g.
 workplace) determine whether individuals catch the flu.
-If this were the case we could use a standard duration model and estimate how different factors impact on the baseline risk.
-But there might also be individuals who are immune to the flu, because they received a flu shot, had the same flu virus in the past year, or have some other characteristic that makes it impossible for them to get sick.
-Given that this is the case we have two underlying populations: An at risk population and an immune one.
-If the immune population is relatively large, estimates using a standard duration model will be biased and predictions resulting from such a model inaccurate.
+Under these assumptions a standard duration model and should provide us with efficient estimates of how different behavioral and structural factors impact on the general baseline risk.
+However, in the overall population there might be individuals that are immune to the flu, because they received vaccination, had the specific flu virus in the past, or have some other characteristic that makes it impossible for them to attract the decease.
+In this instance we have two underlying populations: An at risk population and an immune one.
+If the immune population is relatively large, estimates using a standard duration model will be biased and predictions inaccurate.
 

### 2015-12-11 manuscript.Rmd
update second section

@@
 
-Duration models, where the variable to be modeled describes a distribution of times until some event of interest, were originally developed in health and demographic research, where they grew naturally from life tables and survival records for medical patients.
+Duration models, where baseline risk is modeled by some distribution of time, were originally developed in health and demographic research and grew naturally from life tables and survival records for medical patients.
 Basic formulations of such models, like the parametric exponential or Weibull regressions or semi-parametric Cox regression, implicitly assume that all subjects or units under observation, including right-censored observations, will eventually experience the event of interest.
-This assumption is wrong or at least disadvantageous in many substantive areas and empirical applications, where there instead is a sub-population of units or subjects that will never experience an event, and thus are effectively ``cured''.
+This assumption may be violated in many substantive areas and empirical applications, where a sub-population of units or individuals will never experience an event, and thus are effectively ``cured''.
 
-This insight was realized as early as 1949 by @boag1949maximum and @berkson1952survival, who were researching survival rates in cancer patients, and where some fraction of patients survived because their cancer was cured, while others relapsed after apparent remission due to levels of disease below detectable thresholds.
-As with conventional duration models, their origin in health and medicine has shaped the terminology conventionally used with such model, e.g.
-split-population duration models are sometimes also referred to in such contexts as cure rate models, and the basic concepts like survival and failure rates reference the survival of humans.
+The insight that populations might be split in regard to their baseline risk has been formulated as early as 1949 by @boag1949maximum and @berkson1952survival, who were researching survival rates in cancer patients.
+In their application some fraction of patients survived because their cancer was cured, while others relapsed after apparent remission due to levels of disease below detectable thresholds.
+The development of duration methods in health and medicine has also shaped the terminology conventionally used for such model.
+For example, split-population duration models are also referred to as cure rate models.
+Similarly, the basic concepts like survival and failure rates reference to the survival of humans.
 
@@
 In an early and foundational application that reached beyond medicine, @schmidt1989predicting examined criminal recidivism using data on close to 10,000 prisoners from the North Carolina prison system in the late 1970's and early 1980's to identify factors that influence whether a criminal relapses at all, and if so which factors are related to the amount of time between prison stints.
-This work already includes a full formulation of the model with independent covariates for both the duration equation and the risk or cure equation, although only with subject-specific covariates rather than time-varying covariates with multiple data points per subject.
+This work already includes a full formulation of a model with independent covariates for both the duration equation and the risk or cure equation, although only with subject-specific covariates rather than time-varying covariates for multiple data points per individual.
 
-In a complementary set of research from public health, @douglas1994hazard use data from the US to model the age at which smokers started the habit, and @forster2001role examine the impact of tobacco taxes on smoking and quitting decisions.
+In a complementary set of research from public health, @douglas1994hazard use data from the United States to model the age at which individuals began to smoke, and @forster2001role examine the impact of tobacco taxes on smoking and quitting decisions.
 @deyoung2003failure, an economist, models the failure of new commercial banks in the US during the 1980's.
-@svolik2008authoritarian used a split-population duration framework to whether democratic regimes persist or slide back to authoritarianism, and when.
+@svolik2008authoritarian uses a split-population duration framework to examine whether democratic regimes persist or revert to authoritarianism, and when.
 Building on this effort, split-population duration models have also been used, along with other models, to produce regular predictions for five different forms of political conflict for the Integrated Crisis and Early Warning System (ICEWS) project [@ward2013learning] and to model irregular leadership changes for the Political Instability Task Force [PITF; @beger2014ensemble].

### 2015-12-11 manuscript.Rmd
update section three

@@
 
-Conventional duration models assume that all subjects will eventually fail, if we can only observe them long enough.
+Conventional duration models assume that all subjects will eventually fail, die, or experience a specific outcome.
 The likelihood for a data point with survival time $t$ is thus the failure rate $f(t)$ at that time or the probability of survival beyond $t$, $S(t)$, depending on whether the subject has already experienced the event ($\delta_i$) or is right-censored ($1-\delta_i$):
@@
 
-The major modeling question in this setting, which we will return to below, is the choice of a function to describe the evolution of the hazard rate $h(t) = \frac{f(t)}{S(t)}$ over time, e.g.
-with exponential, Weibull, or log-logistic densities.
+The major modeling question in this setting, which we will return to below, is the choice of a function form (e.g.
+exponential, Weibull, or log-logistic) describing the underlying hazard rate $h(t) = \frac{f(t)}{S(t)}$ over time.
 
@@
 In the examples of applied research discussed above this assumption is untenable.
-Some cancer patients are cured after treatment, most young people never become regular smokers, and many contemporary states will not experience the kind of violence that persists in some parts of the world.
-The presence of a large sub-population which is not at risk for an event in practice will inflate estimates of the survival fraction, and reduce hazard estimates for all subjects by essentially, and falsely, distributing risk among subjects that genuinely will fail and those that are cured.
-A model will thus over predict hazard for subjects that are not at risk (cured), and under predict for those who are at risk of eventually experiencing the event of interest.
+Some cancer patients are cured after treatment, most young people never become regular smokers, and many states will not experience the kind of violence that persists in other parts of the world.
+The presence of a large sub-population which is not at risk for an event, will in practice inflate estimates of the survival fraction, and reduce hazard estimates for all subjects.
+This is the case because the underlying risk is estimated based on subjects that genuinely will fail and those that are cured.
+Hence, such a model will over predict the hazard for subjects that are not at risk (cured), and under predict for those who are at risk of eventually experiencing the event of interest.
 
@@
 For example, there is a clear sub-population in a model of the age at first pregnancy for humans---men---which researchers can easily identify and exclude from their data.
-Much less clear are answers to questions such as whether a cancer patient is cured or not cured given that they have no visible signs of cancer following treatment or have hit the 5-year disease free survival mark.
-In such situations split-population models provide a way to infer membership in sub-populations in a probabilistic fashion.
+Less clear are answers to questions such as whether a cancer patient is cured or not cured given that they have no visible signs of cancer following treatment or have hit the 5-year disease free survival mark.
+In such situations split-population models provide a way to infer sub-populations in a probabilistic fashion.
 
@@
 Where $z_i$ is a vector of covariates for a subject at a given time.
-For interpretation, it is important to note that with time-varying covariates, the risk (or cured) estimate for a subject is particular to a given time point rather than constant for that subject over all time periods in the spell.\footnote{We use ``spell'' to designate all time periods observed for a subject up to the failure time.
+For interpretation, it is important to note that with time-varying covariates, the risk (or cured) estimate for a subject is particular to a given time point rather than constant over all time periods in the spell.\footnote{We use ``spell'' to designate all time periods observed for a subject up to the failure time.
 Subjects can theoretically have multiple spells, e.g.
-cancer patients who go into remission and relapse more than once, or states that experience multiple civil war onsets over their history.} Depending on the covariates, the risk estimate for a subject can thus fluctuate rapidly over time.
+cancer patients who go into remission and relapse more than once, or states that experience multiple civil wars over their history.} Depending on the covariates, the risk estimate for a subject can thus fluctuate over time.
 To ease interpretation, it might be convenient to restrict covariates in the logit risk model to slow-moving, stable covariates in order to produce stable risk estimates for subjects.
@@
 
-With these, the main quantity of interest is the conditional hazard $h(t, \pi)$, where both the risk/cure probabilities and hazard are conditional on survival to time $t$:
+Given the density distribution, the main quantity of interest is the conditional hazard $h(t, \pi)$, where both the risk/cure probabilities and hazard are conditional on survival to time $t$:
 
@@
 
-For a given unconditional risk rate $\pi$, the probability that a particular case with survival time $t$ is in the risk set decreases over time because an increasing number of surviving cases consist of immune or cured ($1-\pi$) cases that will never fail.
+For a given unconditional risk rate $\pi$, the probability that a subject with survival time $t$ is in the risk set decreases over time, because an increasing number of surviving cases consist of immune or cured ($1-\pi$) cases that will never fail.
 In the hazard rate, the failure rate in the numerator is conditional on the probability that a case is in the risk set, given survival up to time $t$, and the numerator is an adjusted survivor function that accounts for the fraction of cured cases by time $t$, which is $1-\pi(t)$.

### 2015-12-11 manuscript.Rmd
update section 4

@@
 
-As an example to illustrate the package functionality, we will examine the model of coups d`etat in @belkin2003toward.
+In order to illustrate the package functionality, we examine a model of coups d`etat in @belkin2003toward.
 Belkin and Schofer's paper lends itself to re-analysis with a split-populuation duration model because they explicitly distinguish long-term structural risk factors for coups from more short-term triggering causes that can explain the timing of a coup in an at-risk regime.
 They argue that many countries never experience coups because coups are effectvely impossible due to structural factors, while others that never experience coups are nevertheless at risk due to a different configuration of those same factors.
-Using language which fits nicely with the class of models described above, they say "[t]riggers are not the source of the original risk, and in the absence of structural causes, the presence of triggering factors alone cannot lead to a coup.
+Using language which fits nicely with the class of models described above, they write "[t]riggers are not the source of the original risk, and in the absence of structural causes, the presence of triggering factors alone cannot lead to a coup.
 Hence, triggers should not be equated with coup risk.
@@
 
-For their empirical test Belkin and Schofer develop a measure of ``structural coup risk'' which incorporates factors such as the strength of democratic institutions and civil society, and a recent history of successful coups, but the model itself is implemented using a logistic regression model.
-This does not adequately capture the process described in the quote above, since it implicitly assumes that all observations are at risk for a coup (i.e., the probability of a coup is non-zero for all observations).
-Their structural coup risk indicator is developed precisely to distinguish between cases where coups are possible and cases where they are not, so the split-population model allows one to examine whether the indicator effectively separates at-risk cases from cases where coups are impossible.
+For their empirical test Belkin and Schofer develop a measure of ``structural coup risk'' which incorporates factors such as the strength of democratic institutions and civil society, and a recent history of successful coups.
+However, they implement a logistic regression model, which does not capture the process as described in their quote above.
+This is because the logistic regression model assumes that all observations are at risk for a coup (i.e., the probability of a coup is non-zero for all observations).
+Since their structural coup risk indicator is developed precisely to distinguish between at risk and ``immune'' cases, the split-population model allows one to examine whether the indicator effectively does so.
 
@@
 The \code{add\_duration} function takes as input a data frame with a binary response variable--"coup"--that measures the occurrence of the event, or failure, recorded over discrete time periods.
-We assume that the data frame is in a cross-sectional time-series format that consists of multiple observations over time of a number of subjects, in our case countries.
+We assume that the data frame is in a cross-sectional time-series format that consists of multiple observations over time for a number of subjects, in our case countries.
 
-Within the framework of duration modeling, these data conceptually consist of "spells", which are repated observations of a unit from the time they enter the data (left-censoring) or after they experienced the event of interest, until the next event or the end of observation (right censoring).
+Within the framework of duration modeling, these data conceptually consist of "spells", which are repeated observations of a unit from the time they enter the data (left-censoring) or after they experienced the event of interest, until the next event or the end of observation (right censoring).
 Table \ref{tab-ex} shows how a single country could have two spells during the observation period.
@@
 This is controlled with the \code{ongoing} argument.
-In some cases like civil war occurence, the *y* variable records all years during which a country experienced a civil war.
+In some cases like civil war occurrence, the *y* variable records all years during which a country experienced a civil war.
 Here \code{ongoing} should be set to true so that the models try to predict the onset of the civil war, but disregard ongoing conflicts (\code{failure} is set to \code{NA} for these cases, dropping them from analysis).
@@
 Table \ref{loglog_table} shows estimates from the duration equation, beginning with the intercept, and then estimates from the risk equation.
-The duration component of the model is in accelerated failure time form and the effect of coefficient estimates is on the log of expected time to failure.
+The duration component of the model is in accelerated failure time format and the coefficient estimates are on the log of expected time to failure.
 The negative coefficient for military regimes for example means that the expected time to a coup is shorter in military regimes than non-military regimes, holding all other factors constant.
-In the risk equation positive coefficients mean higher risk of coup, and military regimes thus also have a higher risk of experiencing any coup at all.
+In the risk equation, positive coefficients mean higher risk of coup.
+Thus, military regimes have a higher risk of experiencing a coup.
 
 One may also use the \code{AIC} and \code{BIC} function to calculate the information criterion statistics for \code{spdur} objects.
-In our example both models are close enough in both statistics to be indistinguishable, so we will continue to focus on the log-logistic form.
+In our example, both models are close enough in both statistics to be indistinguishable, so we will continue to focus on the log-logistic form.
 
@@
 
-As mean values are maybe not that interesting, we can also supply our values for a scenario of choice.
+As mean values are maybe not that interesting, we can also generate scenarios by entering values manually.
 The values are used in the same order that variables are specified in the equations used to estimate the models, which also corresponds to the order of variables in Table 2.
-Including an intercept term.
+Note the inclusion of an intercept term.
 This produces plot B in Figure \ref{hazard-ex}

### 2015-12-11 manuscript.Rmd
update last sections

@@
 Finally, we demonstrate how to evaluate a model's out of sample predictions.
-We will use data from 1996 on as the test set, and the prior data for training.
+We will use data from 1996 onwards as the test set, and prior data for training purposes.
 The \code{add_duration} function retrospectively codes the risk variable based on how a spell ended, and we therefore need to take care in how we add the duration variables for each data set.
-For the training data we need to subset the training set first so that coups in the tests set don't influence the risk coding in the training data.
+For the training data we need to subset the training set first, so that coups in the tests set don't influence the risk coding in the training data.
 
@@
 
-For the test set it is better to add the duration variables and *then* subset the test set.
+For the test set it is recommended to add the duration variables and *then* subset the test set.
 Since the test set is later in time than the training set we do not have to worry about contamination of the risk coding, but if we subset the data before building the duration variables we will start all duration counters at 1996, when in fact we can safely use the previous historic coup information.
@@
 
-Truncation and censoring are problematic for split-population duration models as they are for standard duration regression, but also pose some additional considerations.
+Truncation and censoring are problematic for split-population duration models, as they are for standard duration regression, but also pose some additional considerations.
 In left-truncation, we do not observe data for a spell prior to some date, and thus have incomplete and inaccurate values for the duration or time to failure for a spell.
@@
 Secondly, and as a result of the first point, data should be partitioned in a scheme that samples either by spell or block-wise, e.g.
-by witholding the last **x** years of data, in order to be able to proberly build the duration variables.
+by witholding the last **x** years of data, in order to properly build the duration variables.
 
@@
 Many outcomes of interest are rare events.
-Coups, the onset of war and civil war, mass killings, and so on are exceedingly rare events when one considers that most countries most of the time are stable.
-Figure \ref{rates} shows a few examples for published research that models binary outcomes.
+For example, coups, war onset, or mass killings are exceedingly rare events when considering that most countries do not experience high levels of violence.
+Figure \ref{rates} shows a few examples of published research that models binary outcomes.
 @fearon2003ethnicity is a widely cited study of civil war onset that uses yearly observations of all major countries; @beger2014ensemble is an example of the positive rates when moving to monthly data for a similar set of countries.
@@
 It is a well-recognized problem and has led to the development or use of several specialized mixture models like zero-inflated Poisson and negative binomial regression for count data, and a zero-inflated ordered probit for ordinal outcomes [@bagozzi2015modeling].
-Split-population duration regression provides another principled solution to the challenges posed by data in this domain, but, unlike other solutions to the sparse outcome problem, also addresses underlying temporal dynamics that are an important part of the non-independent data political scientists commonly use.
+Split-population duration regression provides another principled solution to the challenges posed by data in this domain, but, unlike other solutions to the sparse outcome problem, also addresses underlying temporal dynamics that are an important part of the non-independent data political scientists and other social scientists commonly use.
 
-Split-population duration models are not only appealing in a technical sense, but they also match the logic or intuituion many analysts of politics use when they distinguish long-term risk factors from more fleeting triggering causes.
+Split-population duration models are not only appealing in a technical sense, but they also match the logic or intuition many social scientists use when they distinguish long-term risk factors from more fleeting triggering causes.
 The example we have used, @belkin2003toward, is a particularly clear illustration of how well the language of theorists maps onto the model intuition.

### 2015-12-18 manuscript.Rmd
added King and Zeng
added King and Zeng cites
catching up

@@
     address: >
-      First line
+      Department of International Affairs, University of Georgia
       Athens, GA, USA
@@
     address: >
-      First line
+      Department of Political Science, University College London
       London, UK
     email: n.metternich@ucl.ac.uk
+  - name: Shahryar Minhas
+    affiliation: Duke University
+    address: >
+      Department of Political Science, Duke University
+      Durham, NC
+    email: sfm12@duke.edu
   - name: Michael D. Ward
@@
 @fearon2003ethnicity is a widely cited study of civil war onset that uses yearly observations of all major countries; @beger2014ensemble is an example of the positive rates when moving to monthly data for a similar set of countries.
-The positive rates range from 5 to less than 0.2\% of all data points.
+The positive rates range from 5 to less than 0.2\% of all data points.\footnote{For a discussion of the difficulties rare events can pose for prediction see \citet{king2001logistic} and \citet{king2001explaining}.}
 

### 2015-12-18 manuscript.Rmd
fix typos and footnote 3

--- 2015-12-11 manuscript.Rmd
@@
     address: >
-      First line
+      Department of Political Science, University College London
       London, UK
     email: n.metternich@ucl.ac.uk
+  - name: Shahryar Minhas
+    affiliation: Duke University
+    address: >
+      Department of Political Science, Duke University
+      Durham, NC, USA
+    email: sfm12@duke.edu
   - name: Michael D. Ward
@@
 workplace) determine whether individuals catch the flu.
-Under these assumptions a standard duration model and should provide us with efficient estimates of how different behavioral and structural factors impact on the general baseline risk.
-However, in the overall population there might be individuals that are immune to the flu, because they received vaccination, had the specific flu virus in the past, or have some other characteristic that makes it impossible for them to attract the decease.
-In this instance we have two underlying populations: An at risk population and an immune one.
+Under these assumptions a standard duration model should provide us with efficient estimates of how different behavioral and structural factors impact on the general baseline risk.
+But, there might be individuals that are immune to the flu in the overall population--because they received vaccination, had the specific flu virus in the past, or have some other characteristic that makes it impossible for them to attract the disease.
+In this instance we have two underlying populations: an at risk population and an immune one.
 If the immune population is relatively large, estimates using a standard duration model will be biased and predictions inaccurate.
@@
 
-Duration models, where baseline risk is modeled by some distribution of time, were originally developed in health and demographic research and grew naturally from life tables and survival records for medical patients.
+Regular duration models, where baseline risk is modeled by some distribution of time, were originally developed in health and demographic research and grew naturally from life tables and survival records for medical patients.
 Basic formulations of such models, like the parametric exponential or Weibull regressions or semi-parametric Cox regression, implicitly assume that all subjects or units under observation, including right-censored observations, will eventually experience the event of interest.
@@
 In their application some fraction of patients survived because their cancer was cured, while others relapsed after apparent remission due to levels of disease below detectable thresholds.
-The development of duration methods in health and medicine has also shaped the terminology conventionally used for such model.
+The development of duration methods in health and medicine has also shaped the terminology conventionally used for such models.
 For example, split-population duration models are also referred to as cure rate models.
-Similarly, the basic concepts like survival and failure rates reference to the survival of humans.
+Similarly, the basic concepts like survival and failure rates reference the survival of humans.
 
@@
 In order to illustrate the package functionality, we examine a model of coups d`etat in @belkin2003toward.
-Belkin and Schofer's paper lends itself to re-analysis with a split-populuation duration model because they explicitly distinguish long-term structural risk factors for coups from more short-term triggering causes that can explain the timing of a coup in an at-risk regime.
-They argue that many countries never experience coups because coups are effectvely impossible due to structural factors, while others that never experience coups are nevertheless at risk due to a different configuration of those same factors.
+Belkin and Schofer's paper lends itself to re-analysis with a split-population duration model because they explicitly distinguish long-term structural risk factors for coups from more short-term triggering causes that can explain the timing of a coup in an at-risk regime.
+They argue that many countries never experience coups because coups are effectively impossible due to structural factors, while others that never experience coups are nevertheless at risk due to a different configuration of those same factors.
 Using language which fits nicely with the class of models described above, they write "[t]riggers are not the source of the original risk, and in the absence of structural causes, the presence of triggering factors alone cannot lead to a coup.
@@
 
-For their empirical test Belkin and Schofer develop a measure of ``structural coup risk'' which incorporates factors such as the strength of democratic institutions and civil society, and a recent history of successful coups.
+For their empirical test Belkin and Schofer develop a measure of "structural coup risk" which incorporates factors such as the strength of democratic institutions and civil society, and a recent history of successful coups.
 However, they implement a logistic regression model, which does not capture the process as described in their quote above.
 This is because the logistic regression model assumes that all observations are at risk for a coup (i.e., the probability of a coup is non-zero for all observations).
-Since their structural coup risk indicator is developed precisely to distinguish between at risk and ``immune'' cases, the split-population model allows one to examine whether the indicator effectively does so.
+Since their structural coup risk indicator is developed precisely to distinguish between at risk and "immune" cases, the split-population model allows one to examine whether the indicator effectively does so.
 
@@
 
-For the test set it is recommended to add the duration variables and *then* subset the test set.
+For the test set it is recommended to add the duration variables and \textit{then} subset the test set.
 Since the test set is later in time than the training set we do not have to worry about contamination of the risk coding, but if we subset the data before building the duration variables we will start all duration counters at 1996, when in fact we can safely use the previous historic coup information.
@@
 
-Furthermore, in out-of-sample testing based on some kind of data partitioning scheme, this coding scheme can lead to unintentioned contamination of in-sample cases with knowledge of out-of-sample failures through the risk coding for failed spells.
+Furthermore, as the example above shows, in out-of-sample testing based on some kind of data partitioning scheme, this coding scheme can lead to unintentional contamination of in-sample cases with knowledge of out-of-sample failures through the risk coding for failed spells.
 This leads to two recommendations.
-First, the duration data (`add_duration`) should not be built until after partitioning the data.
-Secondly, and as a result of the first point, data should be partitioned in a scheme that samples either by spell or block-wise, e.g.
-by witholding the last **x** years of data, in order to properly build the duration variables.
+First, data should be partitioned spell or block-wise, e.g.
+by withholding the last *x* years of data, and not randomly.
+Second, given the two concerns of left-censoring and non-independence induced through the duration and risk coding, care should be taken to ensure that duration data are built without access to future information in another data partition.
 
@@
 @fearon2003ethnicity is a widely cited study of civil war onset that uses yearly observations of all major countries; @beger2014ensemble is an example of the positive rates when moving to monthly data for a similar set of countries.
-The positive rates range from 5 to less than 0.2\% of all data points.
+The positive rates range from 5 to less than 0.2\% of all data points.[^3]
+
+[^3]: For a discussion of the difficulties rare events can pose for prediction see @king2001logistic and @king2001explaining.
 

### 2016-02-19 manuscript.Rmd
style and bibliography changes for JSS

@@
 title:
-  formatted: "Splitting it up: the \\pkg{spduration} split-population duration regression package"
+  formatted: "Splitting it Up: The \\pkg{spduration} Split-population Duration Regression Package"
   # If you use tex in the formatted title, also supply version without
@@
 Consider the risk of acquiring a viral infection like the common flu.
-Let us initially assume that everyone is at risk of infection, and individual behavior (e.g.
-good hygiene) and structural factors (e.g.
-workplace) determine whether individuals catch the flu.
+Let us initially assume that everyone is at risk of infection, and individual behavior (e.g., good hygiene) and structural factors (e.g., workplace) determine whether individuals catch the flu.
 Under these assumptions a standard duration model should provide us with efficient estimates of how different behavioral and structural factors impact on the general baseline risk.
@@
 
-The major modeling question in this setting, which we will return to below, is the choice of a function form (e.g.
-exponential, Weibull, or log-logistic) describing the underlying hazard rate $h(t) = \frac{f(t)}{S(t)}$ over time.
+The major modeling question in this setting, which we will return to below, is the choice of a function form (e.g., exponential, Weibull, or log-logistic) describing the underlying hazard rate $h(t) = \frac{f(t)}{S(t)}$ over time.
 
@@
 For interpretation, it is important to note that with time-varying covariates, the risk (or cured) estimate for a subject is particular to a given time point rather than constant over all time periods in the spell.\footnote{We use ``spell'' to designate all time periods observed for a subject up to the failure time.
-Subjects can theoretically have multiple spells, e.g.
-cancer patients who go into remission and relapse more than once, or states that experience multiple civil wars over their history.} Depending on the covariates, the risk estimate for a subject can thus fluctuate over time.
+Subjects can theoretically have multiple spells, e.g., cancer patients who go into remission and relapse more than once, or states that experience multiple civil wars over their history.} Depending on the covariates, the risk estimate for a subject can thus fluctuate over time.
 To ease interpretation, it might be convenient to restrict covariates in the logit risk model to slow-moving, stable covariates in order to produce stable risk estimates for subjects.
@@
 data(bscoup)
-bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
-bscoup      <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
-                            freq="year", ongoing = FALSE)
+bscoup$coup <- ifelse(bscoup$coup == "yes", 1, 0)
+bscoup      <- add_duration(bscoup, "coup", unitID = "countryid", tID = "year",
+                            freq = "year", ongoing = FALSE)
 ```
@@
 \begin{center}
-\caption{Example data frame with select duration variables}
-\label{tab-ex}
 \begin{tabular}{lrrrrr} \toprule
@@
 \end{center}
+\caption{Example data frame with select duration variables} \label{tab-ex}
 \end{table}
@@
   atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
-  data=bscoup, silent = TRUE)
+  data = bscoup, silent = TRUE)
 
@@
   atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
-  data=bscoup, distr="loglog", silent = TRUE)
+  data = bscoup, distr = "loglog", silent = TRUE)
 ```
@@
 library("xtable")
-tbl <- xtable(loglog_model, caption="Coup model with log-logistic hazard",
-              label="loglog_table")
-print(tbl, caption.placement="top", comment=FALSE, include.rownames=FALSE)
+tbl <- xtable(loglog_model, caption = "Coup model with log-logistic hazard",
+              label = "loglog_table")
+print(tbl, caption.placement = "top", comment = FALSE, include.rownames = FALSE)
 ```
@@
 \centering
-\caption{Coup model with log-logistic hazard} 
-\label{loglog_table}
 \begin{tabular}{lrrr}
@@
 \end{tabular}
+\caption{Coup model with log-logistic hazard} \label{loglog_table}
 \end{table}
@@
 data(bscoup)
-bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
+bscoup$coup <- ifelse(bscoup$coup == "yes", 1, 0)
 coup_train <- bscoup[bscoup$year < 1996, ]
-coup_train <- add_duration(coup_train, "coup", unitID="countryid", tID="year",
-                          freq="year", ongoing = FALSE)
+coup_train <- add_duration(coup_train, "coup", unitID = "countryid", tID = "year",
+                          freq = "year", ongoing = FALSE)
 ```
@@
 ```{r, include = FALSE}
-coup_test  <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
-                           freq="year", ongoing = FALSE)
+coup_test  <- add_duration(bscoup, "coup", unitID = "countryid", tID = "year",
+                           freq = "year", ongoing = FALSE)
 coup_test  <- coup_test[coup_test$year >= 1996, ]
@@
 ```{r, eval = FALSE}
-coup_test  <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
-                           freq="year", ongoing = FALSE)
+coup_test  <- add_duration(bscoup, "coup", unitID = "countryid", tID = "year",
+                           freq = "year", ongoing = FALSE)
 coup_test  <- coup_test[coup_test$year >= 1996, ]
@@
   atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
-  data = coup_train, distr="loglog", silent = TRUE) 
+  data = coup_train, distr = "loglog", silent = TRUE) 
 ```
@@
 
-par(mfrow=c(2,1),mar=c(2,2,2,2))
+par(mfrow = c(2,1),mar = c(2,2,2,2))
 separationplot(weib2_test_p,   obs_y)
@@
 This leads to two recommendations.
-First, data should be partitioned spell or block-wise, e.g.
-by withholding the last *x* years of data, and not randomly.
+First, data should be partitioned spell or block-wise, e.g., by withholding the last *x* years of data, and not randomly.
 Second, given the two concerns of left-censoring and non-independence induced through the duration and risk coding, care should be taken to ensure that duration data are built without access to future information in another data partition.

### 2016-03-06 manuscript.Rmd
set new plot = false for separation plot
add manuscript2.tex and pre-screening-commends

@@
 ---
+
+```{r}
+stop("I've started editing the tex manually")
+```
 
@@
 
-```{r data, warning=FALSE}
+```{r data, warning=FALSE, include=FALSE}
 library("spduration")
@@
 \end{center}
-\caption{Example data frame with select duration variables} \label{tab-ex}
+\caption{Example data frame with select duration variables.} \label{tab-ex}
 \end{table}
@@
 \end{tabular}
-\caption{Coup model with log-logistic hazard} \label{loglog_table}
+\caption{Coup model with log-logistic hazard.} \label{loglog_table}
 \end{table}
@@
 \includegraphics[width=0.8\textwidth]{graphics/sepplots.pdf}
-\caption{In-sample separation plots of Weibull and log-logistic model conditional hazard predictions}
+\caption{In-sample separation plots of Weibull and log-logistic model conditional hazard predictions.}
 \label{insamp}
@@
 
-par(mfrow = c(2,1),mar = c(2,2,2,2))
-separationplot(weib2_test_p,   obs_y)
-separationplot(loglog2_test_p, obs_y)
+par(mfrow=c(2,1),mar=c(2,2,2,2))
+separationplot(weib2_test_p,   obs_y, newplot = FALSE)
+separationplot(loglog2_test_p, obs_y, newplot = FALSE)
 ```
@@
 \includegraphics[width=0.8\textwidth]{graphics/oos-sepplots.pdf}
-\caption{Out-of-sample separation plots}
+\caption{Out-of-sample separation plots.}
 \label{oos-sepplots}
@@
 \includegraphics[width = 4in]{graphics/rates.pdf}
-\caption{Rates of positive outcomes in select publications with binary outcomes}
+\caption{Rates of positive outcomes in select publications with binary outcomes.}
 \label{rates}

### 2016-03-06 pre-screening-comments.Rmd
add manuscript2.tex and pre-screening-commends

---
title: "Pre-screening comments"
author: "Andreas Beger"
date: "March 6, 2016"
output: pdf_document
---

JSS 2513: Beger, Hill, Metternich, Minhas, Ward

Splitting it up: the spduration split-population duration regression package

---------------------------------------------------------
For further instructions on JSS style requirements please see the Style Files (in particular section 2.1 Style Checklist of jss.pdf) and FAQ at http://www.jstatsoft.org/about/submissions.

For further examples please see RECENT JSS papers.
---------------------------------------------------------

From the editorial team:

o The style instructions have not been taken into account appropriately. An `\AND` rather `\And` is needed somewhere in `\author`. Code is not formatted correctly. The bibliography must be produced via BibTeX. (See additional comments below.)

*Changed `And` to `\AND`, added `R>` and `+` where needed to code formatting, and bibliography is not via BibTeX.*

o The xtable for Figure 2 lacks math markup for negative numbers. Also row labels need `\code` markup for JSS and column labels could be formatted better.

o There should be some discussion of software for related purposes (including references), pointing out relative (dis)advantages.

*Added discussion of `smcure`, which is the only alternative implementation for R that we could find.*

o Some standard methods are available for spdur objects but more would be desired, e.g., coef, vcov, possibly fitted, residuals, model.matrix etc.

*We have added methods for these to a new package version, which is up on CRAN.*

o The replication material does not seem to be complete. Please provide all additional files needed to successfully run the code. It should then also be stated how these data files could be generated / how they were obtained.

o The source package (.tar.gz) rather than the git .zip should be included in the submission.

Manuscript style comments:

o Title should be in title style (do capitalize all principal words and the first word after a colon).

o For the code layout in R publications, we typically distinguish input/output using Sinput/Soutput (or equivalently CodeInput/CodeOutput). Unless there are special reasons to format it differently, the input should use the “R>” command prompt and the text width (up to 76 or 77 characters), and be indented by two spaces, e.g.,

```latex
begin{Sinput}
R> example_model <- lm(response ~ variable1 + variable2 + variable3, 
+    weights = w, data = mydata)
\end{Sinput}
```

*Adjusted code accordingly.*

o Code should have enough spaces to facilitate reading.  Please include spaces before and after operators and after commas (unless spaces have syntactical meaning).

o If using "e.g." and "i.e." add a comma after the period to keep LaTeX from interpreting them as the end of a sentence, i.e.: "e.g., " and "i.e., ".

*Done*

o All captions should appear below the corresponding figure/table. The captions should be in sentence style and end with a period.

*All captions are below figure/table, in sentence case, and end with period now.*

o As a reminder, please make sure that:
	- `\proglang`, `\pkg` and `\code` have been used for highlighting throughout the paper (including titles and references), except where explicitly escaped.

References:

o Journal of the Royal Statistical Society B (not: Journal of the Royal Statistical Society. Series B)

o Please make sure that all software packages are \cite{}'d properly.

o All references should be in title style.

o See FAQ for specific reference instructions.

Code:

o As mentioned above, please make sure that the files needed to replicate all code/examples within the manuscript are included in a standalone replication script.

### 2016-03-07 manuscript.tex
how about actually syncing manuscript2.tex
also expanded alternative implementation section

@@
 \author{
-Daniel W. Hill\\University of Georgia \And Nils W. Metternich\\University College London \And Andreas Beger\\Duke University \And Michael D. Ward\\Duke University
+Andreas Beger\\Duke University \And Daniel W.
+Hill, Jr.\\University of Georgia \AND Nils W.
+Metternich\\University College London \And Shahryar Minhas\\Duke University \And Michael D.
+Ward\\Duke University
 }
-\title{Splitting it up: the \pkg{spduration} split-population duration
-regression package}
-\Keywords{keywords, not capitalized, \proglang{R}}
+\title{Splitting it Up: The \pkg{spduration} Split-population Duration
+Regression Package}
+\Keywords{regression, rare events, \proglang{R}}
 
@@
 we can significantly increase the predictive performance compared to
-standard duration model. The package includes estimation and forecasting
-methods for split-population Weibull and Loglogistic models.
+standard duration models. The package includes estimation and several
+post-estimation methods for split-population Weibull and Loglogistic
+models.
 }
 
-\Plainauthor{Daniel W. Hill, Nils W. Metternich, Andreas Beger, Michael D. Ward}
+\Plainauthor{Andreas Beger, Daniel W. Hill, Jr., Nils W. Metternich, Shahryar Minhas, Michael D. Ward}
 \Plaintitle{Splitting it up: the spduration split-population duration regression
@@
 \Shorttitle{\pkg{spduration}: Split-population duration regression}
-\Plainkeywords{keywords, not capitalized, R}
+\Plainkeywords{regression, rare events, R}
 
@@
 \Address{
-    Daniel W. Hill\\
+    Andreas Beger\\
+  Duke University\\
+  Department of Political Science, Duke University Durham, NC, USA\\
+  E-mail: \href{mailto:ab428@duke.edu}{\nolinkurl{ab428@duke.edu}}\\
+  URL: something\\~\\
+      Daniel W. Hill, Jr.\\
   University of Georgia\\
-  First line Second line\\
+  Department of International Affairs, University of Georgia Athens, GA,
+  USA\\
   E-mail: \href{mailto:dwhill@uga.edu}{\nolinkurl{dwhill@uga.edu}}\\
-  URL: \url{http://spia.uga.edu/directory/faculty_staff/hill-daniel}\\~\\
+
+      Nils W. Metternich\\
+  University College London\\
+  Department of Political Science, University College London London, UK\\
+  E-mail: \href{mailto:n.metternich@ucl.ac.uk}{\nolinkurl{n.metternich@ucl.ac.uk}}\\
+
+      Shahryar Minhas\\
+  Duke University\\
+  Department of Political Science, Duke University Durham, NC, USA\\
+  E-mail: \href{mailto:sfm12@duke.edu}{\nolinkurl{sfm12@duke.edu}}\\
+
+      Michael D. Ward\\
+  Duke University\\
+  Department of Political Science, Duke University Durham, NC, USA\\
+  E-mail: \href{mailto:mw160@duke.edu}{\nolinkurl{mw160@duke.edu}}\\
+
   }
 
-\usepackage{amsmath}
+\usepackage{amsmath} \usepackage{booktabs}
 
@@
 
-This template demonstrates some of the basic latex you'll need to know
-to create a JSS article.
+Duration models are an important class of statistical estimators that
+take into account the duration dependency of specific outcomes. A
+prominent example is that the risk of dying depends on the age of an
+individual. Newborns are at a greater risk of dying, but as they grow
+older this risk quickly declines and then gradually starts to increase
+again after the age of 9-10. While individual behavior (smoking,
+exercising, diet) and structural factors (health care, health
+regulations, urban vs rural) can increase or decrease the probability of
+dying over time, the underlying risk is time dependent and impacts on
+all individuals.
 
-\subsection{Code formatting}\label{code-formatting}
-
-Don't use markdown, instead use the more precise latex commands:
-
-\begin{itemize}
-\itemsep1pt\parskip0pt\parsep0pt
-\item
-  \proglang{Java}
-\item
-  \pkg{plyr}
-\item
-  \code{print("abc")}
-\end{itemize}
-
-\section{Introduction}\label{introduction-1}
-
-Duration models are an important class of statistical estimators that
-take into account the duration dependency of outcomes we are interested
-in. For example, the risk of dying is not independent of time. Newborns
-are at a great risk of dying, but as they grow older this risk declines
-and then gradually starts to increase again after the age of 9-10. This
-risk can be increased or decreased by behavioral (smoking, exercise,
-diet, etc) and structural factors (health care, regulations, urban vs
-rural, etc), but there is an underlying risk that is time dependent and
-impacts on all units.
-
-However, sometimes we are interested in situations in which not all
-units are at risk of dying, failing, or acquiring some new
-characteristic. Imagine we are interested in the risk of acquiring the
-flu. First, we might think that everyone is at risk of getting the flu
-and it just depends on individual behavior (e.g.~good hygiene) and
-structural factors (e.g.~workplace) whether individuals will acquire the
-flu. If this would be the case we could use a standard duration model
-and estimate how different factors impact on the baseline risk. But
-there might also be individuals who are immune to the flu, because they
-received a flu shot, had the same flu virus in the past year, or some
-other factors that makes it impossible for them to get sick. Given that
-this is the case we have two underlying populations: An at risk
-population and an immune one. If the immune population is relatively
-large, estimates using a standard duration model will be biased and
-predictions resulting from such a model inaccurate.
+However, there are conditions under which not all individuals have the
+same underlying risk to experience a specific outcome and might not even
+be at risk at all. Consider the risk of acquiring a viral infection like
+the common flu. Let us initially assume that everyone is at risk of
+infection, and individual behavior (e.g., good hygiene) and structural
+factors (e.g., workplace) determine whether individuals catch the flu.
+Under these assumptions a standard duration model should provide us with
+efficient estimates of how different behavioral and structural factors
+impact on the general baseline risk. But, there might be individuals
+that are immune to the flu in the overall population--because they
+received vaccination, had the specific flu virus in the past, or have
+some other characteristic that makes it impossible for them to attract
+the disease. In this instance we have two underlying populations: an at
+risk population and an immune one. If the immune population is
+relatively large, estimates using a standard duration model will be
+biased and predictions inaccurate.
 
@@
 
-Duration models, where the variable to be modeled describes a
-distribution of times until some event of interest, were originally
-developed in health and demographic research, where they grew naturally
-from life tables and survival records for medical patients. Basic
-formulations of such models, like the parametric exponential or Weibull
-regressions or semi-paramaetric Cox regression, implicitly assume that
-all subjects or units under observation, including right-censored
-observations, will eventually experience the event of interest. This
-assumption is outright wrong or at least disadvantageous in many
-substantive areas and empirical applications, where there instead is a
-sub-population of units or subjects that will never experience an event,
+Regular duration models, where baseline risk is modeled by some
+distribution of time, were originally developed in health and
+demographic research and grew naturally from life tables and survival
+records for medical patients. Basic formulations of such models, like
+the parametric exponential or Weibull regressions or semi-parametric Cox
+regression, implicitly assume that all subjects or units under
+observation, including right-censored observations, will eventually
+experience the event of interest. This assumption may be violated in
+many substantive areas and empirical applications, where a
+sub-population of units or individuals will never experience an event,
 and thus are effectively ``cured''.
 
-This insight was realized as early as 1949 by Boag (1949) and Berkson
-and Gage (1952), who were researching survival rates in cancer patients
-following treatment, and where quite obviously some fraction of patients
-survived because their cancer was cured, while others relapsed after
-apparent remission due to levels of disease below detectable thresholds.
-As with conventional duration models, their origin in health and
-medicine has shaped the terminology conventionally used with such model,
-e.g.~split-population duration models are sometimes also referred to in
-such contexts as cure rate models, and the basic concepts like survival
-and failure rates reference the survival of humans.
+The insight that populations might be split in regard to their baseline
+risk has been formulated as early as 1949 by Boag (1949) and Berkson and
+Gage (1952), who were researching survival rates in cancer patients. In
+their application some fraction of patients survived because their
+cancer was cured, while others relapsed after apparent remission due to
+levels of disease below detectable thresholds. The development of
+duration methods in health and medicine has also shaped the terminology
+conventionally used for such models. For example, split-population
+duration models are also referred to as cure rate models. Similarly, the
+basic concepts like survival and failure rates reference the survival of
+humans.
 
@@
 medicine. In an early and foundational application that reached beyond
-criminal science, Schmidt and Witte (1989) examined criminal recidivism
-using data on close to 10,000 prisoners from the North Carolina prison
-system in the late 1970's and early 1980's to identify factors that
-influence whether a criminal lapses at all, and if so which factors are
-related to the amount of time between prison stints. This work already
-includes a full formulation of the model with independent covariates for
-both the duration equation and the risk or cure equation, although only
-with subject-specific covariates rather than time-varying covariates
-with multiple data points per subject.
+medicine, Schmidt and Witte (1989) examined criminal recidivism using
+data on close to 10,000 prisoners from the North Carolina prison system
+in the late 1970's and early 1980's to identify factors that influence
+whether a criminal relapses at all, and if so which factors are related
+to the amount of time between prison stints. This work already includes
+a full formulation of a model with independent covariates for both the
+duration equation and the risk or cure equation, although only with
+subject-specific covariates rather than time-varying covariates for
+multiple data points per individual.
 
-In public health, Douglas and Hariharan (1994) use data from the US to
-model factors related to the age at which smokers started the habit, and
-Forster and Jones (2001) examines the impact of tobacco taxes on smoking
-and quitting decisions. DeYoung (2003) models the failure of new
-commercial banks in the US during the 1980's. Lastly, in our own domain,
-political science, Svolik (2008) used a split-population duration
-framework to study democratic consolidation, i.e.~whether democratic
-regimes persist or slide back to authoritarianism, and when. Building on
-this effort, split-population duration models have also been used, along
-with other models, to produce regular predictions for five different
-forms of political conflict for the Integrated Crisis and Early Warning
-System (ICEWS) project (M. D. Ward et al. 2013) and to model irregular
-leadership changes for the Political Instability Task Force (PITF;
-Beger, Dorff, and Ward 2014).
+In a complementary set of research from public health, Douglas and
+Hariharan (1994) use data from the United States to model the age at
+which individuals began to smoke, and Forster and Jones (2001) examine
+the impact of tobacco taxes on smoking and quitting decisions. DeYoung
+(2003), an economist, models the failure of new commercial banks in the
+US during the 1980's. Svolik (2008) uses a split-population duration
+framework to examine whether democratic regimes persist or revert to
+authoritarianism, and when. Building on this effort, split-population
+duration models have also been used, along with other models, to produce
+regular predictions for five different forms of political conflict for
+the Integrated Crisis and Early Warning System (ICEWS) project (Ward et
+al. 2013) and to model irregular leadership changes for the Political
+Instability Task Force (PITF; Beger, Dorff, and Ward 2014).
 
@@
 Conventional duration models assume that all subjects will eventually
-fail, if we can only observe them long enough. The likelihood for a data
-point with survival time \(t\) is thus the failure rate at that time or
-the probability of survival beyond \(t\), depending on whether the
-observation is right-censored (\(1-\delta_i\)) or not:
+fail, die, or experience a specific outcome. The likelihood for a data
+point with survival time \(t\) is thus the failure rate \(f(t)\) at that
+time or the probability of survival beyond \(t\), \(S(t)\), depending on
+whether the subject has already experienced the event (\(\delta_i\)) or
+is right-censored (\(1-\delta_i\)):
 
@@
 The major modeling question in this setting, which we will return to
-below, is the choice of a function to describe the evolution of the
-hazard rate \(h(t) = \frac{f(t)}{S(t)}\) over time, e.g.~with
-exponential, Weibull, or log-logistic densities.
+below, is the choice of a function form (e.g., exponential, Weibull, or
+log-logistic) describing the underlying hazard rate
+\(h(t) = \frac{f(t)}{S(t)}\) over time.
 
 The cumulative failure rate (\(F(t) = 1 - S(t)\)) over time converges to
-1, meaning all subjects fail eventually. This assumption is untenable in
-many situations. Some cancer patients are cured after treatment while
-others relapse, most young people do not start smoking even though most
-smokers started at a young age, and many states will not experience the
-kind of violence that persists in some parts of the world. A better
-assumption in such contexts is that there is a subpopulation of subjects
-who are at risk of experiencing an event at some point, and another
-subpopulation who will never experience the event.
-
-The presence of a large sub-population which is not at risk for an event
-in practice will inflate estimates of the survival fraction, and reduce
-hazard estimates for all subjects by essentially, and falsely,
-distributing risk among subjects that genuinely will fail and those that
-are cured. A model will thus over predict hazard for subjects that are
-not at risk (cured), and under predict for those who are at risk of
-eventually experiencing the event of interest.
+1, meaning all subjects fail eventually. In the examples of applied
+research discussed above this assumption is untenable. Some cancer
+patients are cured after treatment, most young people never become
+regular smokers, and many states will not experience the kind of
+violence that persists in other parts of the world. The presence of a
+large sub-population which is not at risk for an event, will in practice
+inflate estimates of the survival fraction, and reduce hazard estimates
+for all subjects. This is the case because the underlying risk is
+estimated based on subjects that genuinely will fail and those that are
+cured. Hence, such a model will over predict the hazard for subjects
+that are not at risk (cured), and under predict for those who are at
+risk of eventually experiencing the event of interest.
 
@@
 contexts where sub-populations are not clearly or easily identifiable.
-Even though there is a clear sub-populations in a model of the age at
-first pregnancy for humans---men---no researcher in their right mind
-would think to include male subjects in their data at all, for example.
-On the other hand, whether a cancer patient is cured or not cured given
-that they have no visible signs of cancer following treatment or have
-hit the 5-year disease free survival mark is much less clear and in fact
-the question that initially led to the idea for split-population
-duration modeling (Boag 1949, Berkson and Gage (1952)).
+For example, there is a clear sub-population in a model of the age at
+first pregnancy for humans---men---which researchers can easily identify
+and exclude from their data. Less clear are answers to questions such as
+whether a cancer patient is cured or not cured given that they have no
+visible signs of cancer following treatment or have hit the 5-year
+disease free survival mark. In such situations split-population models
+provide a way to infer sub-populations in a probabilistic fashion.
 
@@
 covariates, the risk (or cured) estimate for a subject is particular to
-a given time point rather than constant for that subject over all time
-periods in the
+a given time point rather than constant over all time periods in the
 spell.\footnote{We use ``spell'' to designate all time periods observed for a subject up to the failure time.
-Subjects can theoretically have multiple spells, e.g.
-cancer patients who go into remission and relapse more than once, or states that experience multiple civil war onsets over their history.}
+Subjects can theoretically have multiple spells, e.g., cancer patients who go into remission and relapse more than once, or states that experience multiple civil wars over their history.}
 Depending on the covariates, the risk estimate for a subject can thus
-fluctuate rapidly over time. To ease interpretation, it might be
-convenient to restrict covariates in the logit risk model to
-slow-moving, stable covariates in order to produce stable risk estimates
-for subjects.
+fluctuate over time. To ease interpretation, it might be convenient to
+restrict covariates in the logit risk model to slow-moving, stable
+covariates in order to produce stable risk estimates for subjects.
 
@@
 distribution for the shape of the hazard rate. The \pkg{spduration}
-package implements two hazard rate shapes,
-Weibull\footnote{In the code, $f(t) = \alpha \lambda^{\alpha} t^{\alpha - 1} e^{-(\lambda t)^\alpha}$.}
-and log-logistic:
+package implements two hazard rate shapes, Weibull and log-logistic:
 
@@
 
-Where \(\lambda = e^{-x_i\beta}\) is a parameter of covariates.
+Where \(\lambda = e^{-x_i\beta}\) is a parameter of covariates. The
+Weibull density allows for hazard rates that are increasing, constant,
+or decreasing over survival time, while the log-logistic density also
+can fit rates that have a peak at a particular survival time.
 
-With these, the main quantity of interest is the conditional hazard
-\(h(t, \pi)\), where both the risk/cure probabilities and hazard are
-conditional on survival to time \(t\):
+Given the density distribution, the main quantity of interest is the
+conditional hazard \(h(t, \pi)\), where both the risk/cure probabilities
+and hazard are conditional on survival to time \(t\):
 
@@
 For a given unconditional risk rate \(\pi\), the probability that a
-particular case with survival time \(t\) is in the risk set decreases
-over time because an increasing number of surviving cases consist of
-immune or cured (\(1-\pi\)) cases that will never fail. In the hazard
-rate, the failure rate in the numerator is conditional on the
-probability that a case is in the risk set, give survival up to time
-\(t\), and the numerator is an adjusted survivor function that accounts
-for the fraction of cured cases by time \(t\), which is \(1-\pi(t)\).
+subject with survival time \(t\) is in the risk set decreases over time,
+because an increasing number of surviving cases consist of immune or
+cured (\(1-\pi\)) cases that will never fail. In the hazard rate, the
+failure rate in the numerator is conditional on the probability that a
+case is in the risk set, given survival up to time \(t\), and the
+numerator is an adjusted survivor function that accounts for the
+fraction of cured cases by time \(t\), which is \(1-\pi(t)\).
 
@@
 
-The \pkg{spduration} package for \proglang{R} implements a
-split-population duration regression for data with time-varying
-covariates and with two options for the hazard rate shape, Weibull and
-log-logistic. The Weibull density allows for hazard rates that are
-increasing, constant, or decreasing over survival time, while the
-log-logistic density also can fit rates that have a peak at a particular
-survival time.
-
-As an example to illustrate the package functionality, we will replicate
-and extend the model of coup onsets in Belkin and Schofer (2003). Belkin
-and Schofer's paper lends itself to re-analysis with a split-populuation
-duration model because they explicitly distinguish long-term structural
-risk factors for coups from more short-term triggering causes that can
-explain the timing of a coup in a regime at risk. They argue that many
-countries never experience coups because coups are effectvely impossible
-due to structural factors, while others that never experience coups are
+In order to illustrate the package functionality, we examine a model of
+coups d`etat in Belkin and Schofer (2003). Belkin and Schofer's paper
+lends itself to re-analysis with a split-population duration model
+because they explicitly distinguish long-term structural risk factors
+for coups from more short-term triggering causes that can explain the
+timing of a coup in an at-risk regime. They argue that many countries
+never experience coups because coups are effectively impossible due to
+structural factors, while others that never experience coups are
 nevertheless at risk due to a different configuration of those same
 factors. Using language which fits nicely with the class of models
-described above, they say ``Triggers are not the source of the original
-risk, and in the absence of structural causes, the presence of
+described above, they write ``{[}t{]}riggers are not the source of the
+original risk, and in the absence of structural causes, the presence of
 triggering factors alone cannot lead to a coup. Hence, triggers should
@@
 determine the exact timing of a coup in regimes that suffer from high
-coup risk'' (p.~598 Belkin and Schofer 2003)
+coup risk'' (Belkin and Schofer 2003, 598).
 
-Belkin and Schofer develop a measure of ``structural coup risk'' which
-incorporates factors such as the strength of democratic institutions and
-civil society, and a recent history of successful coups. They employ a
-logistic regression model which does not adequately capture the process
-described in the quote above, since it implicitly assumes that all
+For their empirical test Belkin and Schofer develop a measure of
+``structural coup risk'' which incorporates factors such as the strength
+of democratic institutions and civil society, and a recent history of
+successful coups. However, they implement a logistic regression model,
+which does not capture the process as described in their quote above.
+This is because the logistic regression model assumes that all
 observations are at risk for a coup (i.e., the probability of a coup is
-non-zero for all observations). Their structural coup risk indicator is
-developed precisely to distinguish between cases where coups are
-possible and cases where they are not, so the split-population model
-allows one to examine whether the indicator effectively separates
-at-risk cases from cases where coups are impossible.
+non-zero for all observations). Since their structural coup risk
+indicator is developed precisely to distinguish between at risk and
+``immune'' cases, the split-population model allows one to examine
+whether the indicator effectively does so.
 
@@
 
+\begin{CodeInput}
+library("spduration")
+data(bscoup)
+bscoup$coup <- ifelse(bscoup$coup == "yes", 1, 0)
+bscoup      <- add_duration(bscoup, "coup", unitID = "countryid", tID = "year",
+                            freq = "year", ongoing = FALSE)
+\end{CodeInput}
+
+The \code{add\_duration} function takes as input a data frame with a
+binary response variable--``coup''--that measures the occurrence of the
+event, or failure, recorded over discrete time periods. We assume that
+the data frame is in a cross-sectional time-series format that consists
+of multiple observations over time for a number of subjects, in our case
+countries.
+
+Within the framework of duration modeling, these data conceptually
+consist of ``spells'', which are repeated observations of a unit from
+the time they enter the data (left-censoring) or after they experienced
+the event of interest, until the next event or the end of observation
+(right censoring). Table \ref{tab-ex} shows how a single country could
+have two spells during the observation period. Observations for Portugal
+make up two spells, one that begins in 1960, when our data start, to a
+coup event in 1975, when a second spell starts and continues until
+either the next coup or the data end. Canada on the other hand
+experienced no coups and so observations for that country from 1960 to
+2000 make up one spell.
+
+\begin{table}
+\begin{center}
+\begin{tabular}{lrrrrr} \toprule
+Country & Year & Coup & Spell ID & Duration \\ \midrule
+Portugal & 1960 & 0 & 1 & 1  \\ 
+\ldots & \ldots & 0 & 1 & \ldots \\
+\ldots & 1975 & 1 & 1 & 16 \\ \midrule
+Portugal & 1976 & 0 & 2 & 1 \\
+\ldots & \ldots & 0 & 2 & \ldots \\
+\ldots & 2000 & 0 & 2 & 20  \\ \midrule
+Canada & 1960 & 0 & 3 & 1 \\
+\ldots & \ldots & 0 & 3 & \ldots \\ 
+\ldots & 2000 & 0 & 3 & 41 \\ \bottomrule
+\end{tabular}
+\end{center}
+\caption{Example data frame with select duration variables.} \label{tab-ex}
+\end{table}
+
+The \code{add_duration} functions identifies spells in the data and
+returns the data frame with several additional variables needed for
+estimation. The most important among these are a counter of elapsed time
+in a spell, ``duration''. It also codes that a spell was at risk
+(``atrisk==1'') if it ended in a failure event, and not at risk
+otherwise. These are then used as response variables for the formula in
+the \code{spdur} function.
+
+Though the data used in this example are recorded annually, the function
+supports annual, monthly, or daily data and will try to convert the
+\code{tID} input to class ``Date'' given the provided argument in
+\code{freq}. Another important question is how to handle consecutive 1's
+in the response variable provided. This is controlled with the
+\code{ongoing} argument. In some cases like civil war occurrence, the
+\emph{y} variable records all years during which a country experienced a
+civil war. Here \code{ongoing} should be set to true so that the models
+try to predict the onset of the civil war, but disregard ongoing
+conflicts (\code{failure} is set to \code{NA} for these cases, dropping
+them from analysis). With ongoing set to false, successive 1's in
+\emph{y} are treated as distinct and new failures, and kept in the
+analysis. This makes sense for discrete, short-term events like coups.
+Countries can experience distinct coups in successive years.
+
+The \code{spdur} function is the primary function in the package and
+produces a regression model object of class \code{spdur} which can then
+be used with further methods. We begin by fitting first a Weibull and
+then a log-logistic split-population duration model using the coups
+data, including the measure of coup risk in the logit (risk) equation.
+
 \begin{CodeChunk}
 \begin{CodeInput}
-library("spduration")
-#data(bscoup)
-library("foreign")
-bscoup <- read.dta("data/BelkinSchoferTable4.dta")
-bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
-bscoup      <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
-                            freq="year")
+weib_model <- spdur(
+  duration ~ milreg + instab + regconf,
+  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
+  data = bscoup, silent = TRUE)
+
+loglog_model <- spdur(
+  duration ~ milreg + instab + regconf,
+  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
+  data = bscoup, distr = "loglog", silent = TRUE)
 \end{CodeInput}
-\begin{CodeOutput}
-Warning in attempt_date(data[, tID], freq): Converting to 'Date' class with
-yyyy-06-30
-\end{CodeOutput}
 \end{CodeChunk}
-
-\normalsize
-The \code{add\_duration} function takes as input a data frame with a
-binary response variable (in this example, ``coup''``) that measures the
-occurrence of the event, or failure, recorded over discrete time
-periods, and produces a data frame with a duration variable
-(named''duration``) for each spell as well as a binary variable
-(''atrisk``) that is constant across the spell and indicates whether the
-spell ended in failure. These are then used as response variables for
-the formula in the \texttt{spdur} function. The \code{spdur} function is
-the primary function in the package and produces a regression model
-object of class \code{spdur} which can then be used with further
-methods.
-
-Though the data used in this example are recorded annually, the function
-supports annual, monthly, or daily data. We begin by fitting first a
-Weibull and then a log-logistic split-population duration model using
-the coups data, including the measure of coup risk in the logit (risk)
-equation.
 
@@
 standard output showing the model formula, estimates for the duration
-and risk equations, and test statistics with \(p\)-values. One may also
-use the \code{AIC} and \code{BIC} function to calculate the information
-criterion statistics for \code{spdur} objects.
+and risk equations, and test statistics with \(p\)-values. These can be
+exported by calling \code{xtable} directly on the ``spdur'' object to
+produce Table 2.
 
@@
 \begin{CodeInput}
-AIC(weib_model)
+library("xtable")
+tbl <- xtable(loglog_model, caption = "Coup model with log-logistic hazard",
+              label = "loglog_table")
+print(tbl, caption.placement = "top", comment = FALSE, include.rownames = FALSE)
 \end{CodeInput}
-\begin{CodeOutput}
-[1] 1207.992
-\end{CodeOutput}
-\begin{CodeInput}
-AIC(loglog_model)
-\end{CodeInput}
-\begin{CodeOutput}
-[1] 1110.829
-\end{CodeOutput}
-\begin{CodeInput}
-BIC(weib_model)
-\end{CodeInput}
-\begin{CodeOutput}
-[1] 1290.452
-\end{CodeOutput}
-\begin{CodeInput}
-BIC(loglog_model)
-\end{CodeInput}
-\begin{CodeOutput}
-[1] 1193.289
-\end{CodeOutput}
 \end{CodeChunk}
-
-The objects produced by \code{spdur} can also be used with functions in
-the \code{xtable} package to produce \LaTeX~code for tables of results.
-The \code{xtable} function used on an \code{spdur} object will produce
-code such as that used to create Table \ref{loglog_table}.
 
@@
   \hline
-(Dur. Intercept) & 3.30 & 0.14 & 0.00 \\ 
-  milreg & -1.53 & 0.15 & 0.00 \\ 
-  instab & -0.22 & 0.03 & 0.00 \\ 
-  regconf & 3.04 & 2.22 & 0.17 \\ 
-  log(alpha) & -0.52 & 0.07 & 0.00 \\ 
-  (Risk Intercept) & -14.61 & 9.02 & 0.11 \\ 
-  couprisk & 9.11 & 4.37 & 0.04 \\ 
-  wealth & 3.51 & 1.75 & 0.05 \\ 
-  milreg.1 & 21.76 & 0.24 & 0.00 \\ 
-  rwar & -6.37 & 5.38 & 0.24 \\ 
-  regconf.1 & -20.22 & 54.04 & 0.71 \\ 
-  samerica & -3.31 & 3.28 & 0.31 \\ 
-  camerica & -6.46 & 2.57 & 0.01 \\ 
+(Dur. Intercept) & 2.40 & 0.21 & 0.00 \\ 
+  milreg & -1.13 & 0.21 & 0.00 \\ 
+  instab & -0.09 & 0.02 & 0.00 \\ 
+  regconf & -2.52 & 2.16 & 0.24 \\ 
+  log(alpha) & -0.45 & 0.06 & 0.00 \\ 
+  (Risk Intercept) & 2.93 & 1.87 & 0.12 \\ 
+  couprisk & 0.59 & 0.32 & 0.07 \\ 
+  wealth & -0.36 & 0.28 & 0.20 \\ 
+  milreg.1 & 10.82 & 9.31 & 0.25 \\ 
+  rwar & -0.52 & 0.94 & 0.58 \\ 
+  regconf.1 & -5.43 & 5.62 & 0.33 \\ 
+  samerica & 2.09 & 1.45 & 0.15 \\ 
+  camerica & -0.39 & 0.73 & 0.59 \\ 
    \hline
 \end{tabular}
+\caption{Coup model with log-logistic hazard.} \label{loglog_table}
 \end{table}
 
-The table shows estimates from the duration equation, beginning with the
-intercept, and then estimates from the risk equation. Interestingly, the
-coup risk variable has a rather large and statistically significant
-coefficient in the risk equation.
+Table \ref{loglog_table} shows estimates from the duration equation,
+beginning with the intercept, and then estimates from the risk equation.
+The duration component of the model is in accelerated failure time
+format and the coefficient estimates are on the log of expected time to
+failure. The negative coefficient for military regimes for example means
+that the expected time to a coup is shorter in military regimes than
+non-military regimes, holding all other factors constant. In the risk
+equation, positive coefficients mean higher risk of coup. Thus, military
+regimes have a higher risk of experiencing a coup.
+
+One may also use the \code{AIC} and \code{BIC} function to calculate the
+information criterion statistics for \code{spdur} objects. In our
+example, both models are close enough in both statistics to be
+indistinguishable, so we will continue to focus on the log-logistic
+form.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+matrix(c(
+  AIC(weib_model), AIC(loglog_model), BIC(weib_model), BIC(loglog_model)
+  ), ncol = 2, dimnames = list(c("Weibull", "Loglog"), c("AIC", "BIC")))
+\end{CodeInput}
+\begin{CodeOutput}
+             AIC      BIC
+Weibull 1329.908 1348.972
+Loglog  1331.214 1350.278
+\end{CodeOutput}
+\end{CodeChunk}
 
@@
 \begin{CodeInput}
-plot(loglog_model, type="hazard", main = "A")
+plot_hazard(loglog_model, main = "A")
 \end{CodeInput}
@@
 
+As mean values are maybe not that interesting, we can also generate
+scenarios by entering values manually. The values are used in the same
+order that variables are specified in the equations used to estimate the
+models, which also corresponds to the order of variables in Table 2.
+Note the inclusion of an intercept term. This produces plot B in Figure
+\ref{hazard-ex}
+
 \begin{CodeChunk}
 \begin{CodeInput}
-plot(loglog_model, type="hazard", 
+plot_hazard(loglog_model, 
      xvals = c(1, 1, 10, 0.05), 
@@
 
-Importantly, the package also includes functions to evaluate model
-predictions. The generic function \code{predict} can be used on an
-object of class \code{spdur} to generate several kinds of predictions,
-including the probability that an observation is ``at-risk'' and the
-probability of failure for a given time period. In addition, the generic
-\code{plot} function can be used on \code{spdur} objects to produce a
-separation plot (Greenhill, Ward, and Sacks 2011), which is a graphical
-display for evaluating model predictions. The code below produces Figure
-\ref{insamp}.
+The hazard plots can also be called with
+\code{plot(..., type = "hazard")}. Without a type argument,
+\code{plot.spdur} will produce a separation plot, which is a graphical
+display for evaluating model predictions (Greenhill, Ward, and Sacks
+2011). The code below produces Figure \ref{insamp}.
 
@@
 
+Underlying both plotting functions is the \code{predict} function, which
+can be used on an object of class \code{spdur} to generate several kinds
+of predictions, including the probability that an observation is
+``at-risk'' and the probability of failure for a given time period.
+
+\subsection{Out-of-sample testing}\label{out-of-sample-testing}
+
 Finally, we demonstrate how to evaluate a model's out of sample
-predictions. We begin by splitting the data into training and test sets,
-using the last year of data (1999) as the test set and omitting
-incomplete observations from the test set.
+predictions. We will use data from 1996 onwards as the test set, and
+prior data for training purposes. The \code{add_duration} function
+retrospectively codes the risk variable based on how a spell ended, and
+we therefore need to take care in how we add the duration variables for
+each data set. For the training data we need to subset the training set
+first, so that coups in the tests set don't influence the risk coding in
+the training data.
 
@@
 data(bscoup)
-bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
+bscoup$coup <- ifelse(bscoup$coup == "yes", 1, 0)
 coup_train <- bscoup[bscoup$year < 1996, ]
-coup_train <- add_duration(coup_train, "coup", unitID="countryid", tID="year",
-                          freq="year")
+coup_train <- add_duration(coup_train, "coup", unitID = "countryid", tID = "year",
+                          freq = "year", ongoing = FALSE)
 \end{CodeInput}
-\begin{CodeOutput}
-Warning in attempt_date(data[, tID], freq): Converting to 'Date' class with
-yyyy-06-30
-\end{CodeOutput}
 \end{CodeChunk}
+
+For the test set it is recommended to add the duration variables and \textit{then} subset the test set.
+Since the test set is later in time than the training set we do not have to worry about contamination of the risk coding, but if we subset the data before building the duration variables we will start all duration counters at 1996, when in fact we can safely use the previous historic coup information.
+To do this we need to build the duration variables first:
 
@@
 \begin{CodeInput}
-coup_test  <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
-                           freq="year")
-\end{CodeInput}
-\begin{CodeOutput}
-Warning in attempt_date(data[, tID], freq): Converting to 'Date' class with
-yyyy-06-30
-\end{CodeOutput}
-\begin{CodeInput}
+coup_test  <- add_duration(bscoup, "coup", unitID = "countryid", tID = "year",
+                           freq = "year", ongoing = FALSE)
 coup_test  <- coup_test[coup_test$year >= 1996, ]
@@
 
-\normalsize
-The \code{spdurCrisp} function can then be used to fit a model on the
-training set and generate predictions for the test set.
+Now we can fit new models using the training data, and calculate
+predictions from these models for the test data, using
+\code{predict(..., newdata = coup_test)}.
 
@@
   atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
-  data = coup_train, distr="loglog", silent = TRUE) 
+  data = coup_train, distr = "loglog", silent = TRUE) 
 \end{CodeInput}
 \end{CodeChunk}
-
-Foo
 
@@
 
-\normalsize
-The \code{train},\code{test}, and \code{pred} arguments are used to
-specify data frames for the training set, test set, and forecasting set.
-We do not wish to generate forecasts here, so we simply specify the same
-data frame for the test and forecasting set. Here we use
-\code{spdurCrisp} to calculate the conditional risk, which is the
-default calculation for this function and is the probability that an
-observation is in the at-risk population. The test set predictions are
-stored in the model object as \code{test.p}, and we extract them to
-produce the plot shown below in Figure \ref{outsamp}, which was created
-using the \code{separationplot} package.
+Since we are predicting for data that is not contained in the ``spdur''
+model objects, we have to use the \code{separationplot} functions
+directly from the package to produce Figure \ref{oos-sepplots}.
 
-\textbackslash{}begin\{CodeChunk\} \textbackslash{}begin\{CodeInput\}
-library(``separationplot'')
+\begin{CodeChunk}
+\begin{CodeInput}
+library("separationplot")
+obs_y <- coup_test[complete.cases(coup_test), "coup"]
 
-obs\_y \textless{}- coup\_test\$coup{[}complete.cases(coup\_test){]}
-
-par(mfrow=c(2,1),mar=c(2,2,2,2)) separationplot(weib2\_test\_p, obs\_y)
-separationplot(loglog2\_test\_p, obs\_y)
-\textbackslash{}end\{CodeInput\} \textbackslash{}end\{CodeChunk\}
+par(mfrow=c(2,1),mar=c(2,2,2,2))
+separationplot(weib2_test_p,   obs_y, newplot = FALSE)
+separationplot(loglog2_test_p, obs_y, newplot = FALSE)
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 \includegraphics[width=0.8\textwidth]{graphics/oos-sepplots.pdf}
+\caption{Out-of-sample separation plots.}
+\label{oos-sepplots}
 \end{center}
 \end{figure}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-glm_model2    <- glm(
-  coup ~ milreg + instab + regconf + couprisk + wealth + rwar + regconf + 
-    samerica + camerica,
-  data = coup_train, family = "binomial"
-)
-
-c(Weibull = AIC(weib_model2), 
-  Loglog = AIC(loglog_model2), 
-  Logistic = AIC(glm_model2))
-\end{CodeInput}
-\begin{CodeOutput}
-  Weibull    Loglog  Logistic 
- 990.5818  985.2633 1171.3556 
-\end{CodeOutput}
-\end{CodeChunk}
 
@@
 Truncation and censoring are problematic for split-population duration
-models as they are for standard duration regression, but also pose some
+models, as they are for standard duration regression, but also pose some
 additional considerations. In left-truncation, we do not observe data
@@
 
-Furthermore, in out-of-sample testing based on some kind of data
-partitioning scheme, this coding scheme can lead to unintentioned
-contamination of in-sample cases with knowledge of out-of-sample
-failures through the risk coding for failed spells. This leads to two
-recommendations. First, the duration data (\texttt{add\_duration})
-should not be built until after partitioning the data. Secondly, and as
-a result of the first point, data should be partitioned in a scheme that
-samples either by spell or block-wise, e.g.~by witholding the last
-\textbf{x} years of data, in order to be able to proberly build the
-duration variables.
+Furthermore, as the example above shows, in out-of-sample testing based
+on some kind of data partitioning scheme, this coding scheme can lead to
+unintentional contamination of in-sample cases with knowledge of
+out-of-sample failures through the risk coding for failed spells. This
+leads to two recommendations. First, data should be partitioned spell or
+block-wise, e.g., by withholding the last \emph{x} years of data, and
+not randomly. Second, given the two concerns of left-censoring and
+non-independence induced through the duration and risk coding, care
+should be taken to ensure that duration data are built without access to
+future information in another data partition.
 
@@
 
-Many outcomes in the area that we work in, political science and
-political violence more specifically, are rare events. Coups, the onset
-of war and civil war, mass killings, and so on are exceedingly rare
-events when one considers that most countries most of the time are
-stable. Figure \ref{rates} shows a few examples for published research
-that models binary outcomes.
+Many outcomes of interest are rare events. For example, coups, war
+onset, or mass killings are exceedingly rare events when considering
+that most countries do not experience high levels of violence. Figure
+\ref{rates} shows a few examples of published research that models
+binary outcomes. Fearon and Laitin (2003) is a widely cited study of
+civil war onset that uses yearly observations of all major countries;
+Beger, Dorff, and Ward (2014) is an example of the positive rates when
+moving to monthly data for a similar set of countries. The positive
+rates range from 5 to less than 0.2\% of all data points.\footnote{For a
+  discussion of the difficulties rare events can pose for prediction see
+  King and Zeng (2001b) and King and Zeng (2001a).}
 
 In the language of machine learning, we are dealing with highly
-imbalanced classes, although our data are also highly structured with
-clear dependence over space and time in countries and societies that may
-experience political violence. This is a well-recognized problem and has
-led to the development or use of several specialized mixture models like
+imbalanced classes. It is a well-recognized problem and has led to the
+development or use of several specialized mixture models like
 zero-inflated Poisson and negative binomial regression for count data,
@@
 other solutions to the sparse outcome problem, also addresses underlying
-temporal dynamics.
+temporal dynamics that are an important part of the non-independent data
+political scientists and other social scientists commonly use.
 
-Split-population duration models are not only appealing in this
-technical sense, but they also match the logic or intuituion many
-analysts of politics use when they distinguish long-term risk factors
-from more fleeting triggering causes. The example we have used, Belkin
-and Schofer (2003), is a particularly clear illustration of how well the
-language of theorists maps onto the model intuition.
+Split-population duration models are not only appealing in a technical
+sense, but they also match the logic or intuition many social scientists
+use when they distinguish long-term risk factors from more fleeting
+triggering causes. The example we have used, Belkin and Schofer (2003),
+is a particularly clear illustration of how well the language of
+theorists maps onto the model intuition.
 
@@
 
-Bagozzi, Benjamin E, Daniel W Hill, Will H Moore, and Bumba Mukherjee.
-2015. ``Modeling Two Types of Peace the Zero-Inflated Ordered Probit
-(ZiOP) Model in Conflict Research.'' \emph{Journal of Conflict
-Resolution} 59 (4). SAGE Publications: 728--52.
+\hypertarget{refs}{}
+\hypertarget{ref-bagozzi2015modeling}{}
+Bagozzi, Benjamin E., Daniel W. Hill, Will H. Moore, and Bumba
+Mukherjee. 2015. ``Modeling Two Types of Peace the Zero-Inflated Ordered
+Probit (ZiOP) Model in Conflict Research.'' \emph{Journal of Conflict
+Resolution} 59 (4): 728--52.
 
+\hypertarget{ref-beger2014ensemble}{}
 Beger, Andreas, Cassy L. Dorff, and Michael D. Ward. 2014. ``Ensemble
 Forecasting of Irregular Leadership Change.'' \emph{Research \&
-Politics} 1 (3).
+Politics} 1 (3): 1--7.
 
+\hypertarget{ref-belkin2003toward}{}
 Belkin, Aaron, and Evan Schofer. 2003. ``Toward a Structural
 Understanding of Coup Risk.'' \emph{Journal of Conflict Resolution} 47
-(5). Sage Publications: 594--620.
+(5): 594--620.
 
-Berkson, Joseph, and Robert P Gage. 1952. ``Survival Curve for Cancer
+\hypertarget{ref-berkson1952survival}{}
+Berkson, Joseph, and Robert P. Gage. 1952. ``Survival Curve for Cancer
 Patients Following Treatment.'' \emph{Journal of the American
-Statistical Association} 47 (259). Taylor \& Francis Group: 501--15.
+Statistical Association} 47 (259): 501--15.
 
+\hypertarget{ref-boag1949maximum}{}
 Boag, John W. 1949. ``Maximum Likelihood Estimates of the Proportion of
 Patients Cured by Cancer Therapy.'' \emph{Journal of the Royal
-Statistical Society. Series B (Methodological)} 11 (1). JSTOR: 15--53.
+Statistical Society B} 11 (1): 15--53.
 
+\hypertarget{ref-deyoung2003failure}{}
 DeYoung, Robert. 2003. ``The Failure of New Entrants in Commercial
 Banking Markets: A Split-Population Duration Analysis.'' \emph{Review of
-Financial Economics} 12 (1). Elsevier: 7--33.
+Financial Economics} 12 (1): 7--33.
 
+\hypertarget{ref-douglas1994hazard}{}
 Douglas, Stratford, and Govind Hariharan. 1994. ``The Hazard of Starting
 Smoking: Estimates from a Split Population Duration Model.''
-\emph{Journal of Health Economics} 13 (2). Elsevier: 213--30.
+\emph{Journal of Health Economics} 13 (2): 213--30.
 
-Forster, Martin, and Andrew M Jones. 2001. ``The Role of Tobacco Taxes
+\hypertarget{ref-fearon2003ethnicity}{}
+Fearon, James D., and David D. Laitin. 2003. ``Ethnicity, Insurgency,
+and Civil War.'' \emph{American Political Science Review} 97 (1):
+75--90.
+
+\hypertarget{ref-forster2001role}{}
+Forster, Martin, and Andrew M. Jones. 2001. ``The Role of Tobacco Taxes
 in Starting and Quitting Smoking: Duration Analysis of British Data.''
-\emph{Journal of the Royal Statistical Society: Series A (Statistics in
-Society)} 164 (3). Wiley Online Library: 517--47.
+\emph{Journal of the Royal Statistical Society A} 164 (3): 517--47.
 
-Greenhill, Brian, Michael D Ward, and Audrey Sacks. 2011. ``The
+\hypertarget{ref-greenhill2011separation}{}
+Greenhill, Brian, Michael D. Ward, and Audrey Sacks. 2011. ``The
 Separation Plot: A New Visual Method for Evaluating the Fit of Binary
-Models.'' \emph{American Journal of Political Science} 55 (4). Wiley
-Online Library: 991--1002.
+Models.'' \emph{American Journal of Political Science} 55 (4):
+991--1002.
 
+\hypertarget{ref-king2001explaining}{}
+King, Gary, and Langche Zeng. 2001a. ``Explaining Rare Events in
+International Relations.'' \emph{International Organization} 55 (03):
+693--715.
+
+\hypertarget{ref-king2001logistic}{}
+---------. 2001b. ``Logistic Regression in Rare Events Data.''
+\emph{Political Analysis} 9 (2): 137--63.
+
+\hypertarget{ref-schmidt1989predicting}{}
 Schmidt, Peter, and Ann Dryden Witte. 1989. ``Predicting Criminal
 Recidivism Using Split Population Survival Time Models.'' \emph{Journal
-of Econometrics} 40 (1). Elsevier: 141--59.
+of Econometrics} 40 (1): 141--59.
 
+\hypertarget{ref-svolik2008authoritarian}{}
 Svolik, Milan. 2008. ``Authoritarian Reversals and Democratic
-Consolidation.'' \emph{American Political Science Review} 102 (02).
-Cambridge Univ Press: 153--68.
+Consolidation.'' \emph{American Political Science Review} 102 (2):
+153--68.
 
-Ward, Michael D, Nils W Metternich, Cassy L Dorff, Max Gallop, Florian M
-Hollenbach, Anna Schultz, and Simon Weschle. 2013. ``Learning from the
-Past and Stepping into the Future: Toward a New Generation of Conflict
-Prediction.'' \emph{International Studies Review} 15 (4). Wiley Online
-Library: 473--90.
+\hypertarget{ref-ward2013learning}{}
+Ward, Michael D., Nils W. Metternich, Cassy L. Dorff, Max Gallop,
+Florian M. Hollenbach, Anna Schultz, and Simon Weschle. 2013. ``Learning
+from the Past and Stepping into the Future: Toward a New Generation of
+Conflict Prediction.'' \emph{International Studies Review} 15 (4):
+473--90.
 

### 2016-03-07 manuscript2.tex
how about actually syncing manuscript2.tex
also expanded alternative implementation section

--- 2016-03-07 manuscript.tex
@@
 \documentclass[article]{jss}
-\usepackage[utf8]{inputenc}
 
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+%% almost as usual
 \author{
 Andreas Beger\\Duke University \And Daniel W.
-Hill, Jr.\\University of Georgia \AND Nils W.
-Metternich\\University College London \And Shahryar Minhas\\Duke University \And Michael D.
+Hill, Jr.\\University of Georgia \And Nils W.
+Metternich\\University College London \AND Shahryar Minhas\\Duke University \And Michael D.
 Ward\\Duke University
@@
 Regression Package}
-\Keywords{regression, rare events, \proglang{R}}
 
+%% for pretty printing and a nice hypersummary also set:
+\Plainauthor{Andreas Beger, Daniel W. Hill, Jr., Nils W. Metternich, Shahryar Minhas, Michael D. Ward}
+\Plaintitle{Splitting it up: the spduration split-population duration regression
+package}
+\Shorttitle{\pkg{spduration}: Split-population duration regression}
+
+%% an abstract and keywords
 \Abstract{
 We present an implementation of split-population duration regression in
-the \pkg{spduration} R package and an application to data on military
+the \pkg{spduration} \proglang{R} package and an application to data on military
 coups. The statistical model accounts for units that are immune to a
@@
 
-\Plainauthor{Andreas Beger, Daniel W. Hill, Jr., Nils W. Metternich, Shahryar Minhas, Michael D. Ward}
-\Plaintitle{Splitting it up: the spduration split-population duration regression
-package}
-\Shorttitle{\pkg{spduration}: Split-population duration regression}
-\Plainkeywords{regression, rare events, R}
+\Keywords{regression, rare events, \proglang{R}}
+\Plainkeywords{regression, rare events, R} %% without formatting
+%% at least one keyword must be supplied
 
 %% publication information
+%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
 %% \Volume{50}
@@
 %% \Year{2012}
-\Submitdate{}
+%% \Submitdate{2012-06-04}
 %% \Acceptdate{2012-06-04}
 
+%% The address of (at least) one author should be given
+%% in the following format:
 \Address{
@@
 
-\usepackage{amsmath} \usepackage{booktabs}
+%% It is also possible to add a telephone and fax number
+%% before the e-mail in the following format:
+%% Telephone: +43/512/507-7103
+%% Fax: +43/512/507-2851
+
+%% for those who use Sweave please include the following line (with % symbols):
+%% need no \usepackage{Sweave.sty}
+
+%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\usepackage{amsmath} 
+\usepackage{booktabs}
 
@@
 
-\section{Introduction}\label{introduction}
+\section{Introduction}
 
@@
 \section{Immune populations and inference of duration
-processes}\label{immune-populations-and-inference-of-duration-processes}
+processes}
 
@@
 The insight that populations might be split in regard to their baseline
-risk has been formulated as early as 1949 by Boag (1949) and Berkson and
-Gage (1952), who were researching survival rates in cancer patients. In
+risk has been formulated as early as 1949 by \citet{boag1949maximum} and \citet{berkson1952survival}, who were researching survival rates in cancer patients. In
 their application some fraction of patients survived because their
@@
 medicine. In an early and foundational application that reached beyond
-medicine, Schmidt and Witte (1989) examined criminal recidivism using
+medicine, \citet{schmidt1989predicting} examined criminal recidivism using
 data on close to 10,000 prisoners from the North Carolina prison system
@@
 
-In a complementary set of research from public health, Douglas and
-Hariharan (1994) use data from the United States to model the age at
-which individuals began to smoke, and Forster and Jones (2001) examine
-the impact of tobacco taxes on smoking and quitting decisions. DeYoung
-(2003), an economist, models the failure of new commercial banks in the
-US during the 1980's. Svolik (2008) uses a split-population duration
+In a complementary set of research from public health, \citet{douglas1994hazard} use data from the United States to model the age at
+which individuals began to smoke, and \citet{forster2001role} examine
+the impact of tobacco taxes on smoking and quitting decisions. \citet{deyoung2003failure}, an economist, models the failure of new commercial banks in the
+US during the 1980's. \citet{svolik2008authoritarian} uses a split-population duration
 framework to examine whether democratic regimes persist or revert to
@@
 regular predictions for five different forms of political conflict for
-the Integrated Crisis and Early Warning System (ICEWS) project (Ward et
-al. 2013) and to model irregular leadership changes for the Political
-Instability Task Force (PITF; Beger, Dorff, and Ward 2014).
+the Integrated Crisis and Early Warning System (ICEWS) project \citep{ward2013learning} and to model irregular leadership changes for the Political
+Instability Task Force \citep[PITF;][]{beger2014ensemble}.
 
-\section{Model development}\label{model-development}
+\section{Model development}
 
@@
 \section{Fit a split-population model on coups
-data}\label{fit-a-split-population-model-on-coups-data}
+data}
 
 In order to illustrate the package functionality, we examine a model of
-coups d`etat in Belkin and Schofer (2003). Belkin and Schofer's paper
+coups d`etat in \citet{belkin2003toward}. Belkin and Schofer's paper
 lends itself to re-analysis with a split-population duration model
@@
 
+\begin{CodeChunk}
 \begin{CodeInput}
-library("spduration")
-data(bscoup)
-bscoup$coup <- ifelse(bscoup$coup == "yes", 1, 0)
-bscoup      <- add_duration(bscoup, "coup", unitID = "countryid", tID = "year",
-                            freq = "year", ongoing = FALSE)
+R> library("spduration")
+R> data(bscoup)
+R> bscoup$coup <- ifelse(bscoup$coup == "yes", 1, 0)
+R> bscoup <- add_duration(bscoup, "coup", unitID = "countryid", 
++    tID = "year", freq = "year", ongoing = FALSE)
 \end{CodeInput}
+\end{CodeChunk}
 
@@
 \begin{CodeInput}
-weib_model <- spdur(
-  duration ~ milreg + instab + regconf,
-  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
-  data = bscoup, silent = TRUE)
+R> weib_model <- spdur(
++    duration ~ milreg + instab + regconf,
++    atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + 
++      camerica,
++    data = bscoup, silent = TRUE)
 
-loglog_model <- spdur(
-  duration ~ milreg + instab + regconf,
-  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
-  data = bscoup, distr = "loglog", silent = TRUE)
+R> loglog_model <- spdur(
++    duration ~ milreg + instab + regconf,
++    atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + 
++      camerica,
++    data = bscoup, distr = "loglog", silent = TRUE)
 \end{CodeInput}
@@
 \begin{CodeInput}
-library("xtable")
-tbl <- xtable(loglog_model, caption = "Coup model with log-logistic hazard",
-              label = "loglog_table")
-print(tbl, caption.placement = "top", comment = FALSE, include.rownames = FALSE)
+R> library("xtable")
+R> tbl <- xtable(loglog_model, caption = "Coup model with log-logistic hazard",
++    label = "loglog_table")
+R> print(tbl, caption.placement = "top", comment = FALSE, 
++    include.rownames = FALSE)
 \end{CodeInput}
@@
 (Dur. Intercept) & 2.40 & 0.21 & 0.00 \\ 
-  milreg & -1.13 & 0.21 & 0.00 \\ 
-  instab & -0.09 & 0.02 & 0.00 \\ 
-  regconf & -2.52 & 2.16 & 0.24 \\ 
-  log(alpha) & -0.45 & 0.06 & 0.00 \\ 
+  milreg & $-$1.13 & 0.21 & 0.00 \\ 
+  instab & $-$0.09 & 0.02 & 0.00 \\ 
+  regconf & $-$2.52 & 2.16 & 0.24 \\ 
+  log(alpha) & $-$0.45 & 0.06 & 0.00 \\ 
   (Risk Intercept) & 2.93 & 1.87 & 0.12 \\ 
   couprisk & 0.59 & 0.32 & 0.07 \\ 
-  wealth & -0.36 & 0.28 & 0.20 \\ 
+  wealth & $-$0.36 & 0.28 & 0.20 \\ 
   milreg.1 & 10.82 & 9.31 & 0.25 \\ 
-  rwar & -0.52 & 0.94 & 0.58 \\ 
-  regconf.1 & -5.43 & 5.62 & 0.33 \\ 
+  rwar & $-$0.52 & 0.94 & 0.58 \\ 
+  regconf.1 & $-$5.43 & 5.62 & 0.33 \\ 
   samerica & 2.09 & 1.45 & 0.15 \\ 
-  camerica & -0.39 & 0.73 & 0.59 \\ 
+  camerica & $-$0.39 & 0.73 & 0.59 \\ 
    \hline
@@
 
-Table \ref{loglog_table} shows estimates from the duration equation,
+Table~\ref{loglog_table} shows estimates from the duration equation,
 beginning with the intercept, and then estimates from the risk equation.
@@
 \begin{CodeInput}
-matrix(c(
-  AIC(weib_model), AIC(loglog_model), BIC(weib_model), BIC(loglog_model)
-  ), ncol = 2, dimnames = list(c("Weibull", "Loglog"), c("AIC", "BIC")))
+R> matrix(c(
++    AIC(weib_model), AIC(loglog_model), BIC(weib_model), BIC(loglog_model)
++    ), ncol = 2, dimnames = list(c("Weibull", "Loglog"), c("AIC", "BIC")))
 \end{CodeInput}
@@
 \begin{CodeInput}
-plot_hazard(loglog_model, main = "A")
+R> plot_hazard(loglog_model, main = "A")
 \end{CodeInput}
@@
 \begin{CodeInput}
-plot_hazard(loglog_model, 
-     xvals = c(1, 1, 10, 0.05), 
-     zvals = c(1, 7, 8.64, 1, 1, 0.05, 0, 0),
-     main = "B")
+R> plot_hazard(loglog_model, xvals = c(1, 1, 10, 0.05), 
++    zvals = c(1, 7, 8.64, 1, 1, 0.05, 0, 0), main = "B")
 \end{CodeInput}
@@
 \code{plot.spdur} will produce a separation plot, which is a graphical
-display for evaluating model predictions (Greenhill, Ward, and Sacks
-2011). The code below produces Figure \ref{insamp}.
+display for evaluating model predictions \citep{greenhill2011separation}. The code below produces Figure \ref{insamp}.
 
@@
 \begin{CodeInput}
-plot(weib_model)
-plot(loglog_model)
+R> plot(weib_model)
+R> plot(loglog_model)
 \end{CodeInput}
@@
 
-\subsection{Out-of-sample testing}\label{out-of-sample-testing}
+\subsection{Out-of-sample testing}
 
@@
 \begin{CodeInput}
-data(bscoup)
-bscoup$coup <- ifelse(bscoup$coup == "yes", 1, 0)
-coup_train <- bscoup[bscoup$year < 1996, ]
-coup_train <- add_duration(coup_train, "coup", unitID = "countryid", tID = "year",
-                          freq = "year", ongoing = FALSE)
+R> data(bscoup)
+R> bscoup$coup <- ifelse(bscoup$coup == "yes", 1, 0)
+R> coup_train <- bscoup[bscoup$year < 1996, ]
+R> coup_train <- add_duration(coup_train, "coup", unitID = "countryid", 
++    tID = "year", freq = "year", ongoing = FALSE)
 \end{CodeInput}
@@
 \begin{CodeInput}
-coup_test  <- add_duration(bscoup, "coup", unitID = "countryid", tID = "year",
-                           freq = "year", ongoing = FALSE)
-coup_test  <- coup_test[coup_test$year >= 1996, ]
+R> coup_test  <- add_duration(bscoup, "coup", unitID = "countryid", 
++    tID = "year", freq = "year", ongoing = FALSE)
+R> coup_test  <- coup_test[coup_test$year >= 1996, ]
 \end{CodeInput}
@@
 \begin{CodeInput}
-weib_model2   <- spdur(
-  duration ~ milreg + instab + regconf,
-  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
-  data = coup_train, silent = TRUE)
+R> weib_model2   <- spdur(
++    duration ~ milreg + instab + regconf,
++    atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + 
++      camerica,
++    data = coup_train, silent = TRUE)
 
-loglog_model2 <- spdur(
-  duration ~ milreg + instab + regconf,
-  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
-  data = coup_train, distr = "loglog", silent = TRUE) 
+R> loglog_model2 <- spdur(
++    duration ~ milreg + instab + regconf,
++    atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + 
++      camerica,
++    data = coup_train, distr = "loglog", silent = TRUE) 
 \end{CodeInput}
@@
 \begin{CodeInput}
-weib2_test_p   <- predict(weib_model2, newdata = coup_test)
-loglog2_test_p <- predict(loglog_model2, newdata = coup_test)
+R> weib2_test_p   <- predict(weib_model2, newdata = coup_test)
+R> loglog2_test_p <- predict(loglog_model2, newdata = coup_test)
 \end{CodeInput}
@@
 \begin{CodeInput}
-library("separationplot")
-obs_y <- coup_test[complete.cases(coup_test), "coup"]
+R> library("separationplot")
+R> obs_y <- coup_test[complete.cases(coup_test), "coup"]
 
-par(mfrow=c(2,1),mar=c(2,2,2,2))
-separationplot(weib2_test_p,   obs_y, newplot = FALSE)
-separationplot(loglog2_test_p, obs_y, newplot = FALSE)
+R> par(mfrow=c(2,1),mar=c(2,2,2,2))
+R> separationplot(weib2_test_p,   obs_y, newplot = FALSE)
+R> separationplot(loglog2_test_p, obs_y, newplot = FALSE)
 \end{CodeInput}
@@
 
-\section{Conclusion}\label{conclusion}
+\section{Alternative implementations}
+
+The \pkg{survival} library, arguable the most well-established library for survival and event history analysis in \proglang{R}, allows for the estimation of a wide variet of semi-, non-, and parametric survival models, and provides facilities for handling and descriptive summaries of survival and event history data.
+It does not include cure rate or split-population mixture models of the type we have implemented though.
+
+One interesting extension of our project in the future would be to integrate and adapt the \code{Surv} class for survival data in \pkg{survival} to the \pkg{spduration} library.
+This would require some modification of the class as our model also requires information on the risk status of spells, but would give access to the much broader functionality, especially for descriptive summaries of data, in the \pkg{survival} library.
+
+Otherwise, the \pkg{smcure} package \citep{cai2012smcureR}, described in \citet{cai2012smcure}, appears to be the only alternative split-population/cure rate duration model implementation in \proglang{R}.
+It implements semi-parametric proportional hazards and accelerated failure time cure models.
+The semi-paramatric estimation of the hazard function shape obviates the need to make an assumption about its shape, which can be difficult in practice to empirically evalute, as the \pkg{smcure} discuss.
+However, it is much more limited in practical application as it only accomodates case-based data, not time-varying covariates.
+
+Furthermore, \proglang{Stata} implements cure models similar to our own, and which can handle time-varying covariates, but this is of course of limited appeal to \proglang{R} users.
+
+\section{Conclusion}
 
@@
 \ref{rates} shows a few examples of published research that models
-binary outcomes. Fearon and Laitin (2003) is a widely cited study of
+binary outcomes. \citet{fearon2003ethnicity} is a widely cited study of
 civil war onset that uses yearly observations of all major countries;
-Beger, Dorff, and Ward (2014) is an example of the positive rates when
+\citet{beger2014ensemble} is an example of the positive rates when
 moving to monthly data for a similar set of countries. The positive
@@
   discussion of the difficulties rare events can pose for prediction see
-  King and Zeng (2001b) and King and Zeng (2001a).}
+  \citet{king2001explaining} and \citet{king2001logistic}.}
 
@@
 zero-inflated Poisson and negative binomial regression for count data,
-and a zero-inflated ordered probit for ordinal outcomes (Bagozzi et al.
-2015). Split-population duration regression provides another principled
+and a zero-inflated ordered probit for ordinal outcomes \citep{bagozzi2015modeling}. Split-population duration regression provides another principled
 solution to the challenges posed by data in this domain, but, unlike
@@
 use when they distinguish long-term risk factors from more fleeting
-triggering causes. The example we have used, Belkin and Schofer (2003),
+triggering causes. The example we have used, \citet{belkin2003toward},
 is a particularly clear illustration of how well the language of
@@
 
-\section*{References}\label{references}
-\addcontentsline{toc}{section}{References}
-
-\hypertarget{refs}{}
-\hypertarget{ref-bagozzi2015modeling}{}
-Bagozzi, Benjamin E., Daniel W. Hill, Will H. Moore, and Bumba
-Mukherjee. 2015. ``Modeling Two Types of Peace the Zero-Inflated Ordered
-Probit (ZiOP) Model in Conflict Research.'' \emph{Journal of Conflict
-Resolution} 59 (4): 728--52.
-
-\hypertarget{ref-beger2014ensemble}{}
-Beger, Andreas, Cassy L. Dorff, and Michael D. Ward. 2014. ``Ensemble
-Forecasting of Irregular Leadership Change.'' \emph{Research \&
-Politics} 1 (3): 1--7.
-
-\hypertarget{ref-belkin2003toward}{}
-Belkin, Aaron, and Evan Schofer. 2003. ``Toward a Structural
-Understanding of Coup Risk.'' \emph{Journal of Conflict Resolution} 47
-(5): 594--620.
-
-\hypertarget{ref-berkson1952survival}{}
-Berkson, Joseph, and Robert P. Gage. 1952. ``Survival Curve for Cancer
-Patients Following Treatment.'' \emph{Journal of the American
-Statistical Association} 47 (259): 501--15.
-
-\hypertarget{ref-boag1949maximum}{}
-Boag, John W. 1949. ``Maximum Likelihood Estimates of the Proportion of
-Patients Cured by Cancer Therapy.'' \emph{Journal of the Royal
-Statistical Society B} 11 (1): 15--53.
-
-\hypertarget{ref-deyoung2003failure}{}
-DeYoung, Robert. 2003. ``The Failure of New Entrants in Commercial
-Banking Markets: A Split-Population Duration Analysis.'' \emph{Review of
-Financial Economics} 12 (1): 7--33.
-
-\hypertarget{ref-douglas1994hazard}{}
-Douglas, Stratford, and Govind Hariharan. 1994. ``The Hazard of Starting
-Smoking: Estimates from a Split Population Duration Model.''
-\emph{Journal of Health Economics} 13 (2): 213--30.
-
-\hypertarget{ref-fearon2003ethnicity}{}
-Fearon, James D., and David D. Laitin. 2003. ``Ethnicity, Insurgency,
-and Civil War.'' \emph{American Political Science Review} 97 (1):
-75--90.
-
-\hypertarget{ref-forster2001role}{}
-Forster, Martin, and Andrew M. Jones. 2001. ``The Role of Tobacco Taxes
-in Starting and Quitting Smoking: Duration Analysis of British Data.''
-\emph{Journal of the Royal Statistical Society A} 164 (3): 517--47.
-
-\hypertarget{ref-greenhill2011separation}{}
-Greenhill, Brian, Michael D. Ward, and Audrey Sacks. 2011. ``The
-Separation Plot: A New Visual Method for Evaluating the Fit of Binary
-Models.'' \emph{American Journal of Political Science} 55 (4):
-991--1002.
-
-\hypertarget{ref-king2001explaining}{}
-King, Gary, and Langche Zeng. 2001a. ``Explaining Rare Events in
-International Relations.'' \emph{International Organization} 55 (03):
-693--715.
-
-\hypertarget{ref-king2001logistic}{}
----------. 2001b. ``Logistic Regression in Rare Events Data.''
-\emph{Political Analysis} 9 (2): 137--63.
-
-\hypertarget{ref-schmidt1989predicting}{}
-Schmidt, Peter, and Ann Dryden Witte. 1989. ``Predicting Criminal
-Recidivism Using Split Population Survival Time Models.'' \emph{Journal
-of Econometrics} 40 (1): 141--59.
-
-\hypertarget{ref-svolik2008authoritarian}{}
-Svolik, Milan. 2008. ``Authoritarian Reversals and Democratic
-Consolidation.'' \emph{American Political Science Review} 102 (2):
-153--68.
-
-\hypertarget{ref-ward2013learning}{}
-Ward, Michael D., Nils W. Metternich, Cassy L. Dorff, Max Gallop,
-Florian M. Hollenbach, Anna Schultz, and Simon Weschle. 2013. ``Learning
-from the Past and Stepping into the Future: Toward a New Generation of
-Conflict Prediction.'' \emph{International Studies Review} 15 (4):
-473--90.
+\bibliography{spdur}
 

### 2016-03-07 manuscript2.tex
more JSS style conformation, expanded examples
added package citations (and bibtex JSS style changes); added data
descriptives like ?bscoup, str(bscoup); changed bibtex titles to title
case; added a few more \pkg’s and \proglang’s, esp. in bibs;
replication runs through on new R session

@@
 }
-\title{Splitting it Up: The \pkg{spduration} Split-population Duration
+\title{Splitting It Up: The \pkg{spduration} Split-Population Duration
 Regression Package}
@@
 \Plainauthor{Andreas Beger, Daniel W. Hill, Jr., Nils W. Metternich, Shahryar Minhas, Michael D. Ward}
-\Plaintitle{Splitting it up: the spduration split-population duration regression
-package}
-\Shorttitle{\pkg{spduration}: Split-population duration regression}
+\Plaintitle{Splitting It Up: The spduration Split-Population Duration Regression
+Package}
+\Shorttitle{\pkg{spduration}: Split-Population Duration Regression}
 
@@
 We present an implementation of split-population duration regression in
-the \pkg{spduration} \proglang{R} package and an application to data on military
+the \pkg{spduration} \citep{beger2016spduration} package for \proglang{R} \citep{rstats2016} and an application to data on military
 coups. The statistical model accounts for units that are immune to a
@@
 R> data(bscoup)
+R> ?bscoup
+R> head(bscoup[, 1:5])
+\end{CodeInput}
+\begin{CodeOutput}
+  countryid year  couprisk recentcoups rwar
+1       909 1960 0.2200166           0    0
+2       909 1961 0.2200166           0    0
+3       909 1962 0.2200166           0    0
+4       909 1963 0.2200166           0    0
+5       909 1964 2.5702426           1    0
+6       909 1965 2.5702426           2    0
+\end{CodeOutput}
+\end{CodeChunk}
+
+The data are documented in `?bscoup`, which also gives a reference to the source citation.
+It consists of a little more than 5,000 observations of 162 countries from 1960 to 2000.
+Each row corresponds to a country $c$ in year $t$.
+Excluding the country and year identifiers, the data include 12 variables, including a binary indicator for successful coups.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> str(bscoup)
+\end{CodeInput}
+\begin{CodeOutput}
+'data.frame':	5463 obs. of  14 variables:
+ $ countryid  : num  909 909 909 909 909 909 909 920 920 920 ...
+ $ year       : num  1960 1961 1962 1963 1964 ...
+ $ couprisk   : num  0.22 0.22 0.22 0.22 2.57 ...
+ $ recentcoups: num  0 0 0 0 1 2 4 0 0 0 ...
+ $ rwar       : num  0 0 0 0 0 0 1 0 0 0 ...
+ $ milreg     : num  0 0 0 1 0 1 1 0 0 0 ...
+ $ wealth     : num  NA NA NA NA NA ...
+ $ instab     : num  4 2 9 9 12 13 12 NA 0 4 ...
+ $ coup       : Factor w/ 2 levels "no","yes": 1 1 1 2 2 2 1 1 1 1 ...
+ $ africa     : num  0 0 0 0 0 0 0 0 0 0 ...
+ $ eurnam     : num  0 0 0 0 0 0 0 1 1 1 ...
+ $ samerica   : num  0 0 0 0 0 0 0 0 0 0 ...
+ $ camerica   : num  0 0 0 0 0 0 0 0 0 0 ...
+ $ regconf    : num  0 0 0 0.019 0.019 ...
+ - attr(*, "datalabel")= chr ""
+ - attr(*, "time.stamp")= chr "30 Jul 2003 12:15"
+ - attr(*, "formats")= chr  "%9.0g" "%9.0g" "%9.0g" "%9.0g" ...
+ - attr(*, "types")= int  102 102 102 102 102 102 102 102 102 102 ...
+ - attr(*, "val.labels")= chr  "" "" "" "" ...
+ - attr(*, "var.labels")= chr  "country code .pr89" "year coded" "" "" ...
+ - attr(*, "version")= int 7
+ - attr(*, "label.table")=List of 1
+  ..$ hadcoup: Named int  0 1
+  .. ..- attr(*, "names")= chr  "no" "yes"
+\end{CodeOutput}
+\begin{CodeInput}
 R> bscoup$coup <- ifelse(bscoup$coup == "yes", 1, 0)
+R> table(bscoup$coup)
+\end{CodeInput}
+\begin{CodeOutput}
+   0    1
+5173  290 
+\end{CodeOutput}
+\end{CodeChunk}
+
+Alltogether the data include 290 coups, which makes for a positive rate of slightly more than 5\% of all observations.
+
+Before we can estimate a split-population duration model with the data, we need to add several variables that capture the survival characteristics of the data.
+
+\begin{CodeChunk}
+\begin{CodeInput}
 R> bscoup <- add_duration(bscoup, "coup", unitID = "countryid", 
@@
 \end{CodeInput}
+\begin{CodeOutput}
+Warning message:
+In attempt_date(data[, tID], freq) :
+  Converting to 'Date' class with yyyy-06-30
+\end{CodeOutput}
 \end{CodeChunk}
@@
 \code{tID} input to class ``Date'' given the provided argument in
-\code{freq}. Another important question is how to handle consecutive 1's
+\code{freq}, as indicated by the warning it returned in the example above. This 
+can be avoided by converting any numeric dates to \proglang{R} \code{Date}s first.
+
+Another important question is how to handle consecutive 1's
 in the response variable provided. This is controlled with the
@@
 and risk equations, and test statistics with \(p\)-values. These can be
-exported by calling \code{xtable} directly on the ``spdur'' object to
+exported by calling \code{xtable}\citep[see][]{dahl2016xtable} directly on the ``spdur'' object to
 produce Table 2.
@@
 \centering
-\begin{tabular}{lrrr}
+\begin{tabular}{lrrrr}
   \hline
-Parameter & Estimate & StdErr & p \\ 
+Parameter & Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
   \hline
-(Dur. Intercept) & 2.40 & 0.21 & 0.00 \\ 
-  milreg & $-$1.13 & 0.21 & 0.00 \\ 
-  instab & $-$0.09 & 0.02 & 0.00 \\ 
-  regconf & $-$2.52 & 2.16 & 0.24 \\ 
-  log(alpha) & $-$0.45 & 0.06 & 0.00 \\ 
-  (Risk Intercept) & 2.93 & 1.87 & 0.12 \\ 
-  couprisk & 0.59 & 0.32 & 0.07 \\ 
-  wealth & $-$0.36 & 0.28 & 0.20 \\ 
-  milreg.1 & 10.82 & 9.31 & 0.25 \\ 
-  rwar & $-$0.52 & 0.94 & 0.58 \\ 
-  regconf.1 & $-$5.43 & 5.62 & 0.33 \\ 
-  samerica & 2.09 & 1.45 & 0.15 \\ 
-  camerica & $-$0.39 & 0.73 & 0.59 \\ 
+Dur\_(Intercept) & 2.40 & 0.21 & 11.43 & 0.00 \\ 
+  Dur\_milreg & $-$1.13 & 0.21 & $-$5.46 & 0.00 \\ 
+  Dur\_instab & $-$0.09 & 0.02 & $-$4.79 & 0.00 \\ 
+  Dur\_regconf & $-$2.52 & 2.16 & $-$1.16 & 0.24 \\ 
+  log(alpha) & $-$0.45 & 0.06 & $-$7.09 & 0.00 \\ 
+  Risk\_(Intercept) & 2.93 & 1.87 & 1.57 & 0.12 \\ 
+  Risk\_couprisk & 0.59 & 0.32 & 1.82 & 0.07 \\ 
+  Risk\_wealth & $-$0.36 & 0.28 & $-$1.29 & 0.20 \\ 
+  Risk\_milreg & 10.82 & 9.31 & 1.16 & 0.25 \\ 
+  Risk\_rwar & $-$0.52 & 0.94 & $-$0.56 & 0.58 \\ 
+  Risk\_regconf & $-$5.43 & 5.62 & $-$0.97 & 0.33 \\ 
+  Risk\_samerica & 2.09 & 1.45 & 1.45 & 0.15 \\ 
+  Risk\_camerica & $-$0.39 & 0.73 & $-$0.53 & 0.59 \\ 
    \hline
 \end{tabular}
-\caption{Coup model with log-logistic hazard.} \label{loglog_table}
+\caption{Coup model with log-logistic hazard.} 
+\label{loglog_table}
 \end{table}
@@
 
-The package also includes a function, \code{plot_hazard}, that can be
-used with an \code{spdur} object to plot the estimated hazard rate.
-\code{plot_hazard} plots the conditional hazard, which is the
-probability of survival conditional on the covariates in the risk and
-duration equations, and conditional on survival up to time \(t\). The
-function calculates the estimated hazard rate as well as 90\% confidence
+The package includes two types of plots that show the estimated hazard rates and 
+predictive performance, respectively. The hazard rates can be plotted by either 
+calling \code{plot_hazard} directly, or with \code{plot(x, type = "hazard")}, on
+a fitted \code{spdur} object. It will produce a plot of the conditional hazard, which is the probability of survival conditional on the covariates in the risk and
+duration equations, and conditional on survival up to time \(t\), when holding
+all covariates at their sample means. The
+function calculates the average hazard rate as well as 90\% confidence
 intervals, which are produced by simulating values from the estimated
-sampling distributions of the model parameters. By default the function
+sampling distributions of the model parameters.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> plot(weib_model,   type="hazard", main="Weibull")
+R> plot(loglog_model, type="hazard", main="Loglog")
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.8\textwidth]{graphics/hazard-rates.pdf}
+\caption{Hazard rate plots for the Weibull and Loglogistic coup models.}
+\label{hazard-ex}
+\end{center}
+\end{figure}
+
+By default the \code{plot_hazard} function
 uses the mean values of the covariates during the simulations, but users
@@
 
-The hazard plots can also be called with
-\code{plot(..., type = "hazard")}. Without a type argument,
-\code{plot.spdur} will produce a separation plot, which is a graphical
+While \code{plot(x, type = "hazard")} will produce a hazard rate plot, without a type argument,
+\code{plot.spdur} will produce a separation plot. Separation plots are a graphical
 display for evaluating model predictions \citep{greenhill2011separation}. The code below produces Figure \ref{insamp}.
@@
 
-\section{Censoring considerations}
+\section{Additional discussion}
+
+\subsection{Censoring considerations}
 
@@
 
-\section{Alternative implementations}
+\subsection{Comparison with other software}
 
-The \pkg{survival} library, arguable the most well-established library for survival and event history analysis in \proglang{R}, allows for the estimation of a wide variet of semi-, non-, and parametric survival models, and provides facilities for handling and descriptive summaries of survival and event history data.
+The \pkg{survival} library \citep{thernau2000modeling, thernau2015survival}, arguable the most well-established library for survival and event history analysis in \proglang{R}, allows for the estimation of a wide variet of semi-, non-, and parametric survival models, and provides facilities for handling and descriptive summaries of survival and event history data.
 It does not include cure rate or split-population mixture models of the type we have implemented though.

### 2016-03-07 pre-screening-comments.Rmd
more JSS style conformation, expanded examples
added package citations (and bibtex JSS style changes); added data
descriptives like ?bscoup, str(bscoup); changed bibtex titles to title
case; added a few more \pkg’s and \proglang’s, esp. in bibs;
replication runs through on new R session

@@
 title: "Pre-screening comments"
-author: "Andreas Beger"
+author: "The Authors"
 date: "March 6, 2016"
@@
 
-*Changed `And` to `\AND`, added `R>` and `+` where needed to code formatting, and bibliography is not via BibTeX.*
+*Changed `And` to `\AND`, added `R>` and `+` where needed to code formatting, and bibliography is now via BibTeX.*
 
@@
 
+*Changed `-` to `$-$`, and the column names in `xtable.spdur` now match those output by the `xtable` internal methods, e.g. `xtable.lm`.*
+
 o There should be some discussion of software for related purposes (including references), pointing out relative (dis)advantages.
 
-*Added discussion of `smcure`, which is the only alternative implementation for R that we could find.*
+*Added discussion of `smcure`, which is the only alternative implementation for R that we could find, as well `survival`, which handles non-mixture duration models.*
 
@@
 
+*The two included data sets are documented in the package, e.g. `?coups`. We added some code and examples to the paper describing the data in some more details.*
+
 o The source package (.tar.gz) rather than the git .zip should be included in the submission.
@@
 o Title should be in title style (do capitalize all principal words and the first word after a colon).
+
+*Done, and the same for all citations.*
 
@@
 
+*The code examples have spaces before/after operators and after commas.*
+
 o If using "e.g." and "i.e." add a comma after the period to keep LaTeX from interpreting them as the end of a sentence, i.e.: "e.g., " and "i.e., ".
@@
 	- `\proglang`, `\pkg` and `\code` have been used for highlighting throughout the paper (including titles and references), except where explicitly escaped.
+
+*Checked and added where needed, including in the references.*
 

### 2016-03-07 replication.R

@@ -9,9 +9,17 @@ library("scales")
 
 
 data(bscoup)
+?bscoup
+head(bscoup[, 1:5])
+
+str(bscoup)
+
 bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
-bscoup      <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
-                            freq="year", ongoing = FALSE)
+table(bscoup$coup)
+
+bscoup <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
+                       freq="year", ongoing = FALSE)
+
 
 weib_model <- spdur(
   duration ~ milreg + instab + regconf,
@@ -29,9 +37,9 @@ matrix(c(
 
 
 library("xtable")
-tbl <- xtable(loglog_model, caption="Coup model with log-logistic hazard",
+tbl <- xtable(loglog_model, caption="Coup model with log-logistic hazard.",
               label="loglog_table")
-print(tbl, caption.placement="top", comment=FALSE, include.rownames=FALSE)
+print(tbl, caption.placement="bottom", comment=FALSE, include.rownames=FALSE)
 
 #   Plotting functions
 #   ____________________
@@ -114,7 +122,8 @@ glm_model2    <- glm(
 c(Weibull = AIC(weib_model2), 
   Loglog = AIC(loglog_model2), 
   Logistic = AIC(glm_model2))
-  
+
+
 #   Plot of positive rates in several published papers
 #   __________________________
 

### 2016-03-08 manuscript2.tex
added cites for smcure

@@
 
-Alltogether the data include 290 coups, which makes for a positive rate of slightly more than 5\% of all observations.
+Altogether the data include 290 coups, which makes for a positive rate of slightly more than 5\% of all observations.
 
@@
 
-The \pkg{survival} library \citep{thernau2000modeling, thernau2015survival}, arguable the most well-established library for survival and event history analysis in \proglang{R}, allows for the estimation of a wide variet of semi-, non-, and parametric survival models, and provides facilities for handling and descriptive summaries of survival and event history data.
+The \pkg{survival} library \citep{thernau2000modeling, thernau2015survival}, arguably the most well-established library for survival and event history analysis in \proglang{R}, allows for the estimation of a wide variety of semi-, non-, and parametric survival models, and provides facilities for handling and descriptive summaries of survival and event history data.
 It does not include cure rate or split-population mixture models of the type we have implemented though.
@@
 Otherwise, the \pkg{smcure} package \citep{cai2012smcureR}, described in \citet{cai2012smcure}, appears to be the only alternative split-population/cure rate duration model implementation in \proglang{R}.
-It implements semi-parametric proportional hazards and accelerated failure time cure models.
-The semi-paramatric estimation of the hazard function shape obviates the need to make an assumption about its shape, which can be difficult in practice to empirically evalute, as the \pkg{smcure} discuss.
-However, it is much more limited in practical application as it only accomodates case-based data, not time-varying covariates.
+It implements semi-parametric proportional hazards and accelerated failure time cure models using methods outlined in \citep{peng2003fitting} and \citep{zhang2007new}.
+The semi-parametric estimation of the hazard function shape requires no assumption about its shape, which can be difficult in practice to empirically evaluate, as the authors of \pkg{smcure} discuss.
+However, \pkg{spduration} offers two advantages over \pkg{smcure}.
+First, it is much less limited in practical application as it allows for the inclusion of time-varying covariates.
+Second, it features more post-estimation methods, including those discussed above to create and evaluate model predictions.
 
-Furthermore, \proglang{Stata} implements cure models similar to our own, and which can handle time-varying covariates, but this is of course of limited appeal to \proglang{R} users.
+Finally, \proglang{Stata} implements cure models similar to our own, and which can handle time-varying covariates, but this is of course of limited appeal to \proglang{R} users.
 

### 2016-03-11 manuscript2.tex
added discussion and cites for other packages

@@
 
-One interesting extension of our project in the future would be to integrate and adapt the \code{Surv} class for survival data in \pkg{survival} to the \pkg{spduration} library.
+One useful extension of our project in the future would be to integrate and adapt the \code{Surv} class for survival data in \pkg{survival} to the \pkg{spduration} library.
 This would require some modification of the class as our model also requires information on the risk status of spells, but would give access to the much broader functionality, especially for descriptive summaries of data, in the \pkg{survival} library.
 
-Otherwise, the \pkg{smcure} package \citep{cai2012smcureR}, described in \citet{cai2012smcure}, appears to be the only alternative split-population/cure rate duration model implementation in \proglang{R}.
-It implements semi-parametric proportional hazards and accelerated failure time cure models using methods outlined in \citep{peng2003fitting} and \citep{zhang2007new}.
-The semi-parametric estimation of the hazard function shape requires no assumption about its shape, which can be difficult in practice to empirically evaluate, as the authors of \pkg{smcure} discuss.
-However, \pkg{spduration} offers two advantages over \pkg{smcure}.
-First, it is much less limited in practical application as it allows for the inclusion of time-varying covariates.
-Second, it features more post-estimation methods, including those discussed above to create and evaluate model predictions.
+Other currently available \proglang{R} routines for estimation of split-population/cure rate duration models include the packages \pkg{smcure}  \citep{cai2012smcureR}, described in \citet{cai2012smcure}, and \pkg{nltm} \citep{garibotti2010nltm}.
+\pkg{smcure} implements semi-parametric proportional hazards and accelerated failure time cure models using estimation procedures presented in \citet{peng2003fitting} and \citet{zhang2007new}.\footnote{There is also the older \pkg{gfcure} package, which estimates the parametric accelerated failure time cure model discussed in \citet{peng1998generalized}.
+A version for Windows is available at \url{http://post.queensu.ca/~pengp/software.html}.} The package \pkg{nltm} will estimate a semi-parametric proportional hazard cure model, and a number of other survival models, using the approach developed in \citet{tsodikov2003semiparametric,tsodikov2007profile}.
+Semi-parametric estimation of the hazard function is attractive because it requires no assumption about its shape, and such assumptions can be difficult in practice to empirically evaluate.
+The main advantage \pkg{spduration} offers is the ability to include time varying covariates, which gives it a broader range of practical applications, as \pkg{smcure} and\pkg{nltm} will only accommodate case-based data.
+\pkg{spduration} also features more post-estimation methods than alternative packages, including those discussed above to create and evaluate model predictions.
 
-Finally, \proglang{Stata} implements cure models similar to our own, and which can handle time-varying covariates, but this is of course of limited appeal to \proglang{R} users.
+Finally, there are \proglang{Stata} routines for implementing cure models \citep[see, e.g.,][]{lambert2007modeling,buxton2013cureregr}, some of which can handle time-varying covariates, but this is of course of limited appeal to \proglang{R} users.
 

### 2017-02-17 spdurarXiv.tex

--- 2016-03-11 manuscript2.tex
@@
-\documentclass[article]{jss}
+\documentclass[12pt,onesided]{amsart} 
+\usepackage{natbib}
+\usepackage[default,osfigures,scale=0.95]{opensans}{}
+\usepackage[T1]{fontenc}
+% to colorize links in document. See color specification below
+\usepackage[pdftex,hyperref,x11names]{xcolor}
+% load the hyper-references package and set document info
+\usepackage[pdftex]{hyperref}
+% layout control
+\usepackage{geometry}
+\geometry{verbose,tmargin=1.25in,bmargin=1.25in,lmargin=1.1in,rmargin=1.1in}
+\usepackage{fancyhdr}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% math typesetting
+\usepackage{array}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{amsfonts}
+\usepackage{relsize}
+\usepackage{mathtools}
+\usepackage{bm}
 
-%% almost as usual
-\author{
-Andreas Beger\\Duke University \And Daniel W.
-Hill, Jr.\\University of Georgia \And Nils W.
-Metternich\\University College London \AND Shahryar Minhas\\Duke University \And Michael D.
-Ward\\Duke University
+% restricts float objects to be inserted before end of section
+% creates float barriers
+\usepackage[section]{placeins}
+
+% tables
+\usepackage{tabularx}
+\usepackage{booktabs}
+\usepackage{multicol}
+\usepackage{multirow}
+
+% to adapt caption style
+\usepackage[font={small},labelfont=bf]{caption}
+
+% footnotes at bottom
+\usepackage[bottom]{footmisc}
+
+% graphics stuff
+\usepackage{subfig}
+\usepackage{graphicx}
+\usepackage[space]{grffile} % allows us to specify directories that have spaces
+\usepackage{placeins} % prevents floats from moving past a \FloatBarrier
+\usepackage{tikz}
+\usepackage{rotating}
+
+% Spacing
+\usepackage[singlespacing]{setspace}
+
+% -------------------------------------------------- %
+\usepackage{listings}   
+\lstset{ 
+  language=R,                     % the language of the code
+  basicstyle=\small\ttfamily, % the size of the fonts that are used for the code
+  %numbers=left,                   % where to put the line-numbers
+  %numberstyle=\tiny\color{blue},  % the style that is used for the line-numbers
+  %stepnumber=1,                   % the step between two line-numbers. If it is 1, 
+                                  % each line will be numbered
+  numbersep=5pt,                  % how far the line-numbers are from the code
+  backgroundcolor=\color{white},  % choose the background color. You must add 
+  showspaces=false,               % show spaces adding particular underscores
+  showstringspaces=false,         % underline spaces within strings
+  showtabs=false,                 % show tabs within strings adding particular underscores
+  rulecolor=\color{black},        % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. commens (green here))
+  tabsize=2,                      % sets default tabsize to 2 spaces
+  captionpos=b,                   % sets the caption-position to bottom
+  breaklines=true,                % sets automatic line breaking
+  breakatwhitespace=false,        % sets if automatic breaks only happen at whitespace
+  keywordstyle=\color{blue},      % keyword style
+  commentstyle=\color{teal},      % comment style
+  stringstyle=\color{green}       % string literal style
 }
+
+% -------------------- page template -------------------- %
+
+\setlength{\headheight}{15pt}
+\setlength{\headsep}{20pt}
+\pagestyle{fancyplain}
+
+\fancyhf{}
+
+\lhead{\fancyplain{}{}}
+\chead{\fancyplain{}{Split Population Duration Models}}
+\rhead{\fancyplain{}{\today}}
+\rfoot{\fancyplain{}{\thepage}}
+
+% ----------------------------------------------- %
+\usepackage{amsmath} 
+\usepackage{booktabs}
+
+% -------------------- customizations -------------------- %
+
+% command for R package font
+\newcommand{\pkg}[1]{{\fontseries{b}\selectfont #1}}
+
+% approx iid
+\newcommand\simiid{\stackrel{\mathclap{\normalfont\mbox{\tiny{iid}}}}{\sim}}
+
+% ----------------------------------
+
 \title{Splitting It Up: The \pkg{spduration} Split-Population Duration
@@
 
-%% for pretty printing and a nice hypersummary also set:
-\Plainauthor{Andreas Beger, Daniel W. Hill, Jr., Nils W. Metternich, Shahryar Minhas, Michael D. Ward}
-\Plaintitle{Splitting It Up: The spduration Split-Population Duration Regression
-Package}
-\Shorttitle{\pkg{spduration}: Split-Population Duration Regression}
+\author[Beger]{Andreas Beger}
+\address{Andreas Beger: Department of Political Science}
+\curraddr{Predictive Heuristics}
+\email{adbeger@gmail.com}
 
-%% an abstract and keywords
-\Abstract{
-We present an implementation of split-population duration regression in
-the \pkg{spduration} \citep{beger2016spduration} package for \proglang{R} \citep{rstats2016} and an application to data on military
-coups. The statistical model accounts for units that are immune to a
-certain outcome and not part of the duration process the researcher is
-primarily interested in. We provide insights that if immune units exist,
-we can significantly increase the predictive performance compared to
-standard duration models. The package includes estimation and several
-post-estimation methods for split-population Weibull and Loglogistic
-models.
-}
+\author[Hill]{Daniel W. Hill}
+\address{Daniel Hill: Department of International Affairs}
+\curraddr{University of Georgia Athens, GA, USA}
+\email{dwhill@uga.edudwhill@uga.edu}
 
-\Keywords{regression, rare events, \proglang{R}}
-\Plainkeywords{regression, rare events, R} %% without formatting
-%% at least one keyword must be supplied
+\author[Metternich]{Nils W. Metternich}
+\address{Nils W. Metternich: Department of Political Science}
+\curraddr{University College London, London, WC1H 9QU, UK}
+\email{n.metternich@ucl.ac.uk}
 
-%% publication information
-%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
-%% \Volume{50}
-%% \Issue{9}
-%% \Month{June}
-%% \Year{2012}
-%% \Submitdate{2012-06-04}
-%% \Acceptdate{2012-06-04}
+\author[Minhas]{Shahryar Minhas}
+\address{Shahryar Minhas: Department of Political Science}
+\curraddr{Duke University, Durham, NC, 27708, USA}
+\email{shahyar.minhas@duke.edu}
 
-%% The address of (at least) one author should be given
-%% in the following format:
-\Address{
-    Andreas Beger\\
-  Duke University\\
-  Department of Political Science, Duke University Durham, NC, USA\\
-  E-mail: \href{mailto:ab428@duke.edu}{\nolinkurl{ab428@duke.edu}}\\
-  URL: something\\~\\
-      Daniel W. Hill, Jr.\\
-  University of Georgia\\
-  Department of International Affairs, University of Georgia Athens, GA,
-  USA\\
-  E-mail: \href{mailto:dwhill@uga.edu}{\nolinkurl{dwhill@uga.edu}}\\
-
-      Nils W. Metternich\\
-  University College London\\
-  Department of Political Science, University College London London, UK\\
-  E-mail: \href{mailto:n.metternich@ucl.ac.uk}{\nolinkurl{n.metternich@ucl.ac.uk}}\\
-
-      Shahryar Minhas\\
-  Duke University\\
-  Department of Political Science, Duke University Durham, NC, USA\\
-  E-mail: \href{mailto:sfm12@duke.edu}{\nolinkurl{sfm12@duke.edu}}\\
-
-      Michael D. Ward\\
-  Duke University\\
-  Department of Political Science, Duke University Durham, NC, USA\\
-  E-mail: \href{mailto:mw160@duke.edu}{\nolinkurl{mw160@duke.edu}}\\
-
-  }
-
-%% It is also possible to add a telephone and fax number
-%% before the e-mail in the following format:
-%% Telephone: +43/512/507-7103
-%% Fax: +43/512/507-2851
-
-%% for those who use Sweave please include the following line (with % symbols):
-%% need no \usepackage{Sweave.sty}
-
-%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\usepackage{amsmath} 
-\usepackage{booktabs}
+\author[Ward]{Michael D. Ward}
+\address{Michael D. Ward: Department of Political Science}
+\curraddr{Duke University, Durham, NC, 27708, USA}
+\email{michael.d.ward@duke.edu}
 
 \begin{document}
+\maketitle
 
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
+\begin{lstlisting}
 R> library("spduration")
@@
 R> head(bscoup[, 1:5])
-\end{CodeInput}
-\begin{CodeOutput}
+
   countryid year  couprisk recentcoups rwar
@@
 6       909 1965 2.5702426           2    0
-\end{CodeOutput}
-\end{CodeChunk}
+
+\end{lstlisting}
 
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
+\begin{lstlisting}
+
 R> str(bscoup)
-\end{CodeInput}
-\begin{CodeOutput}
+
 'data.frame':	5463 obs. of  14 variables:
@@
   .. ..- attr(*, "names")= chr  "no" "yes"
-\end{CodeOutput}
-\begin{CodeInput}
+
 R> bscoup$coup <- ifelse(bscoup$coup == "yes", 1, 0)
 R> table(bscoup$coup)
-\end{CodeInput}
-\begin{CodeOutput}
+
    0    1
 5173  290
-\end{CodeOutput}
-\end{CodeChunk}
+
+\end{lstlisting}
 
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
+\begin{lstlisting}
+
 R> bscoup <- add_duration(bscoup, "coup", unitID = "countryid", 
 +    tID = "year", freq = "year", ongoing = FALSE)
-\end{CodeInput}
-\begin{CodeOutput}
+
 Warning message:
@@
   Converting to 'Date' class with yyyy-06-30
-\end{CodeOutput}
-\end{CodeChunk}
 
-The \code{add\_duration} function takes as input a data frame with a
+\end{lstlisting}
+
+The \texttt{add\_duration} function takes as input a data frame with a
 binary response variable--``coup''--that measures the occurrence of the
@@
 
-The \code{add_duration} functions identifies spells in the data and
+The \verb|add_duration| functions identifies spells in the data and
 returns the data frame with several additional variables needed for
@@
 otherwise. These are then used as response variables for the formula in
-the \code{spdur} function.
+the \texttt{spdur} function.
 
@@
 supports annual, monthly, or daily data and will try to convert the
-\code{tID} input to class ``Date'' given the provided argument in
-\code{freq}, as indicated by the warning it returned in the example above. This 
-can be avoided by converting any numeric dates to \proglang{R} \code{Date}s first.
+\texttt{tID} input to class ``Date'' given the provided argument in
+\texttt{freq}, as indicated by the warning it returned in the example above. This 
+can be avoided by converting any numeric dates to $\mathcal{R}$ \texttt{Date}s first.
 
@@
 in the response variable provided. This is controlled with the
-\code{ongoing} argument. In some cases like civil war occurrence, the
+\texttt{ongoing} argument. In some cases like civil war occurrence, the
 \emph{y} variable records all years during which a country experienced a
-civil war. Here \code{ongoing} should be set to true so that the models
+civil war. Here \texttt{ongoing} should be set to true so that the models
 try to predict the onset of the civil war, but disregard ongoing
-conflicts (\code{failure} is set to \code{NA} for these cases, dropping
+conflicts (\texttt{failure} is set to \texttt{NA} for these cases, dropping
 them from analysis). With ongoing set to false, successive 1's in
@@
 
-The \code{spdur} function is the primary function in the package and
-produces a regression model object of class \code{spdur} which can then
+The \texttt{spdur} function is the primary function in the package and
+produces a regression model object of class \texttt{spdur} which can then
 be used with further methods. We begin by fitting first a Weibull and
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
+\begin{lstlisting}
+
 R> weib_model <- spdur(
@@
 +    data = bscoup, distr = "loglog", silent = TRUE)
-\end{CodeInput}
-\end{CodeChunk}
 
-Using the \code{summary} function on either model object will produce
+\end{lstlisting}
+
+Using the \texttt{summary} function on either model object will produce
 standard output showing the model formula, estimates for the duration
 and risk equations, and test statistics with \(p\)-values. These can be
-exported by calling \code{xtable}\citep[see][]{dahl2016xtable} directly on the ``spdur'' object to
+exported by calling \texttt{xtable}\citep[see][]{dahl2016xtable} directly on the ``spdur'' object to
 produce Table 2.
+\clearpage
 
-\begin{CodeChunk}
-\begin{CodeInput}
+\begin{lstlisting}
 R> library("xtable")
@@
 +    include.rownames = FALSE)
-\end{CodeInput}
-\end{CodeChunk}
+\end{lstlisting}
 
@@
 
-One may also use the \code{AIC} and \code{BIC} function to calculate the
-information criterion statistics for \code{spdur} objects. In our
+One may also use the \texttt{AIC} and \texttt{BIC} function to calculate the
+information criterion statistics for \texttt{spdur} objects. In our
 example, both models are close enough in both statistics to be
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
+\begin{lstlisting}
+
 R> matrix(c(
@@
 +    ), ncol = 2, dimnames = list(c("Weibull", "Loglog"), c("AIC", "BIC")))
-\end{CodeInput}
-\begin{CodeOutput}
+
              AIC      BIC
@@
 Loglog  1331.214 1350.278
-\end{CodeOutput}
-\end{CodeChunk}
+
+\end{lstlisting}
 
@@
 predictive performance, respectively. The hazard rates can be plotted by either 
-calling \code{plot_hazard} directly, or with \code{plot(x, type = "hazard")}, on
-a fitted \code{spdur} object. It will produce a plot of the conditional hazard, which is the probability of survival conditional on the covariates in the risk and
+calling \verb|plot_hazard| directly, or with \texttt{plot(x, type = "hazard")}, on
+a fitted \texttt{spdur} object. It will produce a plot of the conditional hazard, which is the probability of survival conditional on the covariates in the risk and
 duration equations, and conditional on survival up to time \(t\), when holding
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
+\begin{lstlisting}
+
 R> plot(weib_model,   type="hazard", main="Weibull")
 R> plot(loglog_model, type="hazard", main="Loglog")
-\end{CodeInput}
-\end{CodeChunk}
+
+\end{lstlisting}
 
@@
 
-By default the \code{plot_hazard} function
+By default the \verb|plot_hazard| function
 uses the mean values of the covariates during the simulations, but users
 can choose specific covariate values by entering them as vectors in the
-arguments \code{xvals} and \code{zvals}, which correspond to the
+arguments \texttt{xvals} and \texttt{zvals}, which correspond to the
 covariates in the duration and risk equations, respectively. The command
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
+\begin{lstlisting}
+
 R> plot_hazard(loglog_model, main = "A")
-\end{CodeInput}
-\end{CodeChunk}
+
+\end{lstlisting}
 
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
+\begin{lstlisting}
+
 R> plot_hazard(loglog_model, xvals = c(1, 1, 10, 0.05), 
 +    zvals = c(1, 7, 8.64, 1, 1, 0.05, 0, 0), main = "B")
-\end{CodeInput}
-\end{CodeChunk}
 
-While \code{plot(x, type = "hazard")} will produce a hazard rate plot, without a type argument,
-\code{plot.spdur} will produce a separation plot. Separation plots are a graphical
+\end{lstlisting}
+
+While \texttt{plot(x, type = "hazard")} will produce a hazard rate plot, without a type argument,
+\texttt{plot.spdur} will produce a separation plot. Separation plots are a graphical
 display for evaluating model predictions \citep{greenhill2011separation}. The code below produces Figure \ref{insamp}.
 
-\begin{CodeChunk}
-\begin{CodeInput}
+\begin{lstlisting}
+
 R> plot(weib_model)
 R> plot(loglog_model)
-\end{CodeInput}
-\end{CodeChunk}
+
+\end{lstlisting}
 
@@
 
-The option \code{endSpellOnly} is set to \code{FALSE} so that every
+The option \texttt{endSpellOnly} is set to \texttt{FALSE} so that every
 observation, not only those at the end of a spell, is used in the plot.
-By default the \code{plot} function will calculate the conditional
+By default the \texttt{plot} function will calculate the conditional
 hazard for each observation. The separation plot sorts observations from
@@
 
-Underlying both plotting functions is the \code{predict} function, which
-can be used on an object of class \code{spdur} to generate several kinds
+Underlying both plotting functions is the \texttt{predict} function, which
+can be used on an object of class \texttt{spdur} to generate several kinds
 of predictions, including the probability that an observation is
@@
 predictions. We will use data from 1996 onwards as the test set, and
-prior data for training purposes. The \code{add_duration} function
+prior data for training purposes. The \verb|add_duration| function
 retrospectively codes the risk variable based on how a spell ended, and
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
+\begin{lstlisting}
+
 R> data(bscoup)
@@
 +    tID = "year", freq = "year", ongoing = FALSE)
-\end{CodeInput}
-\end{CodeChunk}
+
+\end{lstlisting}
 
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
+\begin{lstlisting}
+
 R> coup_test  <- add_duration(bscoup, "coup", unitID = "countryid", 
@@
 R> coup_test  <- coup_test[coup_test$year >= 1996, ]
-\end{CodeInput}
-\end{CodeChunk}
+
+\end{lstlisting}
 
@@
 predictions from these models for the test data, using
-\code{predict(..., newdata = coup_test)}.
+\verb|predict(..., newdata = coup_test)|.
 
-\begin{CodeChunk}
-\begin{CodeInput}
+\begin{lstlisting}
+
 R> weib_model2   <- spdur(
@@
 +    data = coup_train, distr = "loglog", silent = TRUE)
-\end{CodeInput}
-\end{CodeChunk}
 
-\begin{CodeChunk}
-\begin{CodeInput}
+\end{lstlisting}
+
+\begin{lstlisting}
+
 R> weib2_test_p   <- predict(weib_model2, newdata = coup_test)
 R> loglog2_test_p <- predict(loglog_model2, newdata = coup_test)
-\end{CodeInput}
-\end{CodeChunk}
+
+\end{lstlisting}
 
 Since we are predicting for data that is not contained in the ``spdur''
-model objects, we have to use the \code{separationplot} functions
+model objects, we have to use the \texttt{separationplot} functions
 directly from the package to produce Figure \ref{oos-sepplots}.
 
-\begin{CodeChunk}
-\begin{CodeInput}
+\begin{lstlisting}
+
 R> library("separationplot")
@@
 R> separationplot(loglog2_test_p, obs_y, newplot = FALSE)
-\end{CodeInput}
-\end{CodeChunk}
+
+\end{lstlisting}
 
@@
 information of previous failures when constructing the necessary
-duration variables with \code{add\_duration()}.
+duration variables with \verb|add_duration()|.
 
@@
 function, they impact the coding of at risk vs.~immune spells. The
-\code{add\_duration()} function retroactively codes all observations in
+\verb|add_duration()| function retroactively codes all observations in
 a spell as at risk if the spell itself ended in failure. Right-censored
@@
 
-The \pkg{survival} library \citep{thernau2000modeling, thernau2015survival}, arguably the most well-established library for survival and event history analysis in \proglang{R}, allows for the estimation of a wide variety of semi-, non-, and parametric survival models, and provides facilities for handling and descriptive summaries of survival and event history data.
+The \pkg{survival} library \citep{thernau2000modeling, thernau2015survival}, arguably the most well-established library for survival and event history analysis in $\mathcal{R}$, allows for the estimation of a wide variety of semi-, non-, and parametric survival models, and provides facilities for handling and descriptive summaries of survival and event history data.
 It does not include cure rate or split-population mixture models of the type we have implemented though.
 
-One useful extension of our project in the future would be to integrate and adapt the \code{Surv} class for survival data in \pkg{survival} to the \pkg{spduration} library.
+One useful extension of our project in the future would be to integrate and adapt the \texttt{Surv} class for survival data in \pkg{survival} to the \pkg{spduration} library.
 This would require some modification of the class as our model also requires information on the risk status of spells, but would give access to the much broader functionality, especially for descriptive summaries of data, in the \pkg{survival} library.
 
-Other currently available \proglang{R} routines for estimation of split-population/cure rate duration models include the packages \pkg{smcure}  \citep{cai2012smcureR}, described in \citet{cai2012smcure}, and \pkg{nltm} \citep{garibotti2010nltm}.
+Other currently available $\mathcal{R}$ routines for estimation of split-population/cure rate duration models include the packages \pkg{smcure}  \citep{cai2012smcureR}, described in \citet{cai2012smcure}, and \pkg{nltm} \citep{garibotti2010nltm}.
 \pkg{smcure} implements semi-parametric proportional hazards and accelerated failure time cure models using estimation procedures presented in \citet{peng2003fitting} and \citet{zhang2007new}.\footnote{There is also the older \pkg{gfcure} package, which estimates the parametric accelerated failure time cure model discussed in \citet{peng1998generalized}.
@@
 
-Finally, there are \proglang{Stata} routines for implementing cure models \citep[see, e.g.,][]{lambert2007modeling,buxton2013cureregr}, some of which can handle time-varying covariates, but this is of course of limited appeal to \proglang{R} users.
+Finally, there are \texttt{Stata} routines for implementing cure models \citep[see, e.g.,][]{lambert2007modeling,buxton2013cureregr}, some of which can handle time-varying covariates, but this is of course of limited appeal to $\mathcal{R}$ users.
 
@@
 
-\bibliography{spdur}
-
+\section{REFERENCES}
+\bibliographystyle{authoryear}
+%\bibliographystyle{apsr}
+\bibliography{spdur.bib}
 \end{document}

### 2017-06-22 R Journal/beger-hill-metternich-minhas-ward.tex
Formatted for R Journal

--- 2016-03-11 manuscript2.tex
@@
-\documentclass[article]{jss}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% almost as usual
-\author{
-Andreas Beger\\Duke University \And Daniel W.
-Hill, Jr.\\University of Georgia \And Nils W.
-Metternich\\University College London \AND Shahryar Minhas\\Duke University \And Michael D.
-Ward\\Duke University
-}
+% !TeX root = RJwrapper.tex
 \title{Splitting It Up: The \pkg{spduration} Split-Population Duration
 Regression Package}
+\author{by Andreas Beger, Daniel W. Hill, Jr., Nils. W. Metternich, Shahryar Minhas, and Michael D. Ward}
 
-%% for pretty printing and a nice hypersummary also set:
-\Plainauthor{Andreas Beger, Daniel W. Hill, Jr., Nils W. Metternich, Shahryar Minhas, Michael D. Ward}
-\Plaintitle{Splitting It Up: The spduration Split-Population Duration Regression
-Package}
-\Shorttitle{\pkg{spduration}: Split-Population Duration Regression}
+\maketitle
 
-%% an abstract and keywords
-\Abstract{
+\abstract{
 We present an implementation of split-population duration regression in
-the \pkg{spduration} \citep{beger2016spduration} package for \proglang{R} \citep{rstats2016} and an application to data on military
+the \CRANpkg{spduration} \citep{beger2016spduration} package for {\sf R} \citep{rstats2016} and an application to data on military
 coups. The statistical model accounts for units that are immune to a
@@
 }
-
-\Keywords{regression, rare events, \proglang{R}}
-\Plainkeywords{regression, rare events, R} %% without formatting
-%% at least one keyword must be supplied
-
-%% publication information
-%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
-%% \Volume{50}
-%% \Issue{9}
-%% \Month{June}
-%% \Year{2012}
-%% \Submitdate{2012-06-04}
-%% \Acceptdate{2012-06-04}
-
-%% The address of (at least) one author should be given
-%% in the following format:
-\Address{
-    Andreas Beger\\
-  Duke University\\
-  Department of Political Science, Duke University Durham, NC, USA\\
-  E-mail: \href{mailto:ab428@duke.edu}{\nolinkurl{ab428@duke.edu}}\\
-  URL: something\\~\\
-      Daniel W. Hill, Jr.\\
-  University of Georgia\\
-  Department of International Affairs, University of Georgia Athens, GA,
-  USA\\
-  E-mail: \href{mailto:dwhill@uga.edu}{\nolinkurl{dwhill@uga.edu}}\\
-
-      Nils W. Metternich\\
-  University College London\\
-  Department of Political Science, University College London London, UK\\
-  E-mail: \href{mailto:n.metternich@ucl.ac.uk}{\nolinkurl{n.metternich@ucl.ac.uk}}\\
-
-      Shahryar Minhas\\
-  Duke University\\
-  Department of Political Science, Duke University Durham, NC, USA\\
-  E-mail: \href{mailto:sfm12@duke.edu}{\nolinkurl{sfm12@duke.edu}}\\
-
-      Michael D. Ward\\
-  Duke University\\
-  Department of Political Science, Duke University Durham, NC, USA\\
-  E-mail: \href{mailto:mw160@duke.edu}{\nolinkurl{mw160@duke.edu}}\\
-
-  }
-
-%% It is also possible to add a telephone and fax number
-%% before the e-mail in the following format:
-%% Telephone: +43/512/507-7103
-%% Fax: +43/512/507-2851
-
-%% for those who use Sweave please include the following line (with % symbols):
-%% need no \usepackage{Sweave.sty}
-
-%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\usepackage{amsmath} 
-\usepackage{booktabs}
-
-\begin{document}
 
@@
 The last component to complete the likelihood is the choice of a
-distribution for the shape of the hazard rate. The \pkg{spduration}
+distribution for the shape of the hazard rate. The \CRANpkg{spduration}
 package implements two hazard rate shapes, Weibull and log-logistic:
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> library("spduration")
-R> data(bscoup)
-R> ?bscoup
-R> head(bscoup[, 1:5])
-\end{CodeInput}
-\begin{CodeOutput}
+\begin{example}
+  library("spduration")
+  data(bscoup)
+  ?bscoup
+  head(bscoup[, 1:5])
     countryid  year  couprisk recentcoups rwar
@@
   6       909  1965 2.5702426           2    0
-\end{CodeOutput}
-\end{CodeChunk}
+\end{example}
 
-The data are documented in `?bscoup`, which also gives a reference to the source citation.
+The data are documented in \samp{?bscoup}, which also gives a reference to the source citation.
 It consists of a little more than 5,000 observations of 162 countries from 1960 to 2000.
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> str(bscoup)
-\end{CodeInput}
-\begin{CodeOutput}
+\begin{example}
+  str(bscoup)
   'data.frame':	5463 obs. of  14 variables:
@@
   .. ..- attr(*, "names")= chr  "no" "yes"
-\end{CodeOutput}
-\begin{CodeInput}
-R> bscoup$coup <- ifelse(bscoup$coup == "yes", 1, 0)
-R> table(bscoup$coup)
-\end{CodeInput}
-\begin{CodeOutput}
+  bscoup$coup <- ifelse(bscoup$coup == "yes", 1, 0)
+  table(bscoup$coup)
   0    1
   5173  290 
-\end{CodeOutput}
-\end{CodeChunk}
+\end{example}
 
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> bscoup <- add_duration(bscoup, "coup", unitID = "countryid", 
+\begin{example}
+  bscoup <- add_duration(bscoup, "coup", unitID = "countryid", 
   +    tID = "year", freq = "year", ongoing = FALSE)
-\end{CodeInput}
-\begin{CodeOutput}
   Warning message:
@@
   Converting to 'Date' class with yyyy-06-30
-\end{CodeOutput}
-\end{CodeChunk}
+\end{example}
 
 The \code{add\_duration} function takes as input a data frame with a
-binary response variable--``coup''--that measures the occurrence of the
+binary response variable--\code{coup}--that measures the occurrence of the
 event, or failure, recorded over discrete time periods. We assume that
@@
 
-The \code{add_duration} functions identifies spells in the data and
+The \code{add\_duration} functions identifies spells in the data and
 returns the data frame with several additional variables needed for
 estimation. The most important among these are a counter of elapsed time
-in a spell, ``duration''. It also codes that a spell was at risk
-(``atrisk==1'') if it ended in a failure event, and not at risk
+in a spell, \code{duration}. It also codes that a spell was at risk
+(\code{atrisk==1}) if it ended in a failure event, and not at risk
 otherwise. These are then used as response variables for the formula in
@@
 supports annual, monthly, or daily data and will try to convert the
-\code{tID} input to class ``Date'' given the provided argument in
+\code{tID} input to class \code{Date} given the provided argument in
 \code{freq}, as indicated by the warning it returned in the example above. This 
-can be avoided by converting any numeric dates to \proglang{R} \code{Date}s first.
+can be avoided by converting any numeric dates to {\sf R} \code{Date}s first.
 
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> weib_model <- spdur(
+\begin{example}
+  weib_model <- spdur(
   +    duration ~ milreg + instab + regconf,
@@
 
-R> loglog_model <- spdur(
+  loglog_model <- spdur(
   +    duration ~ milreg + instab + regconf,
@@
   +    data = bscoup, distr = "loglog", silent = TRUE)
-\end{CodeInput}
-\end{CodeChunk}
+\end{example}
 
@@
 and risk equations, and test statistics with \(p\)-values. These can be
-exported by calling \code{xtable}\citep[see][]{dahl2016xtable} directly on the ``spdur'' object to
+exported by calling \code{xtable}\citep[see][]{dahl2016xtable} directly on the \code{spdur} object to
 produce Table 2.
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> library("xtable")
-R> tbl <- xtable(loglog_model, caption = "Coup model with log-logistic hazard",
+\begin{example}
+  library("xtable")
+  tbl <- xtable(loglog_model, caption = "Coup model with log-logistic hazard",
   +    label = "loglog_table")
-R> print(tbl, caption.placement = "top", comment = FALSE, 
+  print(tbl, caption.placement = "top", comment = FALSE, 
   +    include.rownames = FALSE)
-\end{CodeInput}
-\end{CodeChunk}
+\end{example}
 
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> matrix(c(
+\begin{example}
+  matrix(c(
   +    AIC(weib_model), AIC(loglog_model), BIC(weib_model), BIC(loglog_model)
   +    ), ncol = 2, dimnames = list(c("Weibull", "Loglog"), c("AIC", "BIC")))
-\end{CodeInput}
-\begin{CodeOutput}
                AIC      BIC
@@
   Loglog  1331.214 1350.278
-\end{CodeOutput}
-\end{CodeChunk}
+\end{example}
 
@@
 predictive performance, respectively. The hazard rates can be plotted by either 
-calling \code{plot_hazard} directly, or with \code{plot(x, type = "hazard")}, on
+calling \code{plot\_hazard} directly, or with \code{plot(x, type = "hazard")}, on
 a fitted \code{spdur} object. It will produce a plot of the conditional hazard, which is the probability of survival conditional on the covariates in the risk and
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> plot(weib_model,   type="hazard", main="Weibull")
-R> plot(loglog_model, type="hazard", main="Loglog")
-\end{CodeInput}
-\end{CodeChunk}
+\begin{example}
+  plot(weib_model,   type="hazard", main="Weibull")
+  plot(loglog_model, type="hazard", main="Loglog")
+\end{example}
 
@@
 
-By default the \code{plot_hazard} function
+By default the \code{plot\_hazard} function
 uses the mean values of the covariates during the simulations, but users
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> plot_hazard(loglog_model, main = "A")
-\end{CodeInput}
-\end{CodeChunk}
+\begin{example}
+  plot_hazard(loglog_model, main = "A")
+\end{example}
 
@@
 \caption{Plots of the hazard rate for the log-logistic model of coups.
-The left graph uses the default mean values for covariates, while graph B uses user-specified variable values for a high-risk military regime.}
+Graph A uses the default mean values for covariates, while graph B uses user-specified variable values for a high-risk military regime.}
 \label{hazard-ex}
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> plot_hazard(loglog_model, xvals = c(1, 1, 10, 0.05), 
+\begin{example}
+  plot_hazard(loglog_model, xvals = c(1, 1, 10, 0.05), 
   +    zvals = c(1, 7, 8.64, 1, 1, 0.05, 0, 0), main = "B")
-\end{CodeInput}
-\end{CodeChunk}
+\end{example}
 
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> plot(weib_model)
-R> plot(loglog_model)
-\end{CodeInput}
-\end{CodeChunk}
+\begin{example}
+  plot(weib_model)
+  plot(loglog_model)
+\end{example}
 
@@
 predictions. We will use data from 1996 onwards as the test set, and
-prior data for training purposes. The \code{add_duration} function
+prior data for training purposes. The \code{add\_duration} function
 retrospectively codes the risk variable based on how a spell ended, and
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> data(bscoup)
-R> bscoup$coup <- ifelse(bscoup$coup == "yes", 1, 0)
-R> coup_train <- bscoup[bscoup$year < 1996, ]
-R> coup_train <- add_duration(coup_train, "coup", unitID = "countryid", 
+\begin{example}
+  data(bscoup)
+  bscoup$coup <- ifelse(bscoup$coup == "yes", 1, 0)
+  coup_train <- bscoup[bscoup$year < 1996, ]
+  coup_train <- add_duration(coup_train, "coup", unitID = "countryid", 
   +    tID = "year", freq = "year", ongoing = FALSE)
-\end{CodeInput}
-\end{CodeChunk}
+\end{example}
 
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> coup_test  <- add_duration(bscoup, "coup", unitID = "countryid", 
+\begin{example}
+  coup_test  <- add_duration(bscoup, "coup", unitID = "countryid", 
   +    tID = "year", freq = "year", ongoing = FALSE)
-R> coup_test  <- coup_test[coup_test$year >= 1996, ]
-\end{CodeInput}
-\end{CodeChunk}
+  coup_test  <- coup_test[coup_test$year >= 1996, ]
+\end{example}
 
@@
 predictions from these models for the test data, using
-\code{predict(..., newdata = coup_test)}.
+\code{predict(..., newdata = coup\_test)}.
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> weib_model2   <- spdur(
+\begin{example}
+  weib_model2   <- spdur(
   +    duration ~ milreg + instab + regconf,
@@
 
-R> loglog_model2 <- spdur(
+  loglog_model2 <- spdur(
   +    duration ~ milreg + instab + regconf,
@@
   +    data = coup_train, distr = "loglog", silent = TRUE) 
-\end{CodeInput}
-\end{CodeChunk}
+\end{example}
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> weib2_test_p   <- predict(weib_model2, newdata = coup_test)
-R> loglog2_test_p <- predict(loglog_model2, newdata = coup_test)
-\end{CodeInput}
-\end{CodeChunk}
+\begin{example}
+  weib2_test_p   <- predict(weib_model2, newdata = coup_test)
+  loglog2_test_p <- predict(loglog_model2, newdata = coup_test)
+\end{example}
 
-Since we are predicting for data that is not contained in the ``spdur''
+Since we are predicting for data that is not contained in the \code{spdur}
 model objects, we have to use the \code{separationplot} functions
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> library("separationplot")
-R> obs_y <- coup_test[complete.cases(coup_test), "coup"]
+\begin{example}
+  library("separationplot")
+  obs_y <- coup_test[complete.cases(coup_test), "coup"]
 
-R> par(mfrow=c(2,1),mar=c(2,2,2,2))
-R> separationplot(weib2_test_p,   obs_y, newplot = FALSE)
-R> separationplot(loglog2_test_p, obs_y, newplot = FALSE)
-\end{CodeInput}
-\end{CodeChunk}
+  par(mfrow=c(2,1),mar=c(2,2,2,2))
+  separationplot(weib2_test_p,   obs_y, newplot = FALSE)
+  separationplot(loglog2_test_p, obs_y, newplot = FALSE)
+\end{example}
 
-\begin{figure}
+\begin{figure}[htbp!]
 \begin{center}
@@
 
-The \pkg{survival} library \citep{thernau2000modeling, thernau2015survival}, arguably the most well-established library for survival and event history analysis in \proglang{R}, allows for the estimation of a wide variety of semi-, non-, and parametric survival models, and provides facilities for handling and descriptive summaries of survival and event history data.
+The \CRANpkg{survival} library \citep{thernau2000modeling, thernau2015survival}, arguably the most well-established library for survival and event history analysis in {\sf R}, allows for the estimation of a wide variety of semi-, non-, and parametric survival models, and provides facilities for handling and descriptive summaries of survival and event history data.
 It does not include cure rate or split-population mixture models of the type we have implemented though.
 
-One useful extension of our project in the future would be to integrate and adapt the \code{Surv} class for survival data in \pkg{survival} to the \pkg{spduration} library.
-This would require some modification of the class as our model also requires information on the risk status of spells, but would give access to the much broader functionality, especially for descriptive summaries of data, in the \pkg{survival} library.
+One useful extension of our project in the future would be to integrate and adapt the \code{Surv} class for survival data in \CRANpkg{survival} to the \CRANpkg{spduration} library.
+This would require some modification of the class as our model also requires information on the risk status of spells, but would give access to the much broader functionality, especially for descriptive summaries of data, in the \CRANpkg{survival} library.
 
-Other currently available \proglang{R} routines for estimation of split-population/cure rate duration models include the packages \pkg{smcure}  \citep{cai2012smcureR}, described in \citet{cai2012smcure}, and \pkg{nltm} \citep{garibotti2010nltm}.
-\pkg{smcure} implements semi-parametric proportional hazards and accelerated failure time cure models using estimation procedures presented in \citet{peng2003fitting} and \citet{zhang2007new}.\footnote{There is also the older \pkg{gfcure} package, which estimates the parametric accelerated failure time cure model discussed in \citet{peng1998generalized}.
+Other currently available {\sf R} routines for estimation of split-population/cure rate duration models include the packages \CRANpkg{smcure}  \citep{cai2012smcureR}, described in \citet{cai2012smcure}, and \pkg{nltm} \citep{garibotti2010nltm}.
+\CRANpkg{smcure} implements semi-parametric proportional hazards and accelerated failure time cure models using estimation procedures presented in \citet{peng2003fitting} and \citet{zhang2007new}.\footnote{There is also the older \pkg{gfcure} package, which estimates the parametric accelerated failure time cure model discussed in \citet{peng1998generalized}.
 A version for Windows is available at \url{http://post.queensu.ca/~pengp/software.html}.} The package \pkg{nltm} will estimate a semi-parametric proportional hazard cure model, and a number of other survival models, using the approach developed in \citet{tsodikov2003semiparametric,tsodikov2007profile}.
 Semi-parametric estimation of the hazard function is attractive because it requires no assumption about its shape, and such assumptions can be difficult in practice to empirically evaluate.
-The main advantage \pkg{spduration} offers is the ability to include time varying covariates, which gives it a broader range of practical applications, as \pkg{smcure} and\pkg{nltm} will only accommodate case-based data.
-\pkg{spduration} also features more post-estimation methods than alternative packages, including those discussed above to create and evaluate model predictions.
+The main advantage \CRANpkg{spduration} offers is the ability to include time varying covariates, which gives it a broader range of practical applications, as \CRANpkg{smcure} and \pkg{nltm} will only accommodate case-based data.
+\CRANpkg{spduration} also features more post-estimation methods than alternative packages, including those discussed above to create and evaluate model predictions.
 
-Finally, there are \proglang{Stata} routines for implementing cure models \citep[see, e.g.,][]{lambert2007modeling,buxton2013cureregr}, some of which can handle time-varying covariates, but this is of course of limited appeal to \proglang{R} users.
+%Finally, there are {\sf Stata} routines for implementing cure models \citep[see, e.g.,][]{lambert2007modeling,buxton2013cureregr}, some of which can handle time-varying covariates, but this is of course of limited appeal to {\sf R} users.
 
@@
 
-\bibliography{spdur}
+\bibliography{beger-hill-metternich-minhas-ward}
 
-\end{document}
+\address{Andreas Beger\\
+  Ward Associates\\
+  Tallinn\\
+  Estonia\\
+  \email{adbeger@gmail.com}}
+
+\address{Daniel W. Hill, Jr.\\
+  Department of International Affairs\\
+  School of Public and International Affairs, University of Georgia\\
+  Candler Hall, Athens GA 30602\\
+  United States of America\\
+  \email{dwhill@uga.edu}}
+
+\address{Nils W. Metternich\\
+  Department of Political Science\\
+  School of Public Policy, University College London\\
+  London\\
+  United Kingdom\\
+  \email{n.metternich@ucl.ac.uk}}
+
+\address{Shahryar Minhas\\
+  Department of Political Science\\
+  Duke University\\
+  Durham, NC\\
+  United States of America\\
+  \email{sfm12@duke.edu}}
+
+\address{Michael D. Ward\\
+  Department of Political Science\\
+  Duke University\\
+  Durham, NC\\
+  United States of America\\
+  \email{michael.d.ward@duke.edu}}

### 2017-06-24 R Journal/beger-hill-metternich-minhas-ward.tex
small updates

@@
 \title{Splitting It Up: The \pkg{spduration} Split-Population Duration
-Regression Package}
+Regression Package for Time-varying Covariates}
 \author{by Andreas Beger, Daniel W. Hill, Jr., Nils. W. Metternich, Shahryar Minhas, and Michael D. Ward}
@@
 We present an implementation of split-population duration regression in
-the \CRANpkg{spduration} \citep{beger2016spduration} package for {\sf R} \citep{rstats2016} and an application to data on military
-coups. The statistical model accounts for units that are immune to a
+the \CRANpkg{spduration} \citep{beger2016spduration} package for {\sf R} \citep{rstats2016} that allows for time-varying covariates. The statistical model accounts for units that are immune to a
 certain outcome and not part of the duration process the researcher is
@@
 post-estimation methods for split-population Weibull and Loglogistic
-models.
+models. We provide an empirical application to data on military
+coups.
 }
@@
   Department of Political Science\\
-  School of Public Policy, University College London\\
+University College London\\
   London\\
   United Kingdom\\
-  \email{n.metternich@ucl.ac.uk}}
+  \email{n.metternich@ucl.ac.uk}\\
+  Nils W. Metternich acknowledges support from the Economic and Social Research Council (ES/L011506/1)}
 

### 2017-06-24 R Journal/beger-hill-metternich-minhas-ward.tex
added short paragraph

@@
 biased and predictions inaccurate.
+
+Additionally, risk factors might change over time which requires estimators 
+that can account for such dynamics. We therefore present a split-population 
+estimator that allows for the inclusion of time-varying covariates, which is an 
+important distinction to other R implementations of split-population or cure models.
 

### 2017-07-11 R Journal/beger-hill-metternich-minhas-ward.tex
this version submitted to R Journal

@@
 We present an implementation of split-population duration regression in
-the \CRANpkg{spduration} \citep{beger2016spduration} package for {\sf R} \citep{rstats2016} that allows for time-varying covariates. The statistical model accounts for units that are immune to a
+the \CRANpkg{spduration} \citep{beger2016spduration} package for {\sf R} that allows for time-varying covariates. The statistical model accounts for units that are immune to a
 certain outcome and not part of the duration process the researcher is
@@
 The last component to complete the likelihood is the choice of a
-distribution for the shape of the hazard rate. The \CRANpkg{spduration}
+distribution for the shape of the hazard rate. The \pkg{spduration}
 package implements two hazard rate shapes, Weibull and log-logistic:
@@
 \textrm{Weibull} \\
- f(t) & = & \alpha \lambda (\lambda t)^{\alpha - 1} e^{-(\lambda t)^\alpha} \\
-  S(t) & = & e^{ -(\lambda t )^\alpha } \\
- h(t) & = & \alpha \lambda (\lambda t)^{\alpha-1} \\
+ f(t) & = & \alpha \lambda \left(\lambda t \right)^{\alpha - 1} e^{-\left(\lambda t \right)^\alpha} \\
+  S(t) & = & e^{ -\left(\lambda t \right)^\alpha } \\
+ h(t) & = & \alpha \lambda \left(\lambda t \right)^{\alpha-1} \\
 \textrm{Log-logistic} \\
- f(t) & = & \frac{ \alpha \lambda (\lambda t)^{\alpha-1} }{ (1 + (\lambda t)^\alpha)^2 } \\
- S(t) & = & \frac{1}{ 1+  (\lambda t)^\alpha }  \\
- h(t) & = & \frac{ \alpha \lambda (\lambda t)^{\alpha-1} }{ 1+  (\lambda t)^\alpha }
+ f(t) & = & \frac{ \alpha \lambda \left(\lambda t \right)^{\alpha-1} }{ \left(1 + (\lambda t)^\alpha \right)^2 } \\
+ S(t) & = & \frac{1}{ 1+  \left(\lambda t \right)^\alpha }  \\
+ h(t) & = & \frac{ \alpha \lambda \left(\lambda t \right)^{\alpha-1} }{ 1+  \left(\lambda t \right)^\alpha }
 \end{eqnarray*}
@@
 \begin{eqnarray}
-h(t, \pi) = \frac{f(t, \pi)}{S(t, \pi)} & = & \frac{ \pi(t) \times f(t) }{ (1-\pi(t)) + \pi(t) \times S(t) } \\
- \pi(t) & = & \frac{ 1-\pi }{ S(t) + (1-\pi) (1 - S(t)) }
+h \left(t, \pi \right) = \frac{f \left(t, \pi \right)}{S \left(t, \pi \right)} & = & \frac{ \pi \left(t \right) \times f \left(t \right) }{ \left(1-\pi \left(t \right) \right) + \pi \left(t \right) \times S \left(t \right) } \\
+ \pi \left(t \right) & = & \frac{ 1-\pi }{ S \left(t \right) + \left(1-\pi \right) \left(1 - S \left(t \right) \right) }
 \end{eqnarray}
@@
 
-One useful extension of our project in the future would be to integrate and adapt the \code{Surv} class for survival data in \CRANpkg{survival} to the \CRANpkg{spduration} library.
-This would require some modification of the class as our model also requires information on the risk status of spells, but would give access to the much broader functionality, especially for descriptive summaries of data, in the \CRANpkg{survival} library.
+One useful extension of our project in the future would be to integrate and adapt the \code{Surv} class for survival data in \pkg{survival} to the \pkg{spduration} library.
+This would require some modification of the class as our model also requires information on the risk status of spells, but would give access to the much broader functionality, especially for descriptive summaries of data, in the \pkg{survival} library.
 
 Other currently available {\sf R} routines for estimation of split-population/cure rate duration models include the packages \CRANpkg{smcure}  \citep{cai2012smcureR}, described in \citet{cai2012smcure}, and \pkg{nltm} \citep{garibotti2010nltm}.
-\CRANpkg{smcure} implements semi-parametric proportional hazards and accelerated failure time cure models using estimation procedures presented in \citet{peng2003fitting} and \citet{zhang2007new}.\footnote{There is also the older \pkg{gfcure} package, which estimates the parametric accelerated failure time cure model discussed in \citet{peng1998generalized}.
+\pkg{smcure} implements semi-parametric proportional hazards and accelerated failure time cure models using estimation procedures presented in \citet{peng2003fitting} and \citet{zhang2007new}.\footnote{There is also the older \pkg{gfcure} package, which estimates the parametric accelerated failure time cure model discussed in \citet{peng1998generalized}.
 A version for Windows is available at \url{http://post.queensu.ca/~pengp/software.html}.} The package \pkg{nltm} will estimate a semi-parametric proportional hazard cure model, and a number of other survival models, using the approach developed in \citet{tsodikov2003semiparametric,tsodikov2007profile}.
 Semi-parametric estimation of the hazard function is attractive because it requires no assumption about its shape, and such assumptions can be difficult in practice to empirically evaluate.
-The main advantage \CRANpkg{spduration} offers is the ability to include time varying covariates, which gives it a broader range of practical applications, as \CRANpkg{smcure} and \pkg{nltm} will only accommodate case-based data.
-\CRANpkg{spduration} also features more post-estimation methods than alternative packages, including those discussed above to create and evaluate model predictions.
+The main advantage \pkg{spduration} offers is the ability to include time varying covariates, which gives it a broader range of practical applications, as \pkg{smcure} and \pkg{nltm} will only accommodate case-based data.
+\pkg{spduration} also features more post-estimation methods than alternative packages, including those discussed above to create and evaluate model predictions.
 
@@
 \bibliography{beger-hill-metternich-minhas-ward}
-
+\newpage
 \address{Andreas Beger\\
-  Ward Associates\\
-  Tallinn\\
-  Estonia\\
+  Predictive Heuristics\\
   \email{adbeger@gmail.com}}
@@
   Department of International Affairs\\
-  School of Public and International Affairs, University of Georgia\\
-  Candler Hall, Athens GA 30602\\
+  University of Georgia\\
+  Athens, GA\\
   United States of America\\

### 2017-09-26 R Journal/beger-hill-metternich-minhas-ward.tex
expand Table 1, fixes #3

--- 2017-06-24 R Journal/beger-hill-metternich-minhas-ward.tex
@@
 \begin{center}
-\begin{tabular}{lrrrrr} \toprule
-Country & Year & Coup & Spell ID & Duration \\ \midrule
-Portugal & 1960 & 0 & 1 & 1  \\ 
-\ldots & \ldots & 0 & 1 & \ldots \\
-\ldots & 1975 & 1 & 1 & 16 \\ \midrule
-Portugal & 1976 & 0 & 2 & 1 \\
-\ldots & \ldots & 0 & 2 & \ldots \\
-\ldots & 2000 & 0 & 2 & 20  \\ \midrule
-Canada & 1960 & 0 & 3 & 1 \\
-\ldots & \ldots & 0 & 3 & \ldots \\ 
-\ldots & 2000 & 0 & 3 & 41 \\ \bottomrule
+\begin{tabular}{lrrrrrrr} \toprule
+Country & Year & Coup & Spell ID & Failure & Censor & At risk & Duration \\ \midrule
+Portugal & 1960   & 0 & 1 & 0 & 0 & 1 & 1  \\ 
+\ldots   & \ldots & 0 & 1 & 0 & 0 & 1 & \ldots \\
+\ldots   & 1975   & 1 & 1 & 1 & 0 & 1 & 16 \\ \midrule
+Portugal & 1976   & 0 & 2 & 0 & 0 & 0 & 1 \\
+\ldots   & \ldots & 0 & 2 & 0 & 0 & 0 & \ldots \\
+\ldots   & 2000   & 0 & 2 & 0 & 1 & 0 & 20  \\ \midrule
+Canada   & 1960   & 0 & 3 & 0 & 0 & 0 & 1 \\
+\ldots   & \ldots & 0 & 3 & 0 & 0 & 0 & \ldots \\ 
+\ldots   & 2000   & 0 & 3 & 0 & 1 & 0 & 41 \\ \bottomrule
 \end{tabular}
@@
 returns the data frame with several additional variables needed for
-estimation. The most important among these are a counter of elapsed time
-in a spell, \code{duration}. It also codes that a spell was at risk
-(\code{atrisk==1}) if it ended in a failure event, and not at risk
-otherwise. These are then used as response variables for the formula in
-the \code{spdur} function.
+estimation.
+The most important of these are shown in Table \ref{tab-ex}.
+Using coups as the outcome, failure is coded when a coup occured.
+This also triggers a new spell if there are more observations for a country.
+If a spell ends without a failure event it is coded as right censored (\code{censor}).
+This can occur either because we have reached the end of observed data--the year 2000, or because a country dropped out of the data, e.g.
+if it was subsumed by another country.
+Duration is then coded each spell by counting the number of time periods until a spell ends.
+The at risk coding depends on whether a spell ended in failure--spells that end in failure are coded as being at risk for their entire duration, otherwise they are coded as not being at risk.
+
+The \code{add\_duration} function also adds several other variables that should be of less concern in typical use.
+If the data contain outcomes that can occur over multiple time periods, like civil wars, \code{add\_duration} should be used with the \code{ongoing = TRUE} flag so that only the first time period is treated as a failure, rather than as repeated failures.
+In that case \code{ongoing} is a marker for ongoing events beyond the initial onset.
+The variable \code{cured} is the inverse of \code{atrisk}, i.e.
+$1 -$~\code{atrisk}.
+Lastly, \code{t.0} marks the starting time for the row.
+It is by default \code{duration}~$ - 1$, but is retained to allow extension, if needed in the future, for observations that do not span a fixed time period, e.g.
+if observations are taken at irregular time intervals.
 

### 2017-10-02 R Journal/beger-hill-metternich-minhas-ward.tex
add 'distr = "weib"' instead of assuming default; fixes #2

@@
 \begin{example}
+  bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
   bscoup <- add_duration(bscoup, "coup", unitID = "countryid", 
@@
   +      camerica,
-  +    data = bscoup, silent = TRUE)
+  +    data = bscoup, distr = "weib", silent = TRUE)
 
@@
   +      camerica,
-  +    data = coup_train, silent = TRUE)
+  +    data = coup_train, distr = "weib", silent = TRUE)
 

### 2017-10-02 R Journal/beger-hill-metternich-minhas-ward.tex
fixes #4, not sure about #1

@@
   .. ..- attr(*, "names")= chr  "no" "yes"
+\end{example}
+
+Altogether the data include 290 coups, which makes for a positive rate of slightly more than 5\% of all observations.
+
+\begin{example}
   bscoup$coup <- ifelse(bscoup$coup == "yes", 1, 0)
@@
 \end{example}
-
-Altogether the data include 290 coups, which makes for a positive rate of slightly more than 5\% of all observations.
 
@@
 produces a regression model object of class \code{spdur} which can then
-be used with further methods. We begin by fitting first a Weibull and
+be used with further methods. We begin by fitting first a Weibull (the default) and
 then a log-logistic split-population duration model using the coups
@@
   +      camerica,
-  +    data = bscoup, distr = "weib", silent = TRUE)
+  +    data = bscoup, distr = "weibull", silent = TRUE)
 
@@
 standard output showing the model formula, estimates for the duration
-and risk equations, and test statistics with \(p\)-values. These can be
+and risk equations, and test statistics with \(p\)-values.
+The estimates are based on maximum likelihood estimation and standard errors are derived using the Hessian matrix of the optimization results.
+The \code{spdur} parameter estimates can be
 exported by calling \code{xtable}\citep[see][]{dahl2016xtable} directly on the \code{spdur} object to
@@
 \address{Andreas Beger\\
-  Predictive Heuristics\\
+  Ward Associates / Predictive Heuristics\\
+  Tallinn\\
+  Estonia\\
   \email{adbeger@gmail.com}}

### 2017-10-26 R Journal/beger-hill-metternich-minhas-ward.tex
revisions and memo complete

@@
 of predictions, including the probability that an observation is
-``at-risk'' and the probability of failure for a given time period.
+``at-risk'' and the probability of failure for a given time period. By default 
+\code{predict} calculates the conditional hazard, but the probability that an 
+observation is at risk may also be of interest. These models 
+are most appropriate in cases where there is a strong theoretical reason to suspect 
+some units are not at risk, as in the canonical applications (not everyone smokes, 
+some cancer patients are cured, some convicts do not relapse, etc.). Whether 
+an observation is cured or at-risk is ultimately unknowable, but estimating cure probabilities 
+can give users some idea of whether a split-population model is appropriate 
+for their application. Here we use \code{predict} to estimate conditional cure 
+probabilities from the Weibull model.
+
+\begin{example}
+  bscoup$ccure_prob <- predict(weib_model, type = "conditional cure", 
+  +    newdata = bscoup, na.action = "na.exclude")
+  mean(bscoup$ccure_prob, na.rm = TRUE)
+  [1] 0.4333912
+\end{example}
 

### 2017-10-30 R Journal/beger-hill-metternich-minhas-ward.tex
resubmission to R journal
added DOIs, example code all works and is consistent with paper

@@
 \begin{example}
-  bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
   bscoup <- add_duration(bscoup, "coup", unitID = "countryid", 
@@
 \begin{example}
-  plot_hazard(loglog_model, main = "A")
+  plot(loglog_model, type = "hazard", main = "A")
 \end{example}
@@
 \begin{example}
-  plot_hazard(loglog_model, xvals = c(1, 1, 10, 0.05), 
+  plot(loglog_model, type = "hazard", xvals = c(1, 1, 10, 0.05), 
   +    zvals = c(1, 7, 8.64, 1, 1, 0.05, 0, 0), main = "B")
@@
 \begin{example}
-  plot(weib_model)
-  plot(loglog_model)
+  plot(weib_model, lwd1 = 1, lwd2 = 1)
+  title(main="Weibull")
+  plot(loglog_model, lwd1 = 1, lwd2 = 1)
+  title(main="Loglog")
 \end{example}
@@
 \begin{example}
-  weib2_test_p   <- predict(weib_model2, newdata = coup_test)
-  loglog2_test_p <- predict(loglog_model2, newdata = coup_test)
+  weib2_test_p   <- predict(weib_model2, newdata = coup_test, na.action = "na.omit")
+  loglog2_test_p <- predict(loglog_model2, newdata = coup_test, na.action = "na.omit")
 \end{example}

### 2017-10-30 R Journal/beger-hill-metternich-minhas-ward.R

@@ -1,0 +1,175 @@
+#  Replication script for R Journal manuscript examples
+
+library("spduration")
+library("foreign")
+library("magrittr")
+library("ggplot2")
+library("xtable")
+library("scales")
+
+#	Fit a split-population model on coups data
+#   ____________________
+
+data(bscoup)
+?bscoup
+head(bscoup[, 1:5])
+
+str(bscoup)
+
+bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
+table(bscoup$coup)
+
+bscoup <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
+                       freq="year", ongoing = FALSE)
+
+weib_model <- spdur(
+  duration ~ milreg + instab + regconf,
+  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
+  data=bscoup, silent = TRUE)
+
+loglog_model <- spdur(
+  duration ~ milreg + instab + regconf,
+  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
+  data=bscoup, distr="loglog", silent = TRUE)
+
+library("xtable")
+tbl <- xtable(loglog_model, caption="Coup model with log-logistic hazard.",
+              label="loglog_table")
+print(tbl, caption.placement="bottom", comment=FALSE, include.rownames=FALSE)
+
+matrix(c(
+  AIC(weib_model), AIC(loglog_model), BIC(weib_model), BIC(loglog_model)
+  ), ncol = 2, dimnames = list(c("Weibull", "Loglog"), c("AIC", "BIC")))
+
+#   Plotting functions
+#   ____________________
+
+pdf("graphics/hazard-rates.pdf", height=4, width=8)
+par(mfrow=c(1, 2))
+plot(weib_model,   type="hazard", main="Weibull")
+plot(loglog_model, type="hazard", main="Loglog")
+dev.off()
+
+pdf("graphics/hazard-ex.pdf", height=4, width=8)
+par(mfrow = c(1, 2))
+plot_hazard(loglog_model, main = "A")
+plot_hazard(loglog_model, 
+     xvals = c(1, 1, 10, 0.05), 
+     zvals = c(1, 7, 8.64, 1, 1, 0.05, 0, 0),
+     main = "B")
+dev.off()
+
+pdf("graphics/hazard-ex.pdf", height=4, width=8)
+par(mfrow = c(1, 2))
[…]
